---
title: "Benchmarking detettore6110"
output:
  html_document: 
    theme: simplex
date: "2025-03-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

# Overview




# Calculate recall and precision
```{r}
library(ggplot2)
library(magrittr)
library(betareg)
library(margins)
library(cowplot)

library(RColorBrewer)
library(wesanderson)
library(viridis)

input_table <- snakemake@input[[1]]
output_html <- snakemake@output[[1]]

input_table <- '../results/benchmarking/detettore/benchmark_summary.tsv'


t <- read.table(input_table, sep='\t', header=T)

t$assembly <- as.factor(t$assembly)
t$coverage <- as.factor(t$coverage)
t$readlength <- as.factor(t$readlength)
t$side <- as.factor(t$side)

# Recall/sensitivity and precision, false discovery rate
t$recall = t$true_positives / (t$true_positives + t$false_negatives)
t$precision = t$true_positives / (t$true_positives + t$false_positives)
t$fdr = t$false_positives / (t$true_positives + t$false_positives)

# Number of unpaired inverted repeats
t$unpaired = t$unpaired_IRL + t$unpaired_IRR

```


# Check quality of IS annotation
Only include assemblies where the IS copy number equals the number of paired IRs. 
This excludes the 3 assemblies where IS copies are fragmented. I'm not aware of
biological processes that would lead to the presence of fragmented, active (!) IS 
in the genome; the other assemblies don't show this pattern; and, finally, the
three affected assemblies are all from the same source and based on Ion Torrent 
short reads.

```{r}
# Copy nr vs. nr of unpaired IRs
ggplot(t, aes(copy_number, unpaired)) + geom_point() + theme_bw()
t[which(t$unpaired>0), 1] %>% unique() %>% as.character()

# Copy nr vs. nr of paired IS
ggplot(t, aes(copy_number, paired_IRs)) + geom_point() + theme_bw()
exclude <- t[which(t$paired_IRs != t$copy_number), 1] %>% unique() %>% as.character()
exclude

# Exclude 3 assemblies
t_filt <- subset(t, !(assembly %in% exclude)) %>% droplevels()

```


# Recall/sensitivity
Plot recall against coverage. 
Show mean recall for all parameter combinations. Fit GLM to see which parameters affect recall.
```{r}
# Line plot
figS1a <- t_filt %>% 
  ggplot(aes(y=recall, x=coverage, col=side,group=interaction(assembly,side))) + 
  geom_point(alpha=0.5) + geom_line(alpha=0.5) + 
  facet_wrap(~readlength) + 
  theme(text = element_text(size = 12)) +
  theme_bw() + ylab('Sensitivity') + xlab('Coverage') +
  scale_color_brewer(palette='Dark2')

figS1a

ggsave('/home/cristobal/TB/projects/MTBC_IS6110-poly/workflow/results/figures/figS1a.sensitivity_lines.svg', figS1a, 
       width=12, height = 6, units = 'cm')

# Boxplot
figS1b <- t_filt  %>% 
  ggplot(aes(y=recall, x=coverage, fill=side)) + 
  geom_boxplot(alpha=0.5) + 
  facet_wrap(~readlength) +
  theme(text = element_text(size = 12)) +
  theme_bw() + ylab('Sensitivity') + xlab('Coverage') +
  scale_fill_brewer(palette='Dark2')

figS1b

ggsave('/home/cristobal/TB/projects/MTBC_IS6110-poly/workflow/results/figures/figS1b.sensitivity_boxes.svg', figS1b, 
       width=12, height = 6, units = 'cm')

aggregate(recall~coverage+readlength+side, t_filt, mean)

```

# What affects recall/sensitivity?
Use beta regression to model a proportion:

- slightly change 0s and 1s to avoid boundary issues
- weight by copy number to give less weight to low CN strains
- coefficients from beta regression are log odds. Exponentiate and get marginal effects for interpretation

Results: coverage of 20 can produce quite some false negatives, leading to very low
sensitivity in low-CN strains. Otherwise there seem to be single elements in single
assemblies that are occasionally missed, but overall sensitivity is high!

```{R}
# adjust 0s and 1s
t_filt$recall_adj <- (t_filt$recall * (nrow(t_filt) - 1) + 0.5) / nrow(t_filt)

m_recall_beta <- betareg(recall_adj ~ assembly + coverage + readlength + side, data = t_filt, weights=t_filt$copy_number)

summary(m_recall_beta)

# Get odds from log odds ration
exp(coef(m_recall_beta))

# Marginal effects
m_recall_beta_margins <- margins(m_recall_beta)
summary(m_recall_beta_margins)

```

# Precision
Same for precision. 
```{r}

# Line plot
figS1c <- t_filt %>% 
  ggplot(aes(y=precision, x=coverage, col=side,group=interaction(assembly,side))) + 
  geom_point(alpha=0.5) + geom_line(alpha=0.5) + 
  facet_wrap(~readlength) + 
  theme(text = element_text(size = 12)) +
  theme_bw() + ylab('Precision') + xlab('Coverage') +
  scale_color_brewer(palette='Dark2')

figS1c

ggsave(
  '/home/cristobal/TB/projects/MTBC_IS6110-poly/workflow/results/figures/figS1c.precision_lines.svg', 
  figS1c, width=12, height = 6, units = 'cm')

# Boxplot
figS1d <- t_filt %>% 
  ggplot(aes(y=precision, x=coverage, fill=side)) + 
  geom_boxplot(alpha=0.5) + 
  facet_wrap(~readlength) +
  theme(text = element_text(size = 12)) +
  theme_bw() + ylab('Precision') + xlab('Coverage') +
  scale_fill_brewer(palette='Dark2')

figS1d

ggsave(
  '/home/cristobal/TB/projects/MTBC_IS6110-poly/workflow/results/figures/figS1d.precision_boxes.svg', 
  figS1d, width=12, height = 6, units = 'cm')

aggregate(precision~coverage+readlength+side, t_filt, mean)


# Single assemblies affect precision, especially ASM25337
subset(t_filt, side =='5' & false_positives > 0) %>% 
  ggplot(aes(y=as.factor(false_positives), x=copy_number, col=assembly)) +
  geom_jitter(width=0.05, height = 0.05, alpha=0.8) + xlab('Copy number') + ylab('False positives') +
  scale_x_continuous(breaks = seq(0, 25, by = 2)) +
  scale_color_brewer(palette='Set1') +
  theme_bw()


```


# What affects precision?

Results: there is a single case of 3 false positives, else max FP is 2. The assembly is the 
main factor affecting precision, read length and coverage have nothing to do with it. In these
few assemblies with false positives, the 3' side is for some reason less precise. 

Whether these effects are due to misassembly or IS copies that have diverged from the 
target IS through some biological process is unclear. These are exceptions, however; mean precision
is overall high at 97%. 

```{r}
#adjust 0s and 1s
t_filt$precision_adj <- (t_filt$precision * (nrow(t_filt) - 1) + 0.5) / nrow(t_filt)

m_precision_beta <- betareg(precision_adj ~ assembly + coverage + readlength + side, data = t_filt, weights=t_filt$copy_number)

summary(m_precision_beta)

# Get odds from log odds ration
exp(coef(m_precision_beta))

# Marginal effects!
m_precision_beta_margins <- margins(m_precision_beta)
summary(m_precision_beta_margins)

```

# Are false positives due to unpaired IRs?
Only in one assembly (ASM25337). This strain has 2 IS6110 copies and a pair of dispersed IRL and IRR. 

```{r}
subset(t_filt, side == '5' & (false_positives > 0 | unpaired > 0)) %>% 
  ggplot(aes(x=false_positives, y=unpaired, col=assembly)) + geom_jitter(width = 0.1, height = 0.1) 

```


# Number of supporting reads

- How many supporting reads can we expect at different coverages?
- How do z-scores behave when there are few observations?

```{r}
library(ggplot2)

anchors <- read.csv('workflow/results/benchmarking/detettore/ALL_anchors.tsv', sep='\t')
head(anchors)

# Split sample name
split_col <- strsplit(anchors$strain, "_", fixed = TRUE)
split_df <- as.data.frame(do.call(rbind, split_col))
colnames(split_df) <- c("ASM", "COV", "RL")
anchors <- cbind(anchors, split_df)

# Plot distributions of supporting reads
ggplot(anchors, aes(x=COV, y=num_reads, col=RL)) + 
  geom_boxplot(size=1)  + geom_jitter(alpha=0.1) +
  theme_bw()


# Plot distributions of z-scores
ggplot(anchors, aes(x=COV, y=mod_z_score,col=RL)) + 
  geom_violin(size=1)  + geom_jitter(alpha=0.1) +
  theme_bw()


# Classifier to check if num_reads and/or z-score can identify false positives
library(randomForest)
model_rf <- randomForest(as.factor(is_false_positive) ~ num_support + z_score, data=df, ntree=500)
print(model_rf)

```






