---
title: "Prologue: parse input data"
output:
  html_document: 
    theme: simplex
date: "2025-03-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

Make all results analysis-ready and store tables in an RData object.

# Snakemake I/O
```{r}
paths <- tryCatch(
  {
    list(
      tree             = snakemake@input$tree,
      metadata         = snakemake@input$metadata,
      matrix5          = snakemake@input$matrix5,
      matrix3          = snakemake@input$matrix3,
      matrix_metadata  = snakemake@input$matrix_md,
      refins           = snakemake@input$refins,
      anchormap        = snakemake@input$anchormap,
      copy_numbers     = snakemake@input$copy_numbers,

      output_html      = snakemake@output$html,
      output_rdata     = snakemake@output$rdata
    )
  },
  error = function(e){
    list(
      tree             = 'workflow/results/phylogeny/snp_alignment.rooted.treefile',
      metadata         = 'metadata/10k/metadata_reduced.tsv',
      matrix5          = 'workflow/results/detettore/ALL_presence-absence.5prime.tsv',
      matrix3          = 'workflow/results/detettore/ALL_presence-absence.3prime.tsv',
      matrix_metadata  = 'workflow/results/detettore/ALL_presence-absence.metadata.tsv',
      refins           = 'workflow/results/detettore/ALL_reference_insertions.tsv',
      anchormap        = 'workflow/results/detettore/anchor_map.tsv',
      copy_numbers     = 'workflow/results/detettore/ALL_copy_numbers.tsv',
      output_html      = NULL,
      output_rdata     = 'workflow/results/is6110.RData'
    )
  }
)

source('workflow/notebooks/utils.R')
```


# Load results
and save them in an RData file
```{r}
library(ape)

# Metadata
metadata <- read.csv(paths$metadata, sep='\t')

# Tree
tree <- read.tree(paths$tree)

# Copy numbers
copy_numbers <- read.csv(paths$copy_numbers, sep='\t')

# Presence/absence matrices
matrix5 <- read.csv(paths$matrix5, sep='\t')
rownames(matrix5) <- matrix5[,1]
matrix5 <- matrix5[,-1]
print(paste('Number of 5 prime insertion sites: ', ncol(matrix5)))

matrix3 <- read.csv(paths$matrix3, sep='\t')
rownames(matrix3) <- matrix3[,1]
matrix3 <- matrix3[,-1]
print(paste('Number of 3 prime insertion sites: ', ncol(matrix3)))

# Matrix metadata
matrix_metadata <- read.csv(paths$matrix_metadata, sep='\t')

# Anchor-reference map
anchor_map <- read.csv(paths$anchormap, sep='\t', header=F)

# Reference insertions
refins <- read.csv(paths$refins, sep='\t', header=TRUE)
print(paste('Number of reference insertion sites: ', length(unique(refins$position))))


```


# Aesthetics
```{r}
# Load latex fonts (only has to be done once)
#library(extrafont)
#loadfonts()
# run the following command once
#font_import()


# Set plotting theme (not useful, ggsave onyl renders correct font when specified in the plot itself)
library(ggplot2)
plot_theme = theme_bw(base_family = "CMU Sans Serif", base_size=12)
theme_set(plot_theme)


# Color scales, fonts etc. 
library(ggplot2)

mtb_col_scale = c(
  "L1" = "#ff00ff", 
  "L2" = "#0000ff", 
  "L3" = "#a000cc", 
  "L4" = "#ff0000", 
  "L5" = "#663200", 
  "L6" = "#00cc33", 
  "L7" = "#ede72e", 
  "L8" ="#FF7F00" ,
  "L9" = "#006400",
  "L10" = "black",
  "La1"="#fbf390", 
  "La2"="#c1f9dd",
  "La3"="#dfe1f3",
  "La4"="#f7cfe7"
)

mtb_col_scale_animals_black <- c(
  "A1"="#fbf390", 
  "A2"="#c1f9dd", 
  "A3"="#dfe1f3", 
  "A4"="#f7cfe7", 
  "L1" = "#ff00ff", 
  "L2" = "#0000ff", 
  "L3" = "#a000cc", 
  "L4" = "#ff0000", 
  "L5" = "#663200", 
  "L6" = "#00cc33", 
  "L7" = "#ede72e", 
  "L8" ="#FF7F00" ,
  "L9" = "#006400",
  "L10" = "black",
  "La1"="black",
  "La2" ="black",
  "La3"="black",
  "La4"="black"
)
  

lineage_order = c('L1', 'L2', 'L3', 'L4', 'L5', 'L6', 'L7', 'L8', 'L9', 'L10','La1', 'La2', 'La3', 'LaX', 'Dassie bacillus')

```


# Load and augment gene map 
that matches MTBC0 gene names to H37Rv gene names
```{r}

augment_genemap <- function(genemap, dejesus_path, resistance_path){
  
  # Add DeJesus essentiality
  dejesus <- read.csv(dejesus_path, sep= '\t')
  genemap$essentiality <- as.factor(dejesus$Final.Call[match(genemap$h37rv_id, dejesus$ORF.ID)])
  
  # Distance to oriC
  genemap$dist_ori <- sapply(genemap$h37rv_start, dist_to_oriC)
  
  
  # Add resistance association
  resist = read.delim(resistance_path, sep='\t')
  
  resistance_genes = c()
  for (locus in resist$variant){
    gene = strsplit(locus, '_') %>% unlist()
    resistance_genes = c(resistance_genes, gene[1])
  }
  resistance_genes = unique(resistance_genes)
  resistance_genes
  
  genemap$resistance = rep('no', nrow(genemap))
  genemap$resistance[which(genemap$gene_id %in% resistance_genes | genemap$h37rv_id %in% resistance_genes)] <- 'yes'  
  
  return(genemap)
}


# Add information to gene metadata: essentiality, distance to origin, resistance
genemap = read.delim("data/MTBC0/mtbc0_to_h37rv.genemap.tsv", sep='\t')


# Add essentiality and resistance information
genemap = augment_genemap(
  genemap,
  'manuscript/literatura/DeJesus2017/orfs.csv',
  'data/who_resistance/WHO-UCN-TB-2023.7-eng_genomic_coordinates.txt'
)

head(genemap)

```


# Add blast hits to matrix metadata
```{r, eval=FALSE}
matrix_metadata <- add_blast_info(matrix_metadata)
```


# Add alternative lineage names
```{r}
# Change names of lineages C and D
metadata$lineage = metadata$LINEAGE_x
metadata$sublineage = metadata$SUBLINEAGE_x
metadata$lineage[which(metadata$lineage=='C')] <- 'LaX'
metadata$lineage[which(metadata$lineage=='D')] <- 'Dassie bacillus'

```


# Rescale branch lengths to account for non-variable positions
Information from https://github.com/cstritt/large_variable_alignment/
```{r}
genome_size = 4411532  
excluded_positions = 673325
var_positions = 548052 # from workflow/results/phylogeny/snp_alignment.uniqueseq.phy

scaling_factor <- (genome_size - excluded_positions)/ var_positions
tree$edge.length = tree$edge.length * scaling_factor

```


# Get MRCA of lineages
```{r}
library(ape)
library(reshape2)
# Check monophyly and get mrca
mrcas <- list()
nomono <- list()

for (l in lineage_order){
  md_l <- subset(metadata, lineage==l)
  gnrs <- md_l$GNUMBER[which(md_l$GNUMBER %in% tree$tip.label)]
  
  is_mono <- is.monophyletic(tree, gnrs)
  mrca_node <- getMRCA(tree, gnrs)
  
  if (is_mono){
    mrcas[[l]] <- mrca_node
  }
  else {
    print(paste("Not monophyletic: ", l))
    all_descendants <- extract.clade(tree, mrca_node)$tip.label
    subtree <- keep.tip(tree, all_descendants)
    nomono[[l]] <- subtree
  }
}
mrcas
mrcas = as.data.frame(mrcas) %>% melt()
names(mrcas) <- c("lineage", "node")
mrcas


# Lineages 8, 10, and D are missing as they have a single sample
table(metadata$lineage)
subset(metadata, lineage == 'L8')
subset(metadata, lineage == 'L9')
subset(metadata, lineage == 'L10')
subset(metadata, lineage == 'Dassie bacillus')

# Use L9  mrca to label L9, 10 and Dassie bacillus, which are too close anyway but not monophyletic
singles <- data.frame(
  'lineage' = c('L8', 'L10', 'Dassie bacillus', 'L9/L10/Dassie'),
  'node' = c(
    which(tree$tip.label=='G26689'), 
    which(tree$tip.label=='G50409'),
    which(tree$tip.label=='G01494'),
    18206)
)

mrcas <- rbind(mrcas, singles)
mrcas

```

# Summarize hotspots
Old hotspot summary derived from individual-site ancestral states. 
In the study, sites are aggregated by region and ancestral states are inferred per region. 
```{r, eval=FALSE}

site_summary <- read.csv('workflow/results/detettore/AS_5prime.site_summary.tsv', sep='\t')
site_summary$site_id <- gsub("X", "", site_summary$site_id)

hotspots <- data.frame()
refins_absent = data.frame()  # to further investigate: why are some refins not mapped to the matrix?


# Only consider 5' reads
for (i in 1:nrow(refins)){

  if (i%%1000 == 0){print(i)}

  row = refins[i,]
  region = row$gene
  
  # GEt the corresponding uFS for a reference insertion
  if (row$strand == '+'){ua = subset(anchor_map, V1==row$strain & V2==row$anchor_5)$V3}
  else {ua = subset(anchor_map, V1==row$strain & V2==row$anchor_3)$V3}

  if (length(ua) == 0){
    refins_absent = rbind(refins_absent, row)
    next 
    }  

  if (ua %in% hotspots$site_id){next}

  ua_asr = subset(site_summary, site_id == ua)
  ua_meta = subset(matrix_metadata, site_id == ua)
  
  dfrow = data.frame(
    'region' = region,
    'site_id' = ua,
    'position' = ua_meta$ref_position,
    'births' = ua_asr$births,
    'deaths' = ua_asr$deaths,
    'nstrains' = ua_meta$nr_strains,
    'blasthit' = ifelse(is.null(ua_meta$blasthit), NA, ua_meta$blasthit),
    'rv_homolog' = row$rv_ortholog,
    'cluster' = ua_meta$cluster
  )
  hotspots = rbind(hotspots, dfrow)
}

```

# Clean up Rdata and save image
```{r}
rm(singles, row, md_l, ua_asr, ua_meta)
rm(paths)
save.image('is6110.RData')
```


# Fake tree and matrix for Figure 1
```{r}
library(ape)
library(phytools)
library(ggtree)
library(ggplot2)


# 1. Simulate a random phylogenetic tree with 20 samples
set.seed(123)  # for reproducibility
n_taxa <- 50
tree <- rtree(n_taxa)

# Plot the tree
plot(tree, main = "Simulated Phylogenetic Tree")
tree <- root(tree, 't1')

# Simulate 10 traits evolving on the tree
Q <- matrix(c(-0.1, 0.1,
              0.1, -0.1), 
            nrow = 2, byrow = TRUE)

rownames(Q) <- colnames(Q) <- c("0", "1")

n_traits <- 40
trait_matrix <- matrix(NA, nrow = n_taxa, ncol = n_traits)

for (i in 1:n_traits) {
  trait <- sim.Mk(tree, Q = Q, anc = "0")
  trait_matrix[, i] <- as.numeric(trait)
}
rownames(trait_matrix) <- tree$tip.label
colnames(trait_matrix) <- paste0("Trait_", 1:n_traits)

trait_matrix <- apply(trait_matrix, c(1, 2), as.character)

# Plot
p <- ggtree(tree)

pm <- gheatmap(p, trait_matrix, colnames = FALSE) +  
  scale_fill_manual(values=c('2'='black', '1'='white')) +
  theme(legend.position = "none") +
  scale_x_reverse()

pm

```