---
title: "Benchmarking detettore6110"
output:
  html_document: 
    theme: simplex
date: "2025-03-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


# Snakemake I/O
```{r}
paths <- tryCatch(
  {
    list(
      benchmarking = snakemake@input$benchmark_summary,
      assembly_meta = snakemake@input$assembly_metadata,
      output_html     = snakemake@output[[1]]
    )
  },
  error = function(e){
    list(
      benchmarking = 'workflow/results/benchmarking/detettore/benchmark_summary.tsv',
      assembly_meta = 'metadata/Harrison2024.metadata.tsv',
      output_html     = NULL
    )
  }
)

```


# Calculate sensitivity and precision
```{r}
bm <- read.table(paths$benchmarking, sep='\t', header=T)

bm$assembly <- as.factor(bm$assembly)
bm$coverage <- as.factor(bm$coverage)
bm$readlength <- as.factor(bm$readlength)
bm$side <- as.factor(bm$side)

# Recall/sensitivity and precision, false discovery rate
bm$sensitivity = bm$true_positives / (bm$true_positives + bm$false_negatives)
bm$precision = bm$true_positives / (bm$true_positives + bm$false_positives)
bm$fdr = bm$false_positives / (bm$true_positives + bm$false_positives)

# Number of unpaired inverted repeats
bm$unpaired = bm$unpaired_IRL + bm$unpaired_IRR

summary(bm)

```


## Table S1: benchmarking data set
```{r}
# Load assembly meta data
asm_md <- read.csv(paths$assembly_meta, sep='\t')

# Combine data frames
bm_md <- bm[!duplicated(bm$assembly), c('assembly', 'copy_number', 'paired_IRs', 'paired_IRs_wrong_seq', 'unpaired_IRL', 'unpaired_IRR')]
bm_md <- merge(bm_md, asm_md, by.x='assembly', by.y='Assembly')

head(bm_md)

# Write supplementary table
write.table(bm_md, 'figures/supplement/TableS1.benchmarking_dataset.tsv', sep='\t', quote=F, row.names = F)


```


## Check quality of IS annotation
Only include assemblies where the IS copy number equals the number of paired IRs. 
This excludes the 3 assemblies where IS copies are fragmented. I'm not aware of
biological processes that would lead to the presence of fragmented, active (!) IS 
in the genome; the other assemblies don't show this pattern; and, finally, the
three affected assemblies are all from the same source and based on Ion Torrent 
short reads.

In the legend, "number of annotated intact copies" is not correct, ISEScan annotated them in an untypical way. 

```{r}
library(ggplot2)

# Copy nr vs. nr of unpaired IRs
ggplot(bm, aes(copy_number, unpaired)) + geom_point() + theme_bw()
bm[which(bm$unpaired>0), 1] %>% unique() %>% as.character()

# Copy nr vs. nr of paired IS
figS1 <- ggplot(bm, aes(copy_number, paired_IRs)) + 
  geom_point() + theme_bw(base_family = "CMU Sans Serif", base_size=12) +
  xlab("Number of annotated intact IS6110 copies") + 
  ylab("Number of paired inverted repeats")

figS1

ggsave("figures/supplement/figS1.fragmented_copies.pdf", figS1, width = 12, height = 10, units = "cm", device = cairo_pdf)     


# Exclude 3 assemblies
exclude <- bm[which(bm$paired_IRs != bm$copy_number), 1] %>% unique() %>% as.character()
exclude
bm_filt <- subset(bm, !(assembly %in% exclude)) %>% droplevels()

# How many IS copies in total?
bm_copynr <- bm_filt[!duplicated(bm_filt$assembly), ]
print(paste('Number of IS6110 elements analyzed for benchmarking: ', sum(bm_copynr$copy_number)))

```


## Figure S2: sensitivity and precision
```{r}
library(ggplot2)
library(cowplot)
library(reshape2)

head(bm_filt)

summary(bm_filt)

aggregate(sensitivity~coverage+readlength+side, bm_filt, mean)
aggregate(precision~coverage+readlength+side, bm_filt, mean)

```


# Fig S2: Benchmarking results
```{r}

bm_filt_fig <- melt(bm_filt, id.vars=c('coverage', 'side','readlength'), measure.vars=c('precision', 'sensitivity'))

bm_filt_fig$variable = as.character(bm_filt_fig$variable)
bm_filt_fig$variable[which(bm_filt_fig$variable=='sensitivity')] <- 'Sensitivity'
bm_filt_fig$variable[which(bm_filt_fig$variable=='precision')] <- 'Precision'
bm_filt_fig$variable = factor(bm_filt_fig$variable, levels=c('Sensitivity', 'Precision'))


bm_filt_fig$readlength = as.character(bm_filt_fig$readlength)
bm_filt_fig$readlength[which(bm_filt_fig$readlength=='100')] <- '100 bp reads'
bm_filt_fig$readlength[which(bm_filt_fig$readlength=='150')] <- '150 bp reads'
bm_filt_fig$readlength = as.factor(bm_filt_fig$readlength)


# Reorder factor levels
bm_filt_fig$side = factor(bm_filt_fig$side, levels=c('5', '3'))
str(bm_filt_fig)


figS2 <- ggplot(bm_filt_fig, aes(y=value, x=coverage, fill=side)) +
  geom_boxplot(outlier.size = 0.5) + facet_grid(variable~readlength) +
  theme_bw(base_family = "CMU Sans Serif", base_size=12) +
  scale_fill_brewer(palette='Dark2') +
  xlab('Coverage') + ylab('') + labs(fill = "Side")

figS2

ggsave("figures/supplement/figS2.benchmarking.pdf", figS2, width = 15, height = 12, units = "cm",device = cairo_pdf)    
```