---
title: "IS6110 copy numbers"
output:
  html_document: 
    theme: simplex
date: "2025-03-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


# Snakemake I/O
```{r}
paths <- tryCatch(
  {
    list(
      rdata = snakemake@input$rdata,
      output_html     = snakemake@output[[1]]
    )
  },
  error = function(e){
    list(
      rdata           = 'is6110.RData',
      output_html     = NULL  # or whatever default makes sense
    )
  }
)

load(paths$rdata)
```


# Characterize uFS matrices
Number of identified uFS, site frequency spectrum.

```{r}
# Sort insertions sites by frequency
m_counts <- colSums(m) %>% as.data.frame()
m_counts$ins <- rownames(m_counts)
names(m_counts)[1] <- 'count'
m_counts <- m_counts[order(m_counts$count, decreasing = T),]

ggplot(m_counts, aes(count)) + geom_histogram() + scale_x_log10() +
  theme_bw() + xlab("Number of strains with insertion site") + ylab("Count")

```


# Part 2: Variation in IS6110 copy numbers
Fig. 2. a) Phylogeny of the 10,000 selected strains, estimated with ... The colors of the
tips indicated the copy number. Lineage labelling follows the 'modern' terminology, 
as summarized in ..., b) copy number distributions by lineage 

Add lineage and sublineage to copy number table
```{r}
# Add lineage to copy_numbers
copy_numbers$lineage = metadata$lineage[match(copy_numbers$strain, metadata$GNUMBER)]
copy_numbers$sublineage = metadata$sublineage[match(copy_numbers$strain, metadata$GNUMBER)]

head(copy_numbers)

```


# Correlation between 5' and 3' cFS counts
Given that 5' and 3' flanking sequences are inferred independently, do we find the same number of copies from the two sides?
```{r}
library(ggplot2)

head(copy_numbers)

ggplot(copy_numbers, aes(nr_5_anchors, nr_3_anchors)) + 
  geom_point(alpha=0.2) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) +
  xlab("Five prime cFS count") + ylab("Three prime cFS count") +
  geom_density2d(col='grey80')

r = cor(copy_numbers$nr_5_anchors, copy_numbers$nr_3_anchors)
print(paste('Correlation between 5 and 3 prime cFS counts: ', r))
```


# Ancestral state reconstruction with copy numbers
Infer IS6110 copy number ancestral states using maximum parsimony implemented in castor::asr_max_parsimony().
```{r}
library(castor)
library(treeio)
library(ggtree)
library(viridis)

head(copy_numbers)
summary(copy_numbers)

# Reorder matrix rows so they match the tree!
phyl_order <- match(tree$tip.label, copy_numbers$strain)
copy_numbers_ordered <- as.integer(copy_numbers[phyl_order,]$CN)
copy_numbers_ordered <- copy_numbers_ordered + 1


# Convert to numeric starting with 1
site_vec_mapped <- map_to_state_space(copy_numbers_ordered)

plot(density(site_vec_mapped$mapped_states))  

# Reconstruct ancestral states
asr_result <- asr_max_parsimony(
  tree, 
  tip_states = copy_numbers_ordered
  )

head(asr_result$ancestral_likelihoods)
summary(asr_result$ancestral_states)

ancestral_states = c()
for (i in nrow(asr_result$ancestral_likelihoods)){
  lhs = asr_result$ancestral_likelihoods[i,]
  max_i = which(lhs==max(lhs))
  if (length(max_i) > 1){anc_state = NA}
  else {anc_state = max_i -1 }
  ancestral_states = c(ancestral_states, anc_state)
}
ancestral_states


# Plot on tree!
p <- ggtree(tree, layout='circular',size=0.1, alpha=0.5)  %<+% metadata
p$data$copy_number <- c(copy_numbers_ordered, asr_result$ancestral_states)
#p + geom_tree(aes(col=copy_number),size=0.1, alpha=0.5) +  scale_color_viridis(option='turbo')

write.table(p$data, 'workflow/results/phylogeny/treedata.CN_ASR.tsv', sep='\t')
td <- as.treedata(p)
write.beast(td, 'workflow/results/phylogeny/snp_alignment.rooted.treefile.annotated') 


```


# Tree and boxplot
```{r}
library(ape)
library(ggtree)
library(RColorBrewer)
library(viridis)
library(reshape2)
library(cowplot)
library(magick)
library(ggplot2)

# Fig. 2a: tree -----------------------------------------------------------


# Color copy number, label lineages
metadata$cn <- as.numeric(copy_numbers$CN[match(metadata$GNUMBER, copy_numbers$strain)])

t <- ggtree(tree, layout="circular",size=0, alpha=0) %<+% metadata
# Add CN from ancestral state reconstruction above!
t$data$copy_number = c(copy_numbers_ordered, asr_result$ancestral_states)

t$data$lineage[which(t$data$lineage %in% c('L9', 'L10', 'Dassie bacillus'))] <- 'L9/L10/Dassie'

fig2a <- t + geom_tree(aes(col=copy_number),size=0.1, alpha=0.8) +
  #t + geom_tippoint(aes(col=cn), size=0.7, alpha=0.4, stroke=NA) +
  theme_void(base_family = "CMU Sans Serif", base_line_size = 10) + 
  theme(
    legend.position=c(0.88, 0.35), 
    legend.key.width = unit(2, "mm"),
    legend.title = element_text(size = 4),
    legend.text = element_text(size = 4),
    legend.title.position='bottom',
    legend.key.size = unit(3, "mm")) +
  geom_cladelab(
    data = subset(mrcas, !(lineage %in% c('L9', 'L10', 'Dassie bacillus'))), 
    mapping=aes(node = node, label = lineage),
    fontsize = 1.5, barsize=0.1, offset.text=0.0008,color='grey20') +
  scale_color_viridis(option = 'turbo') +
  labs(col = "IS6110 copy number")

fig2a


# Crop white space
ggsave("figures/main/fig2a.pdf", fig2a,
       width = 13, height = 13, units = "cm",
       device = cairo_pdf)      # full-vector

system("pdfcrop --margins 5 figures/main/fig2a.pdf figures/main/fig2a_crop.pdf")


# 2b: copy number distribution ----------------------------------------------
brewer.pal(3, "Dark2")
brewer.pal(3, "Set1")

copy_numbers$lineage <- factor(copy_numbers$lineage, levels = lineage_order)

copy_numbers$lineage_new <- as.character(copy_numbers$lineage)
copy_numbers$lineage_new[which(copy_numbers$lineage=='Dassie bacillus')] <- 'Db'
copy_numbers$lineage_new <- factor(copy_numbers$lineage_new, levels = c(lineage_order, 'Db'))



# Replace Dassie bacillus with Db
#copy_numbers$lineage[which(copy_numbers$lineage=='Dassie bacillus')] <- 'Db'


fig2b <- ggplot(copy_numbers, aes(y=CN,x=lineage_new)) + 
  geom_jitter(alpha=0.3, size=1, color="grey50", stroke=NA) + 
  geom_boxplot(alpha=0,outlier.shape = NA) +
  annotate("text",
           x = 1:length(table(copy_numbers$lineage)),
           y = 38,
           label = table(copy_numbers$lineage),
           col = "grey50", size=2) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 10) +
  xlab("Lineage") + ylab("IS6110 copy nr.")

fig2b


# 2c new: frequency distribution of CN
fig2c <- ggplot(copy_numbers, aes(CN)) + geom_histogram() +
  theme_bw(base_family = "CMU Sans Serif", base_size = 10) + 
  xlab('Copy number') + ylab('Nr. of strains')

fig2c



# Combine panels --------------------------------------------------------------
img_tree <- image_read("figures/main/fig2a_crop.pdf", density = 600)
p_tree  <- ggdraw() + draw_image(img_tree)

lower_panel <- plot_grid(
  fig2c, fig2b,nrow = 1,
  labels=c('B', 'C'),
  rel_widths = c(1,2),
  label_size = 10,
  label_fontfamily='CMU Sans Serif'
  )

plot_grid(p_tree, lower_panel, 
          labels = c('A', ''),
          label_size = 10,
          ncol=1, 
          rel_heights=c(3,1),
          label_fontfamily='CMU Sans Serif')


ggsave("figures/main/fig2.tree_copy-numbers.png",
       width = 15, height = 17, units = "cm",
       dpi = 300)

```


# Zero copies in L2 and L4 strains?
The L4 strain seems plausible, as the sublineage (L4.2.2) to which it belongs has a mean CN of 2. 

The L2 strain is not plausible, its sublineage (L2.2.1) has a mean CN of 20. In the study where this data was published (https://doi.org/10.1099/mgen.0.001022), there is no indication of read prefiltering, so it is unclear why there would be no IS6110-reads in this strain (SRR20969754). 

```{r}
copy_numbers[which(copy_numbers$CN==0),]

copy_numbers[which(copy_numbers$CN==0 & copy_numbers$lineage !='L1'),]

```


# Sublineage copy numbers 
```{r}
library(dplyr)

l='L1'
cn_subl <- subset(copy_numbers, lineage==l) %>% 
  group_by(sublineage) %>% 
  summarise(mean_value = median(CN))

print(cn_subl, n=nrow(cn_subl))
```


# Do generalists have higher CNs?
Define generalists and specialists as in Stucki et al. 2016, Figure 1.
Not the appropriate data to really test this. Either include more of these sublineage strains or skip.

```{r}
l4_range <- list(
  'L4.1.2' = 'generalist',
  'L4.1.3' = 'specialist',
  'L4.3' = 'generalist',
  'L4.5' = 'specialist',
  'L4.6.1' = 'specialist',
  'L4.6.2' = 'specialist',
  'L4.10' = 'generalist'
)

cn_l4_range = subset(copy_numbers, sublineage %in% names(l4_range))
table(cn_l4_range$sublineage)
table(cn_l4_range$CN, cn_l4_range$sublineage)
```


Plot CN in lineage 4 sublineages!
```{r}
l4_cn <- subset(out$metadata, LINEAGE_x=='L4')

ggplot(l4_cn,aes(y=cn,x=SUBLINEAGE_x)) + 
  geom_jitter(alpha=0.3, size=1, color="#377EB8", stroke=NA) + 
  geom_boxplot(alpha=0,outlier.shape = NA) +
  annotate("text",
           x = 1:length(table(l4_cn$SUBLINEAGE_x)),
           y = 38,
           label = table(l4_cn$SUBLINEAGE_x),
           col = "grey50", size=2) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  xlab("Sublineage") + ylab("IS6110 copy number")
```


## Table 1: lineagae summary
- add diversity (calculated from tree, e.g. recycle pyseer similiarity matrix)
```{r}
library(gt)
library(magrittr)
library(tidyverse)
library(ape)
library(ggplot2)

table1 <- copy_numbers %>%
  group_by(lineage) %>%
  summarise(
    n = n(),
    mean = mean(CN),
    min = min(CN),
    max = max(CN),
    var = var(CN),
    sd = sd(CN)
  )

table1


# Add number of insertion events
table1$nr_events <- rep(0, nrow(table1))
table1$nr_insertions <- rep(0, nrow(table1))

for (i in 1:nrow(table1)){
  
  lineage = table1$lineage[i] %>% as.character()
    
  gnrs = metadata$GNUMBER[metadata$lineage == lineage]
  m_lineage = matrices$`5`[which(rownames(matrices$`5`) %in% gnrs),] 
  dim(m_lineage)
  
  # Remove columns with sum == 0
  if (length(gnrs) > 1){
    m_lineage <- m_lineage[, colSums(m_lineage) != 0]
    dim(m_lineage)
    
    if (is.null(ncol(m_lineage))){table1$nr_events[i] <- 1}
    else {table1$nr_events[i] <- ncol(m_lineage)}  
    table1$nr_insertions[i] <- sum(m_lineage)
  }
  
  else {
    table1$nr_events[i] <- sum(m_lineage)
    table1$nr_insertions[i] <- sum(m_lineage)
  }
}


# Add lineage SNP diversity
dm <- cophenetic(tree)
dim(dm)
dm_smry <- summary(dm[upper.tri(dm)])

gnrs = rownames(dm)

dists_lineages = data.frame()

for (l in lineage_order){
  gnrs_l = subset(metadata, lineage == l)$GNUMBER
  dm_li = match(gnrs_l, gnrs)
  dm_l = dm[dm_li, dm_li]
  pwd_l = dm_l[upper.tri(dm_l)]
  df_l = data.frame(
    dists = pwd_l,
    lineage = rep(l, length(pwd_l))
  )
  dists_lineages = rbind(dists_lineages, df_l)
}

rm(dm)

ggplot(dists_lineages, aes(x=lineage, y=dists)) + 
  geom_boxplot(outliers=F) +
  xlab("") + ylab('Pairwise phylogenetic distance')

median_pwd <- tapply(dists_lineages$dists, dists_lineages$lineage, median)
table1$diversity <- round(median_pwd[table1$lineage], 5)



# Arrange table
table1$nr_events_per_strain <- table1$nr_events/table1$n
table1 <- table1 %>%  arrange(factor(lineage, levels=lineage_order))

# Write to file
write.table(as.data.frame(table1), file = "workflow/results/table1.tsv", quote = FALSE, sep = "\t", row.names = F)

#table1 <- table1[-c(14, 15),]
table1
table1 <- table1[,c("lineage", 'n', 'diversity', 'nr_events', 'nr_insertions', 'mean',  'sd','min', 'max')]

names(table1) <- c('Lineage', 'Strains', 'Diversity', 'Insertion sites', 'Total insertions', 'Mean', 'SD', 'Min', 'Max')

gt_tbl <- 
  gt(table1) |>
  fmt_number(columns= c(Strains, `Insertion sites`, `Total insertions`, Min, Max), decimals=0) |>
  fmt_number(columns= c(Mean, SD), decimals=1) |>
  tab_header(
    title = md("**Table 1**. IS6110 insertions in 10,000 strains")) |>
  tab_spanner(label = "Copy number",columns = c(Mean, SD, Min, Max)) |>
  tab_options(table_body.hlines.style = "none") |>
  opt_table_font(font='CMU Sans Serif', stack = "transitional")  |>
  opt_align_table_header("left") |>
  tab_footnote(
    footnote = "Diversity is calculated as the median pairwise phylogenetic distance within lineages.",
    locations = cells_column_labels(columns = "Diversity")) |>
  tab_footnote(
    footnote = md("This clade comprises *M. microti* and *M. pinnipedii*."),
    locations = cells_body(
      rows = table1$Lineage == "LaX",  # Replace "X" with the specific value
      columns = vars(Lineage)
    ))
  
gt_tbl

gt_tbl_latex <- as_latex(gt_tbl)
as.character(gt_tbl_latex) %>% cat(file="figures/main/table1.tex")

```