---
title: "Birth and death of IS6110"
output:
  html_document: 
    theme: simplex
date: "2025-03-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


# Snakemake I/O
```{r}

paths <- tryCatch(
  {
    list(
      branchsummary_5 = snakemake@input$branchsummary_5,
      branchsummary_3 = snakemake@input$branchsummary_3,
      treedata        = snakemake@input$treedata,
      rdata           = snakemake@input$rdata,
      output_html     = snakemake@output[[1]]
    )
  },
  error = function(e){
    list(
      branchsummary_5 = 'workflow/results/detettore/AS_5prime.branch_summary.tsv',
      branchsummary_3 = 'workflow/results/detettore/AS_3prime.branch_summary.tsv',
      treedata        = 'workflow/results/phylogeny/treedata.CN_ASR.tsv',
      rdata           = 'is6110.RData',
      output_html     = NULL  # or whatever default makes sense
    )
  }
)

```


# IS6110 diversity
Count the number of pairwise IS6110 differences within each lineage. 
```{r}
library(ggplot2)
library(magrittr)
library(ape)
library(reshape2)
library(Matrix)

#table1 <- read.csv('workflow/results/table1.tsv', sep='\t')
#table1$IS_diversity <- rep(NA, nrow(table1))

load(paths$rdata)

# If adding SNP distances to table, subsample to avoid OOM crash
add_snp_dist = TRUE  # OOM!
subsample_n = 1500

is_allele_spec <- function(mat){
  " Get binned IS allele frequencies from a matrix
  "

  mat <- as.matrix(mat)
  nstrains = nrow(mat)
  nsites = ncol(mat)

  allelecounts <- colSums(mat)

  low =  table(allelecounts <= nstrains*0.25)
  lowmid = table(nstrains*0.25 < allelecounts & allelecounts <= nstrains*0.5)
  midhigh = table(nstrains*0.5 < allelecounts & allelecounts <= nstrains*0.75)
  high = table(nstrains*0.75 < allelecounts & allelecounts <= nstrains)

  prop_out = data.frame(
    'low' = tryCatch({low[['TRUE']]}, error = function(e){0})/nsites,
    'lowmid' =  tryCatch({lowmid[['TRUE']]}, error = function(e){0})/nsites,
    'midhigh' =  tryCatch({midhigh[['TRUE']]}, error = function(e){0})/nsites,
    'high' =  tryCatch({high[['TRUE']]}, error = function(e){0})/nsites,
    'nsites' = nsites
  )
  return(prop_out)
}


pairwise_diff_raw <- function(mat) {
  " Count pairwise differences between rows of a presence-absence matrix
  "
  mat <- as.matrix(mat) * 1  # ensure numeric 0/1
  L <- ncol(mat)
  
  # Number of shared 1s between pairs
  shared_ones <- mat %*% t(mat)
  
  # Number of 1s in each strain
  ones_per_strain <- rowSums(mat)
  
  # Using: differences = ones_i + ones_j - 2 * shared_ones
  diff_mat <- outer(ones_per_strain, ones_per_strain, "+") - 2 * shared_ones
  
  return(diff_mat)  # full symmetric matrix of raw differences
}


to_table <- function(mat, value_name = "value") {
  rn <- rownames(mat)
  
  # Upper triangle WITHOUT diagonal
  idx <- which(upper.tri(mat), arr.ind = TRUE)
  
  data.frame(
    sample1 = rn[idx[, 1]],
    sample2 = rn[idx[, 2]],
    value = mat[idx],
    row.names = NULL,
    check.names = FALSE
  ) |>
    dplyr::rename(!!value_name := value)
}

# Optionally add SNP distances (may require subsampling)
if (add_snp_dist){
  dm <- cophenetic(tree)
  gnrs = rownames(dm)
}

# Create data frame with pairwise IS6110 and SNP distances!
pairwise_differences <- data.frame()
shared_allele_proportions <- data.frame()

for (l in lineage_order){
  if (l %in% c('L8', 'L10', 'Dassie bacillus')){next}

  tm_sub <- subset_tree_matrix(tree, matrix5, metadata, l, 'lineage') 
  mat_sub = tm_sub[[2]]
  diff_mat = pairwise_diff_raw(mat_sub)

  # Median pairwise IS differences
  median_pwd = median(diff_mat)
  #table1$IS_diversity[which(table1$Lineage==l)] <- median_pwd
  print(paste(l, round(median_pwd,4)))

  # Frequency spectrum for shared insertions
  aspec <- is_allele_spec(mat_sub)
  aspec$lineage <- l
  shared_allele_proportions <- rbind(shared_allele_proportions, aspec)

  # Subsample if combining with SNP distances
  nstrains = length(tm_sub[[1]]$tip.label)
  if (subsample_n > 0){nsubsample = ifelse(nstrains<subsample_n, nstrains, subsample_n)}

  # Convert to table
  diff_df <- to_table(diff_mat, 'is6110_distance')
  if (subsample_n > 0){diff_df <- diff_df[sample(nrow(diff_df), nsubsample), ]}

  # Optionally add SNP distances
  if (add_snp_dist){
    # Add phylogenetic distance
    gnrs_l = unique(diff_df$sample1, diff_df$sample2)
    dm_li = match(gnrs_l, gnrs)
    dm_l = dm[dm_li, dm_li]  
    diff_df_snps <- to_table(dm_l, 'snp_distance')
    diff_df <- merge(diff_df, diff_df_snps, by = c('sample1', 'sample2'))
  }

  # Add lineage 
  diff_df$lineage = rep(l, nrow(diff_df))
  pairwise_differences <- rbind(pairwise_differences, diff_df)
}

by(pairwise_differences$is6110_distance, pairwise_differences$lineage, summary)


# Normalized IS distances
pairwise_differences$is6110_normalized <- pairwise_differences$is6110_distance / pairwise_differences$snp_distance

# Overall differences
#diff_mat_full = pairwise_diff_raw(matrices$`5`)
#median(diff_mat_full)
```

## Figure 3

```{r}
library(cowplot)

# Raw IS distances
# Panel A1: Raw IS distances
a1 <- ggplot(pairwise_differences, aes(lineage, is6110_distance)) + 
  geom_boxplot(outlier.size = 0.5, size = 0.3, alpha = 0.7) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) +
  xlab("Lineage") + ylab(expression("Pairwise IS6110 differences " ~ d[IS])) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Panel A2: Normalized IS distances
a2 <- ggplot(pairwise_differences, aes(lineage, is6110_normalized)) + 
  geom_boxplot(outlier.size = 0.5, size = 0.3, alpha = 0.7) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) +
  xlab("Lineage") + ylab(expression("Normalized IS6110 differences " ~ d[IS] / d[xy])) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

plot_grid(a1, a2, ncol = 2, labels = c('A', 'B'), label_fontfamily = 'CMU Sans Serif')

ggsave('figures/main/fig3.IS_diversity.png', width = 22, height = 10, units = "cm", dpi = 300) 


```

Supplement: proportion of high-frequency insertions.
```{r}
library(ggrepel)
library(dplyr)


# Allele frequency spectrum
focal = c('L1', 'L2', 'L3', 'L4','L5', 'L6', 'La1', 'La2', 'La3')
#focal = c('L1', 'L2', 'L3', 'L4', 'La3')

subset(shared_allele_proportions, lineage %in% focal) %>%
  melt(measure.vars = c('low', 'lowmid', 'midhigh', 'high'), id.vars='lineage') %>%
  ggplot(aes(x=variable, y=value, fill=lineage)) + geom_histogram(stat='identity', position='dodge') +
  scale_fill_manual(values=mtb_col_scale) + theme_bw(base_family = "CMU Sans Serif", base_size = 12)


# Shared proportion vs count
shared_allele_proportions$f0_0.25 <- 
  shared_allele_proportions$low * shared_allele_proportions$nsites

shared_allele_proportions$f0.25_0.5 <- 
  shared_allele_proportions$lowmid * shared_allele_proportions$nsites

shared_allele_proportions$f0.5_0.75 <- 
  shared_allele_proportions$midhigh * shared_allele_proportions$nsites

shared_allele_proportions$f0.75_1 <- 
  shared_allele_proportions$high * shared_allele_proportions$nsites

subset(shared_allele_proportions, lineage %in% focal) %>%
  melt(measure.vars = c('f0.25_0.5', 'f0.5_0.75', 'f0.75_1'), id.vars='lineage') %>%
  ggplot(aes(x = variable, y = value, fill = lineage)) +
  geom_histogram(stat = 'identity', position = 'dodge') +
  scale_fill_manual(values = mtb_col_scale) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) +
  theme(
    legend.position = c(0.02, 0.98),
    legend.justification = c(0, 1),
    strip.background = element_blank(),   # removes grey facet box
    strip.text = element_blank()
  ) +
  ylab("Number of sites") +
  xlab("Frequency category") +
  facet_wrap(~variable, scales = 'free_x')

ggsave('figures/supplement/figS4.frequency_categories.png', width = 18, height = 14, units = "cm", dpi = 300) 


# Just the high frequency sites
subset(shared_allele_proportions, lineage %in% focal) %>%
  melt(measure.vars = c('f0.75_1'),
       id.vars = 'lineage') %>%
  mutate(variable = factor(variable,
    levels = c("f0.25_0.5", "f0.5_0.75", "f0.75_1"),
    labels = c("L", "M", "H"))) %>%
  ggplot(aes(x = lineage, y = value)) +
  geom_col(fill = "grey90", colour = "black") +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) +
  ylab("Number of high frequency sites") +
  xlab("Lineage")




```


# Birth and death rates
How does IS6110 activity vary across the MTBC? 
Do more copies imply more transposition? Or is this impossible to observe due to the effects of strong purifying selection?

Compare the distribution of birth and death rates among lineages! For this, assign lineage to internal nodes using the MRCA information.


## Load ancestral state summaries and annotated tree
```{r}
library(dplyr)
library(ggplot2)

load(paths$rdata)

branch_smry <- merge(
  read.csv(paths$branchsummary_5, sep='\t'),
  read.csv(paths$branchsummary_3, sep='\t'),
  by =c('parent', 'child')
)

treedata = read.csv(paths$treedata, sep='\t')

# Add births and deaths to tree
treedata <- left_join(treedata, branch_smry, by=c("node" = "child"))

# Assign lineage to internal nodes
node_lineage = data.frame()
node_sublineage = data.frame()

for (i in 1:nrow(mrcas)){
  row = mrcas[i,]
  l = row$lineage
  sl = row$sublineage
  n = row$node

  desc = get_descendants(tree, node=n)
  desc_df = data.frame(
    node = desc,
    lineage = rep(l, length(desc))
  )
  node_lineage = rbind(node_lineage, desc_df)
}

treedata$lineage = node_lineage$lineage[match(treedata$node, node_lineage$node)]

# Assign node type terminal/internal
treedata$node_type = rep('internal', nrow(treedata))
treedata$node_type[which(treedata$isTip)] <- 'terminal'
treedata$node_type <- as.factor(treedata$node_type)

# Add parent_CN (as appropriate predictor of birth rate!)
treedata$parent_CN <- treedata$copy_number[match(treedata$parent.x, treedata$node)]

# Parent vs. child node copy number
ggplot(treedata, aes(copy_number, parent_CN)) + 
  geom_point() + 
  geom_abline(intercept = 0, slope=1) +
  labs(title='Parent vs. child IS6110 copy number')

fit <- lm(copy_number ~ parent_CN, data=treedata)
summary(fit)

```


## Number of births and deaths
```{r}
# How many births and deaths in total?
n_births_five = round(sum(treedata$births.x, na.rm=T))
n_births_three = round(sum(treedata$births.y, na.rm=T))
n_deaths_five = round(sum(treedata$deaths.x, na.rm=T))
n_deaths_three = round(sum(treedata$deaths.y, na.rm=T))

paste('Number of births inferred from 5 prime:', n_births_five)
paste('Number of births inferred from 3 prime:', n_births_three)
paste('Number of deaths inferred from 5 prime:', n_deaths_five)
paste('Number of deaths inferred from 3 prime:', n_deaths_three)

# Correlation between 5' and 3' estimates
a <- ggplot(branch_smry, aes(births.x, births.y)) + geom_point() + geom_smooth() + theme_bw() + 
  xlab('Births inferred from 5-prime') + ylab('Births inferred from 3-prime')
b <- ggplot(branch_smry, aes(deaths.x, deaths.y)) + geom_point() + geom_smooth() + theme_bw() +
  xlab('Deaths inferred from 5-prime') + ylab('Deaths inferred from 3-prime')

cowplot::plot_grid(a,b,ncol=2,labels=c('A', 'B'), label_fontfamily = 'CMU Sans Serif')

cor(branch_smry$births.x,branch_smry$births.y)
cor(branch_smry$deaths.x,branch_smry$deaths.y)

paste('Number of deaths inferred from 5 prime:',  sum(treedata$deaths.x))
paste('Number of deaths inferred from 3 prime:',  sum(treedata$deaths.y))

ggsave('figures/supplement/figS5.births_deaths.corr_five_three.png', width = 16, height = 8, units = "cm", dpi = 300) 

```

How comes there are so much fewer deaths inferred from the 3' flanking sequences?
-> strand bias & known bias of ASR in inffering loss events (Cunningham 1999)
-> solution: focus on births

Now get some more information about the distribution of births/deaths on the tree.
```{r}
# Proportion of births/deaths on terminal branches
n_births_five_tip = round(sum(treedata$births.x[which(treedata$isTip)]))
n_births_three_tip = round(sum(treedata$births.y[which(treedata$isTip)]))
n_deaths_five_tip = round(sum(treedata$deaths.x[which(treedata$isTip)]))
n_deaths_three_tip = round(sum(treedata$deaths.y[which(treedata$isTip)]))

paste('Proportion of terminal 5 prime births:', n_births_five_tip/n_births_five)
paste('Proportion of terminal 3 prime births:', n_births_three_tip/n_births_three)
paste('Proportion of terminal 5 prime deaths:', n_deaths_five_tip/n_deaths_five)
paste('Proportion of terminal 3 prime deaths:', n_deaths_three_tip/n_deaths_three)


# Number of branches without births/deaths
paste('Branches without births 5 prime: ', length(which(treedata$births.x==0)))
paste('Branches without births 3 prime: ', length(which(treedata$births.y==0)))
paste('Branches without deaths 5 prime: ', length(which(treedata$deaths.x==0)))
paste('Branches without deaths 3 prime: ', length(which(treedata$deaths.y==0)))

```


## Estimate birth/death rates
Obtain a first approximation of birth/death rates by dividing the number of births by branch length. 

Remove the root, for which there are no counts. 

For interpretability, convert branch length units to genome-wide rather than per (callable) base pair, because IS births are also per genome rather than per site.

```{r}
# Remove root which lacks birth and death counts
treedata <- treedata[-which(is.na(treedata$births.x)),]

# Scale branch length by callable genome size (from 0_prologue.Rmd)
callable_genome_size = 4411532  - 673325
treedata$branch.length.gw = treedata$branch.length * callable_genome_size

# Estimate birth and death rates
treedata$birthrate.5 = treedata$births.x/treedata$branch.length.gw
treedata$deathrate.5 = treedata$deaths.x/treedata$branch.length.gw
treedata$birthrate.3 = treedata$births.y/treedata$branch.length.gw
treedata$deathrate.3 = treedata$deaths.y/treedata$branch.length.gw
```


### Issue of 0-length branches
There are some branches of length 0. Two ways to deal with them:
 - Add very small length. Problematic because this creates arbitrarilly large rates
 - Remove 0-length branches. 

```{r}
# How many branches with length 0?
nrow(treedata[which(treedata$branch.length==0),])
table(treedata[which(treedata$branch.length==0),]$node_type)

# How many births/deaths are there on branches of length 0?
sum(treedata[which(treedata$branch.length==0),]$births.x)
sum(treedata[which(treedata$branch.length==0),]$deaths.x)

# Add minimal value to 0-length branches
treedata <- treedata %>%
  mutate(
    # handle zero or non-positive branch lengths
    branch_length_adj = ifelse(is.na(branch.length) | branch.length <= 0,
                               min(branch.length[branch.length>0], na.rm=TRUE) * 1e-8,
                               branch.length)
  )

# There are few births/deaths on 0-length branches, so we chose option 2 and remove them. 

treedata_no0 <- subset(treedata, branch.length>0)
nrow(treedata_no0)
```

### Descriptive stats: births, branch lengths, birth rates 
```{r}
# Distribution of births
treedata$Branch <- rep("Internal", nrow(treedata))
treedata$Branch[treedata$isTip] <- "Terminal"

panelA <- ggplot(treedata, aes(births.x, fill=Branch)) + 
  geom_histogram(position='dodge') + 
  theme_bw(base_family = "CMU Sans Serif", base_size = 10) +
  xlab("Number of births") + ylab("Count") +
  scale_fill_brewer(palette = "Dark2") +
  theme(
    legend.position = c(0.8, 0.8),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    legend.key.size = unit(0.4, "cm"),
    legend.background = element_blank(),
    legend.key = element_blank()
  )


# How many branches with no births?
nrow(treedata[which(treedata$births.x==0),])


# Branch length vs. number of births
panelB <- ggplot(treedata, aes(x=branch.length.gw/1000, y=births.x)) + 
  geom_point(alpha=0.2) +
  geom_smooth() +
  theme_bw(base_family = "CMU Sans Serif", base_size = 10) +
  xlab("Branch length") + ylab("Number of births")

cor(treedata$births.x,treedata$branch.length.gw)


# Distribution of birth rate
panelC <- ggplot(subset(treedata_no0), aes(birthrate.5)) + geom_density() +
  theme_bw(base_family = "CMU Sans Serif", base_size = 10) +
  xlab("Birth rate")


# Terminal vs. internal branches
by(treedata$birthrate.5, treedata$isTip, summary)

# OVerall mean birth and death rate (not very useful given the skew)
summary(treedata_no0$birthrate.5)
summary(treedata_no0$birthrate.3)
summary(treedata_no0$deathrate.5)
summary(treedata_no0$deathrate.3)

# Median rate for branches with at least one event
summary(subset(treedata_no0, births.x>0))
summary(subset(treedata_no0, deaths.x>0))

```

### Birth rate vs. copy number
```{r}
# Does birth rate depend on copy number?
y_raw <- treedata$birthrate.5
y_pos <- y_raw[is.finite(y_raw) & !is.na(y_raw) & (y_raw > 0)]
log_min <- floor(log10(min(y_pos)))
log_max <- ceiling(log10(max(y_pos)))
# major breaks: 10^n
major_breaks <- 10^(log_min:log_max)
# minor breaks: 1..9 * 10^n (this gives the conventional minor gridlines)
minor_breaks <- as.vector(outer(1:9, 10^(log_min:log_max), `*`))
# keep only minor breaks within the data range
minor_breaks <- minor_breaks[minor_breaks >= min(y_pos) & minor_breaks <= max(y_pos)]
  
x_levels <- levels(as.factor(treedata$copy_number))

treedata_no7 <- subset(treedata, !(lineage %in% removelineage))

panelD <- ggplot(na.omit(treedata), aes(as.factor(copy_number), birthrate.5)) + 
  geom_boxplot(outlier.size = 0.3) + facet_wrap(~lineage) +
  xlab("Copy number") + ylab("IS6110 birth rate") + 
  scale_x_discrete(
    labels = ifelse(seq_along(x_levels) %% 5 == 1, x_levels, "")) + 
  scale_y_log10(breaks = major_breaks,minor_breaks = minor_breaks)+ 
  theme_bw(base_family = "CMU Sans Serif", base_size = 10) +
  scale_y_log10()

upper <- cowplot::plot_grid(panelA, panelB, labels=c('A','B'), nrow=1)
cowplot::plot_grid(upper, panelD, nrow=2, labels=c("","C"), rel_heights=c(1,2))
ggsave("figures/supplement/figS6.cn_vs_birthrate.observed.png", width = 15, height = 18, units = "cm", dpi=300) 
```


# Model birth and death rates

Potential problems:
  - zero inflation in response
  - imbalance: copy numbers not equally distributed between lineages
  - collinearity: high copy number associated to some lineages

Solutions:
  - models taking into account zero inflation (ziformula). Or simpler: negative binomial 
  - do not add splines in main analysis, but discuss them

## Prepare data
```{r}
summary(treedata)

# Split L1 into 4 sublineages (doesn't work for internal nodes!!)
#table(subset(treedata, lineage=='L1')$sublineage)
#treedata$lineage_mod <- treedata$lineage
#l1_indices = which(treedata$lineage=='L1')
#treedata$lineage_mod[l1_indices] <- treedata$sublineage[l1_indices]

# Remove lineages with few observations
removelineage = which(table(treedata$lineage) < 100)
removelineage <- c(removelineage, c('L7', 'L9', 'LaX'))

# Remove highly imbalances lineages
removelineage <- c(removelineage, c('La1', 'La2', 'La3'))


treedata_no0 <- treedata_no0 %>%
  mutate(
    parent = as.factor(parent.x),
    node_type = as.factor(node_type),
    copy_number_s = as.numeric(scale(copy_number)),
    parent_CN_s = as.numeric(scale(parent_CN))   # standardize predictor
  )


# Plot response variable: zero inflation!
ggplot(treedata_no0, aes(birthrate.5)) + geom_histogram()


# Create smaller dataset with variables used
treedata_sub <- subset(treedata_no0, !(lineage %in% removelineage) & !(is.na(lineage))) %>% droplevels()
treedata_sub <- treedata_sub[,c('births.x', 'deaths.x', 'births.y', 'deaths.y','branch.length', 'branch.length.gw','parent_CN', 'parent_CN_s', 'node_type', 'lineage', 'parent', 'birthrate.5')]
summary(treedata_sub)

treedata_sub$births <- as.integer(treedata_sub$births.x)
treedata_sub$deaths <- as.integer(treedata_sub$deaths.x)

# Log branch length for offset
treedata_sub$log_branch_length <- log(treedata_sub$branch.length.gw)

# Log parent CN
treedata_sub$log_parent_CN <- log(treedata_sub$parent_CN)
```

## Fit GLMM
Compare n models:
  - m1: glmm with parent node as random effect
  - m2: glmm with independent slopes and intercepts for lineages
  - m3: m2 with additional accounting for zero inflation
  - m4: m2 with splines for copy  number
  - m5: fit power law with log(parent_CN) as predictor


```{r}
library(glmmTMB)
library(ggeffects)
library(ggplot2)
library(DHARMa)
library(splines)


m1 <- glmmTMB(
  births ~ parent_CN + node_type + lineage + 
    (1 | parent) + 
    offset(log_branch_length),
  data = treedata_sub,
  family = nbinom2(),
)
summary(m1)


m2 <- glmmTMB(
  births ~ parent_CN + node_type + 
    (1 + parent_CN | lineage) + 
    (1 | parent) + 
    offset(log_branch_length),
  data = treedata_sub,
  family = nbinom2(),
)
summary(m2)

m2b <- glmmTMB(
  births ~ parent_CN + node_type + 
    (1 + parent_CN || lineage) + 
    (1 | parent) + 
    offset(log_branch_length),
  data = treedata_sub,
  family = nbinom2(),
)
summary(m2b)


# ADditionally correct for zero inflation
m3 <- glmmTMB(
  births ~ parent_CN + node_type + 
    (1 + parent_CN | lineage) + 
    (1 | parent) + 
    offset(log_branch_length),
  ziformula = ~ 1,
  data = treedata_sub,
  family = nbinom2(),
)
summary(m3)


# With splines for copy number, no lineage because too complex
m4 <- glmmTMB(
  births ~ ns(parent_CN, df = 4) + node_type + 
    (1 | parent) + 
    offset(log_branch_length),
  data = treedata_sub,
  family = nbinom2(link="log")
)
summary(m4)


# Fit power law with log(parent_CN) as predictor
m5 <- glmmTMB(
  births ~ log(parent_CN) + node_type +
    (1 + log(parent_CN) | lineage) +
    (1 | parent) +
    offset(log_branch_length),
  family = nbinom2(),
  data = treedata_sub
)
summary(m5)

# Add interaction term
m5_interact <- glmmTMB(
  births ~ log(parent_CN) * node_type +
    (1 + log(parent_CN) | lineage) +
    (1 | parent) +
    offset(log_branch_length),
  family = nbinom2(),
  data   = treedata_sub
)

summary(m5_interact)
AIC(m5, m5_interact)

m5_nodevar_nocor <- glmmTMB(
  births ~ log(parent_CN + 1) + node_type +
    (1 + log(parent_CN + 1) + node_type || lineage) +
    (1 | parent) +
    offset(log_branch_length),
  family = nbinom2(),
  data = treedata_sub
)

summary(m5_nodevar_nocor)

AIC(m1, m2, m3, m4, m5, m5_nodevar_nocor)

anova(m2,m5)

```


### Figure: main effect sizes
Copy number effect, with births per substitution per genome. 
Main figure: lineage-averaged effect for both internal and external branches
Supplement: 

```{r}
pred_df <- expand.grid(
  parent_CN         = 1:35,
  node_type         = c("internal", "terminal"),
  log_branch_length = 0,
  lineage           = unique(treedata_sub$lineage),
  parent            = NA
)

pred_link <- predict(
  m5,
  newdata = pred_df,
  type    = "link",
  re.form = NULL,  # include all random effects; parent=NA marginalises over parent RE
  se.fit  = FALSE  # se.fit not supported with re.form=NULL in glmmTMB
)

pred_df <- pred_df %>%
  mutate(fit = exp(pred_link))


# Population-level predictions for the main lines
pred_pop <- expand.grid(
  parent_CN         = 1:35,
  node_type         = c("internal", "terminal"),
  log_branch_length = 0,
  lineage           = NA,
  parent            = NA
)

pred_pop_link <- predict(
  m5,
  newdata = pred_pop,
  type    = "link",
  re.form = NA,
  se.fit  = TRUE
)

pred_pop <- pred_pop %>%
  mutate(
    fit = exp(pred_pop_link$fit),
    lwr = exp(pred_pop_link$fit - 1.96 * pred_pop_link$se.fit),
    upr = exp(pred_pop_link$fit + 1.96 * pred_pop_link$se.fit)
  )


fig4 <- ggplot() +
  geom_line(
    data = pred_pop,
    aes(x = parent_CN, y = fit, colour = node_type) ) +
  geom_ribbon(
    data = pred_pop,
    aes(x = parent_CN, y = fit, ymin = lwr, ymax = upr, fill = node_type),
    alpha = 0.2, colour = NA
  ) +
  coord_cartesian(clip = "off") +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    x      = "IS6110 copy number",
    y      = "Predicted IS6110 birth rate",
    colour = "Branch type",
    fill   = "Branch type"
  ) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 10) +
  theme(legend.position = c(0.05, 0.95), legend.justification = c("left", "top"))

fig4

ggsave("figures/main/figX.birth_model.png", width = 12, height = 10, units = "cm", dpi=300)

```

### Example effects
Ancestral strain with 10 copies increasing to 11
```{r}
# Predictions at CN=10 and CN=11
pred_rates <- data.frame(
  parent_CN = c(10, 10, 11, 11),
  node_type = c("internal", "terminal", "internal", "terminal"),
  log_branch_length = 0,
  lineage = NA,
  parent = NA
)

pred_link <- predict(
  m5,
  newdata = pred_rates,
  type = "link",
  re.form = NA
)

pred_rates$fit <- exp(pred_link)

pred_rates %>%
  mutate(
    CN_label = parent_CN,
    rate = round(fit, 5)
  ) %>%
  select(CN_label, node_type, rate)
```


### Per-element rates
Get the per-copy rate for internal and terminal branches.
```{r}
library(dplyr)

# Create a prediction dataset from observed data
pred_dat <- treedata_sub %>%
  mutate(
    branch_length = exp(log_branch_length)
  )

# Predict expected births
pred_dat$mu <- predict(
  m5,
  newdata = pred_dat,
  type = "response",
  re.form = NA   # marginal over random effects
)

# Convert to per-element rates
pred_dat <- pred_dat %>%
  mutate(
    rate_per_branch = mu / branch_length,
    rate_per_element = rate_per_branch / parent_CN
  )

# Average per node type
pred_dat <- pred_dat %>% filter(parent_CN > 0)

# Plot distributions
ggplot(pred_dat, aes(node_type,rate_per_element)) + geom_violin(fill="#1B9E77FF") +
  theme_bw(base_family = "CMU Sans Serif", base_size = 10)


avg_rates <- pred_dat %>%
  group_by(node_type) %>%
  summarise(
    mean_rate = mean(rate_per_element),
    median_rate = median(rate_per_element),
    sd_rate = sd(rate_per_element),
    se_rate = sd(rate_per_element) / sqrt(n()),
    n = n(),
    q25 = quantile(rate_per_element, 0.25),
    q75 = quantile(rate_per_element, 0.75),
    .groups = "drop"
  ) %>%
  mutate(
    ci_lower = mean_rate - 1.96 * se_rate,
    ci_upper = mean_rate + 1.96 * se_rate
  )
avg_rates

avg_rates_weighted <- pred_dat %>%
  group_by(node_type) %>%
  summarise(
    weighted_rate =
      sum(rate_per_element * branch_length) /
      sum(branch_length)
  )

avg_rates_weighted

```


How does this compare to Tanaka's transposition rate estimates?
```{r}

weighted_rate_terminal <- avg_rates_weighted$weighted_rate[2]
mean_CN <- mean(treedata_sub$parent_CN)

# births/copy/sub_per_genome → births/copy/sub_per_site
rate_per_copy_per_sub_per_genome <- weighted_rate_terminal / mean_CN
rate_per_copy_per_sub_per_site <- rate_per_copy_per_sub_per_genome / callable_genome_size

clock_rates      <- c(low = 1e-8, mid = 1e-7, high = 1e-6)
generations_per_year <- c(low = 300, mid = 365, high = 400)

results <- expand.grid(
    clock  = clock_rates,
    gen_yr = generations_per_year
  ) %>%
  mutate(
    # births/copy/sub/site ÷ (subs/site/year) = births/copy/year
    rate_per_copy_per_year = rate_per_copy_per_sub_per_site / clock,
    # births/copy/year ÷ (generations/year) = births/copy/generation  
    rate_per_copy_per_gen  = rate_per_copy_per_year / gen_yr
  )

results

ggplot(results, aes(rate_per_copy_per_gen, clock)) +
  geom_point() +
  facet_wrap(~gen_yr) + scale_x_log10()

```




### Supplementary figure: lineage effect
```{r}
library(ggrepel)
library(reshape2)

# Get label positions at max CN per lineage, for internal branches only
line_labels <- pred_df %>%
  group_by(lineage) %>%
  slice_max(parent_CN, n = 1)


panelA <- ggplot() +
  geom_line(
    data = pred_df,
    aes(x = parent_CN, y = fit,  group = interaction(lineage, node_type)),
    linetype = "dashed", alpha = 0.5, linewidth = 0.5
  ) +
  geom_ribbon(
    data = pred_pop,
    aes(x = parent_CN, y = fit, ymin = lwr, ymax = upr),
    alpha = 0.2, colour = NA
  ) +
  geom_line(
    data = pred_pop,
    aes(x = parent_CN, y = fit),
    linewidth = 1
  ) +
  facet_wrap(~node_type) +
  geom_text_repel(
    data = line_labels,
    aes(x = parent_CN, y = fit, label = lineage),
    direction   = "y",
    hjust       = 0,
    nudge_x     = 1,
    segment.size = 0.2,
    size        = 3
  ) +
  coord_cartesian(clip = "off") +
  scale_colour_brewer(palette = "Dark2") +
  scale_fill_brewer(palette = "Dark2") +
  labs(
    x      = "IS6110 copy number",
    y      = "Predicted IS6110 birth rate\n(births per substitution per genome)",
    colour = "Branch type",
    fill   = "Branch type"
  ) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 10) +
  theme(legend.position = c(0.05, 0.95), legend.justification = c("left", "top"))

panelA

# Correlation between random lineage slope and intercept
re <- ranef(m5)$cond$lineage
re$lineage <- rownames(re)

fixef_CN <- fixef(m5)$cond["log(parent_CN)"]
fixef_IC <- fixef(m5)$cond["(Intercept)"]
re$intercept <- fixef_IC + re$`(Intercept)`
re$slope     <- fixef_CN + re$`log(parent_CN)`


panelB <- ggplot(re, aes(slope, intercept)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    y = "Lineage intercept (baseline birth rate)",
    x = "Lineage CN slope"
  ) +
  geom_text_repel(
    data = re,
    aes(slope, intercept, label = lineage)) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 10)

cowplot::plot_grid(panelA, panelB, nrow=2, labels=c('A', 'C'))

ggsave("figures/supplement/figS7.lineage_effect.png", width = 12, height = 12, units = "cm", dpi=300)

```


```{r}
lower <- cowplot::plot_grid(panelC,panelD, nrow=1, labels=c('C', 'D'), label_fontfamily = 'CMU Sans Serif')


cowplot::plot_grid(upper, lower, nrow=2)

ggsave("figures/main/fig3.diversity_AND_birth_rate_model.png", width = 23, height = 20, units = "cm", dpi=300)
```





## Get beta value for the niche constraints model
```{r}
beta_hat <- fixef(m5)$cond["log(parent_CN)"]
beta_ci  <- beta_hat + c(-1, 1) * 1.96 * 
  sqrt(vcov(m5)$cond["log(parent_CN)", "log(parent_CN)"])
```






## Model validation
```{r}

# Model validation
sim <- simulateResiduals(fit_offset, n = 2000)
plot(sim)

# Test overdispersion
testDispersion(sim)

# Test zero-inflation
testZeroInflation(sim)

# Check predictor residual patterns
plotResiduals(sim, treedata_sub$parent_CN_s)
plotResiduals(sim, treedata_sub$node_type)
plotResiduals(sim, treedata_sub$lineage)

# QQ to check Tweedie assumption
qqnorm(residuals(sim)); qqline(residuals(sim))
```