---
title: "Birth and death of IS6110"
output:
  html_document: 
    theme: simplex
date: "2025-03-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


# Snakemake I/O
```{r}

paths <- tryCatch(
  {
    list(
      branchsummary_5 = snakemake@input[[1]],
      branchsummary_3 = snakemake@input[[2]],
      treedata        = snakemake@input[[3]],
      rdata           = snakemake@input[[4]],
      output_html     = snakemake@output[[1]]
    )
  },
  error = function(e){
    list(
      branchsummary_5 = 'workflow/results/detettore/AS_5prime.branch_summary.tsv',
      branchsummary_3 = 'workflow/results/detettore/AS_3prime.branch_summary.tsv',
      treedata        = 'workflow/results/phylogeny/treedata.CN_ASR.tsv',
      rdata           = 'is6110.RData',
      output_html     = NULL  # or whatever default makes sense
    )
  }
)

```


# IS6110 diversity
```{r}
library(ggplot2)
library(magrittr)
library(ape)
library(reshape2)
library(Matrix)

#table1 <- read.csv('workflow/results/table1.tsv', sep='\t')
#table1$IS_diversity <- rep(NA, nrow(table1))

load(paths$rdata)

# If adding SNP distances to table, subsample to avoid OOM crash
add_snp_dist = FALSE  # OOM!
subsample_n = 0

is_allele_spec <- function(mat){
  " Get binned IS allele frequencies from a matrix
  "

  mat <- as.matrix(mat)
  nstrains = nrow(mat)
  nsites = ncol(mat)

  allelecounts <- colSums(mat)

  low =  table(allelecounts <= nstrains*0.25)
  lowmid = table(nstrains*0.25 < allelecounts & allelecounts <= nstrains*0.5)
  midhigh = table(nstrains*0.5 < allelecounts & allelecounts <= nstrains*0.75)
  high = table(nstrains*0.75 < allelecounts & allelecounts <= nstrains)

  prop_out = data.frame(
    'low' = tryCatch({low[['TRUE']]}, error = function(e){0})/nsites,
    'lowmid' =  tryCatch({lowmid[['TRUE']]}, error = function(e){0})/nsites,
    'midhigh' =  tryCatch({midhigh[['TRUE']]}, error = function(e){0})/nsites,
    'high' =  tryCatch({high[['TRUE']]}, error = function(e){0})/nsites,
    'nsites' = nsites
  )
  return(prop_out)
}


pairwise_diff_raw <- function(mat) {
  " Count pairwise differences between rows of a presence-absence matrix
  "
  mat <- as.matrix(mat) * 1  # ensure numeric 0/1
  L <- ncol(mat)
  
  # Number of shared 1s between pairs
  shared_ones <- mat %*% t(mat)
  
  # Number of 1s in each strain
  ones_per_strain <- rowSums(mat)
  
  # Using: differences = ones_i + ones_j - 2 * shared_ones
  diff_mat <- outer(ones_per_strain, ones_per_strain, "+") - 2 * shared_ones
  
  return(diff_mat)  # full symmetric matrix of raw differences
}


to_table <- function(mat, value_name = "value") {
  rn <- rownames(mat)
  
  # Upper triangle WITHOUT diagonal
  idx <- which(upper.tri(mat), arr.ind = TRUE)
  
  data.frame(
    sample1 = rn[idx[, 1]],
    sample2 = rn[idx[, 2]],
    value = mat[idx],
    row.names = NULL,
    check.names = FALSE
  ) |>
    dplyr::rename(!!value_name := value)
}

# Optionally add SNP distances (may required subsampling)
if (add_snp_dist){
  dm <- cophenetic(tree)
  gnrs = rownames(dm)
}

# Create data frame with pairwise IS6110 and SNP distances!
pairwise_differences <- data.frame()
shared_allele_proportions <- data.frame()


for (l in lineage_order){
  if (l %in% c('L8', 'L10', 'Dassie bacillus')){next}

  tm_sub <- subset_tree_matrix(tree, matrix5, metadata, l, 'lineage') 
  mat_sub = tm_sub[[2]]
  diff_mat = pairwise_diff_raw(mat_sub)

  # Median pairwise IS differences
  median_pwd = median(diff_mat)
  #table1$IS_diversity[which(table1$Lineage==l)] <- median_pwd
  print(paste(l, round(median_pwd,4)))

  # Frequency spectrum for shared insertions
  aspec <- is_allele_spec(mat_sub)
  aspec$lineage <- l
  shared_allele_proportions <- rbind(shared_allele_proportions, aspec)

  # Subsample if combining with SNP distances
  nstrains = length(tm_sub[[1]]$tip.label)
  if (subsample_n > 0){nsubsample = ifelse(nstrains<subsample_n, nstrains, subsample_n)}

  # Convert to table
  diff_df <- to_table(diff_mat, 'is6110_distance')
  if (subsample_n > 0){diff_df <- diff_df[sample(nrow(diff_df), nsubsample), ]}

  # Optionally add SNP distances
  if (add_snp_dist){
    # Add phylogenetic distance
    gnrs_l = unique(diff_df$sample1, diff_df$sample2)
    dm_li = match(gnrs_l, gnrs)
    dm_l = dm[dm_li, dm_li]  
    diff_df_snps <- to_table(dm_l, 'snp_distance')
    diff_df <- merge(diff_df, diff_df_snps, by = c('sample1', 'sample2'))
  }

  # Add lineage 
  diff_df$lineage = rep(l, nrow(diff_df))
  pairwise_differences <- rbind(pairwise_differences, diff_df)
}

by(pairwise_differences$is6110_distance, pairwise_differences$lineage, summary)

# Overall differences
#diff_mat_full = pairwise_diff_raw(matrices$`5`)
#median(diff_mat_full)
```

## Figure 3
```{r}
library(ggrepel)
library(dplyr)


panelA <- ggplot(pairwise_differences, aes(lineage, is6110_distance)) + geom_boxplot(outlier.size=0.5, size=0.3) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) + xlab("Lineage") + ylab("IS6110 pairwise differences") 


# Allele frequency spectrum
focal = c('L1', 'L2', 'L3', 'L4','L5', 'L6', 'La1', 'La2', 'La3')
#focal = c('L1', 'L2', 'L3', 'L4', 'La3')

b <- subset(shared_allele_proportions, lineage %in% focal) %>%
  melt(measure.vars = c('low', 'lowmid', 'midhigh', 'high'), id.vars='lineage') %>%
  ggplot(aes(x=variable, y=value, fill=lineage)) + geom_histogram(stat='identity', position='dodge') +
  scale_fill_manual(values=mtb_col_scale) + theme_bw(base_family = "CMU Sans Serif", base_size = 12)


# Shared proportion vs count
shared_allele_proportions$f0_0.25 <- 
  shared_allele_proportions$low * shared_allele_proportions$nsites

shared_allele_proportions$f0.25_0.5 <- 
  shared_allele_proportions$lowmid * shared_allele_proportions$nsites

shared_allele_proportions$f0.5_0.75 <- 
  shared_allele_proportions$midhigh * shared_allele_proportions$nsites

shared_allele_proportions$f0.75_1 <- 
  shared_allele_proportions$high * shared_allele_proportions$nsites

panelB <- subset(shared_allele_proportions, lineage %in% focal) %>%
  melt(measure.vars = c('f0.25_0.5', 'f0.5_0.75', 'f0.75_1'), id.vars='lineage') %>%
  ggplot(aes(x = variable, y = value, fill = lineage)) +
  geom_histogram(stat = 'identity', position = 'dodge') +
  scale_fill_manual(values = mtb_col_scale) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) +
  theme(
    legend.position = c(0.02, 0.98),
    legend.justification = c(0, 1),
    strip.background = element_blank(),   # removes grey facet box
    strip.text = element_blank()
  ) +
  ylab("Number of sites") +
  xlab("Frequency category") +
  facet_wrap(~variable, scales = 'free_x')

panelB


# Panel B beta
panelB <- subset(shared_allele_proportions, lineage %in% focal) %>%
  melt(measure.vars = c('f0.75_1'),
       id.vars = 'lineage') %>%
  mutate(variable = factor(variable,
    levels = c("f0.25_0.5", "f0.5_0.75", "f0.75_1"),
    labels = c("L", "M", "H"))) %>%
  ggplot(aes(x = lineage, y = value)) +
  geom_col(fill = "grey90", colour = "black") +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) +
  ylab("Number of high frequency sites") +
  xlab("Lineage")

panelB


ggplot(subset(shared_allele_proportions, lineage %in% focal), aes(f0.75_1, high)) + 
  geom_point() +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) + geom_text_repel(aes(label=lineage), nudge_x = 1, size=2) +
  xlab("Number of high frequency sites") + ylab("Proportion of high freq. sites")


upper <- cowplot::plot_grid(panelA,panelB, ncol=2, labels=c('A', 'B'), label_fontfamily = 'CMU Sans Serif', rel_widths = c(3,2))
upper
#ggsave('figures/main/fig3.IS_diversity.png', width = 22, height = 8, units = "cm", dpi = 300) 



```



# Birth and death rates
How does IS6110 activity vary across the MTBC? 
Do more copies imply more transposition? Or is this impossible to observe due to the effects of strong purifying selection?

Compare the distribution of birth and death rates among lineages! For this, assign lineage to internal nodes using the MRCA information.



## Load ancestral state summaries and annotated tree
```{r}
library(dplyr)
library(ggplot2)

load(paths$rdata)

branch_smry <- merge(
  read.csv(paths$branchsummary_5, sep='\t'),
  read.csv(paths$branchsummary_3, sep='\t'),
  by =c('parent', 'child')
)

treedata = read.csv(paths$treedata, sep='\t')

# Add births and deaths
treedata <- left_join(treedata, branch_smry, by=c("node" = "child"))

# Assign lineage to internal nodes
node_lineage = data.frame()
node_sublineage = data.frame()

for (i in 1:nrow(mrcas)){
  row = mrcas[i,]
  l = row$lineage
  sl = row$sublineage
  n = row$node

  desc = get_descendants(tree, node=n)
  desc_df = data.frame(
    node = desc,
    lineage = rep(l, length(desc))
  )
  node_lineage = rbind(node_lineage, desc_df)
}

node_lineage

treedata$lineage = node_lineage$lineage[match(treedata$node, node_lineage$node)]

# Assign node type terminal/internal
treedata$node_type = rep('internal', nrow(treedata))
treedata$node_type[which(treedata$isTip)] <- 'terminal'
treedata$node_type <- as.factor(treedata$node_type)

# Add parent_CN (as appropriate predictor of birth rate!)
treedata$parent_CN <- treedata$copy_number[match(treedata$parent.x, treedata$node)]

ggplot(treedata, aes(copy_number, parent_CN)) + geom_point() + geom_abline(intercept = 0, slope=1)

fit <- lm(copy_number ~ parent_CN, data=treedata)
summary(fit)

```


## Number of births and deaths
```{r}
# How many births and deaths in total?
n_births_five = round(sum(treedata$births.x))
n_births_three = round(sum(treedata$births.y))
n_deaths_five = round(sum(treedata$deaths.x))
n_deaths_three = round(sum(treedata$deaths.y))

paste('Number of births inferred from 5 prime:', n_births_five)
paste('Number of births inferred from 3 prime:', n_births_three)
paste('Number of deaths inferred from 5 prime:', n_deaths_five)
paste('Number of deaths inferred from 3 prime:', n_deaths_three)

# Proportion of branches with births/deaths
paste('Branches without births 5 prime: ', length(which(treedata$births.x==0)))
paste('Branches without births 3 prime: ', length(which(treedata$births.y==0)))
paste('Branches without deaths 5 prime: ', length(which(treedata$deaths.x==0)))
paste('Branches without deaths 3 prime: ', length(which(treedata$deaths.y==0)))

# No birth or death
paste('Branches without events 5 prime: ', length(which(treedata$births.x>0 | treedata$deaths.x>0)))
paste('Branches without events 3 prime: ', length(which(treedata$births.y>0 | treedata$deaths.y<0)))

# Proportion of births/deaths on terminal branches
n_births_five_tip = round(sum(treedata$births.x[which(treedata$isTip)]))
n_births_three_tip = round(sum(treedata$births.y[which(treedata$isTip)]))
n_deaths_five_tip = round(sum(treedata$deaths.x[which(treedata$isTip)]))
n_deaths_three_tip = round(sum(treedata$deaths.y[which(treedata$isTip)]))

paste('Proportion of terminal 5 prime births:', n_births_five_tip/n_births_five)
paste('Proportion of terminal 3 prime births:', n_births_three_tip/n_births_three)
paste('Proportion of terminal 5 prime deaths:', n_deaths_five_tip/n_deaths_five)
paste('Proportion of terminal 3 prime deaths:', n_deaths_three_tip/n_deaths_three)

```


### Correlation between 5' and 3' estimates
```{r}
# Correlation between 5' and 3' estimates
a <- ggplot(branch_smry, aes(births.x, births.y)) + geom_point() + geom_smooth() + theme_bw() + 
  xlab('Births inferred from 5-prime') + ylab('Births inferred from 3-prime')
b <- ggplot(branch_smry, aes(deaths.x, deaths.y)) + geom_point() + geom_smooth() + theme_bw() +
  xlab('Deaths inferred from 5-prime') + ylab('Deaths inferred from 3-prime')

cowplot::plot_grid(a,b,ncol=2,labels=c('A', 'B'), label_fontfamily = 'CMU Sans Serif')

cor(branch_smry$births.x,branch_smry$births.y)
cor(branch_smry$deaths.x,branch_smry$deaths.y)

paste('Number of deaths inferred from 5 prime:',  sum(treedata$deaths.x))
paste('Number of deaths inferred from 3 prime:',  sum(treedata$deaths.y))

ggsave('figures/supplement/figSx.births_deaths.corr_five_three.png', width = 16, height = 8, units = "cm", dpi = 300) 
```

How comes there are so much fewer deaths inferred from the 3' flanking sequences?
-> strand bias & known bias of ASR in inffering loss events (Cunningham 1999)
-> solution: focus on births


## Estimate birth/death rates
Substitute minimal values for 0-length branches. Remove the root, for which there are no counts.
```{r}
# Add minimal value to 0-length branches
treedata <- treedata %>%
  mutate(
    # handle zero or non-positive branch lengths (for weights)
    branch_length_adj = ifelse(is.na(branch.length) | branch.length <= 0,
                               min(branch.length[branch.length>0], na.rm=TRUE) * 1e-8,
                               branch.length)
  )


# Remove root which lacks birth and death counts
treedata <- treedata[-which(is.na(treedata$births.x)),]


```


Estimate birth and death rates. For interpretability, convert branch length units to genome-wide rather than per (callable) base pair, because IS births are also per genome rather than per site.
```{r}
# Scale branch length by callable genome size (from 0_prologue.Rmd)
callable_genome_size = 4411532  - 673325
treedata$branch.length.gw = treedata$branch.length * callable_genome_size

# Get birth and death rates
treedata$birthrate.5 = treedata$births.x/treedata$branch.length.gw
treedata$deathrate.5 = treedata$deaths.x/treedata$branch.length.gw
treedata$birthrate.3 = treedata$births.y/treedata$branch.length.gw
treedata$deathrate.3 = treedata$deaths.y/treedata$branch.length.gw


```

Issue with branches of length 0
```{r}
# How many branches with length 0?
nrow(treedata[which(treedata$branch.length==0),])
table(treedata[which(treedata$branch.length==0),]$node_type)

# How many births/deaths are there on branches of length 0?
sum(treedata[which(treedata$branch.length==0),]$births.x)
sum(treedata[which(treedata$branch.length==0),]$deaths.x)
```

There are few births/deaths on 0-length branches, these are ignored. The alternative - estimating rates with the assigned minimal values - seems more problematic because this would result in arbitrarily huge rates. 
```{r}
treedata_no0 <- subset(treedata, branch.length>0)
nrow(treedata_no0)
```

Some basic information
```{r}
# Distributions of birth and death rates
ggplot(treedata, aes(x=isTip, y=birthrate.5)) + geom_boxplot() + theme_bw()

# Terminal vs. internal branches
by(treedata$birthrate.5, treedata$isTip, summary)

# OVerall mean birth and death rate (not very useful given the skew)
summary(treedata_no0$birthrate.5)
summary(treedata_no0$birthrate.3)
summary(treedata_no0$deathrate.5)
summary(treedata_no0$deathrate.3)

# Median rate for branches with at least one event
summary(subset(treedata_no0, births.x>0))
summary(subset(treedata_no0, deaths.x>0))

ggplot(subset(treedata_no0, births.x>0), aes(birthrate.5)) +
  geom_density() 


```





## Birth vs. death rates
```{r}
ggplot(treedata_no0, aes(birthrate.5, deathrate.5)) + geom_point() + geom_smooth()
ggplot(treedata_no0, aes(birthrate.3, deathrate.3)) + geom_point() + geom_smooth()

# Ratio birth/death
ggplot(treedata_no0, aes(birthrate.5/deathrate.5)) + geom_histogram()

ggplot(treedata_no0, aes(birthrate.3/deathrate.3)) + geom_histogram()
```


# Model birth and death rates

Potential problems:
  - zero inflation in response
  - imbalance: copy numbers not equally distributed between lineages
  - collinearity: high copy number associated to some lineages

Solutions:
  - models taking into account zero inflation (ziformula). Or simpler: negative binomial 
  - do not add splines in main analysis, but discuss them

## Prepare data
```{r}
summary(treedata)

# Split L1 into 4 sublineages (doesn't work for internal nodes!!)
#table(subset(treedata, lineage=='L1')$sublineage)
#treedata$lineage_mod <- treedata$lineage
#l1_indices = which(treedata$lineage=='L1')
#treedata$lineage_mod[l1_indices] <- treedata$sublineage[l1_indices]

# Remove lineages with few observations
removelineage = which(table(treedata$lineage) < 100)
removelineage <- c(removelineage, c('L7', 'L9', 'LaX'))

treedata_no0 <- treedata_no0 %>%
  mutate(
    parent = as.factor(parent.x),
    node_type = as.factor(node_type),
    copy_number_s = as.numeric(scale(copy_number)),
    parent_CN_s = as.numeric(scale(parent_CN))   # standardize predictor
  )


# Plot response variable
ggplot(treedata_no0, aes(birthrate.5)) + geom_histogram()
ggplot(treedata_no0, aes(birthrate.3)) + geom_histogram() + scale_x_log10()

fig4a <- ggplot(treedata_no0, aes(birthrate.5)) + geom_histogram() + scale_x_log10() +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) +
  xlab('IS6110 birth rate (log scale)') + ylab('Count')

fig4a


# Create smaller dataset with variables used
treedata_sub <- subset(treedata_no0, !(lineage %in% removelineage) & !(is.na(lineage))) %>% droplevels()
treedata_sub <- treedata_sub[,c('births.x', 'deaths.x', 'births.y', 'deaths.y','branch.length', 'branch.length.gw','parent_CN', 'parent_CN_s', 'node_type', 'lineage', 'parent', 'birthrate.5')]
summary(treedata_sub)

treedata_sub$births <- as.integer(treedata_sub$births.x)
treedata_sub$deaths <- as.integer(treedata_sub$deaths.x)

# Protect against log(0) when using offset model (not necessary when removing 0-length branches)
#treedata_sub$branch_length2 <- pmax(treedata_sub$branch.length, 1e-6)

# Log branch length for offset
treedata_sub$log_branch_length <- log(treedata_sub$branch.length.gw)
```

## GLMM
Compare n models:
  - m1: glmm with parent node as random effect
  - m2: glmm with independent slopes and intercepts for lineages
  - m3: m2 with additional accounting for zero inflation
  - m4: m2 with splines for copy  number

### Birth
```{r}
library(glmmTMB)
library(ggeffects)
library(ggplot2)
library(DHARMa)
library(splines)


m1 <- glmmTMB(
  births ~ parent_CN + node_type + lineage + 
    (1 | parent) + 
    offset(log_branch_length),
  data = treedata_sub,
  family = nbinom2(),
)
summary(m1)


m2 <- glmmTMB(
  births ~ parent_CN + node_type + 
    (1 + parent_CN | lineage) + 
    (1 | parent) + 
    offset(log_branch_length),
  data = treedata_sub,
  family = nbinom2(),
)
summary(m2)

m2b <- glmmTMB(
  births ~ parent_CN + node_type + 
    (1 + parent_CN || lineage) + 
    (1 | parent) + 
    offset(log_branch_length),
  data = treedata_sub,
  family = nbinom2(),
)
summary(m2b)


# Correct for zero inflation? Not necessary when using negbinom2...
m3 <- glmmTMB(
  births ~ parent_CN + node_type + 
    (1 + parent_CN | lineage) + 
    (1 | parent) + 
    offset(log_branch_length),
  ziformula = ~ 1,
  data = treedata_sub,
  family = nbinom2(),
)
summary(m3)


# With splines for copy number, no lineage because too complex
m4 <- glmmTMB(
  births ~ ns(parent_CN, df = 4) + node_type + 
    (1 | parent) + 
    offset(log_branch_length),
  data = treedata_sub,
  family = nbinom2(link="log")
)
summary(m4)

# Fit power law
m5 <- glmmTMB(
  births ~ log(parent_CN + 1) + node_type +
    (1 + log(parent_CN + 1) | lineage) +
    (1 | parent) +
    offset(log_branch_length),
  family = nbinom2(),
  data = treedata_sub
)

m5_nodevar_nocor <- glmmTMB(
  births ~ log(parent_CN + 1) + node_type +
    (1 + log(parent_CN + 1) + node_type || lineage) +
    (1 | parent) +
    offset(log_branch_length),
  family = nbinom2(),
  data = treedata_sub
)

summary(m5_nodevar_nocor)

AIC(m1, m2, m3, m4, m5, m5_nodevar_nocor)

anova(m2,m5)

```


## Figure: effect sizes

### Estimate rate from model
To compare to the rates from Tanaka et al. 2004
```{r}
library(dplyr)

# Create a prediction dataset from your observed data
pred_dat <- treedata_sub %>%
  mutate(
    branch_length = exp(log_branch_length)
  )

# Predict expected births
pred_dat$mu <- predict(
  m5,
  newdata = pred_dat,
  type = "response",
  re.form = NA   # marginal over random effects
)

# Convert to per-element rates
pred_dat <- pred_dat %>%
  mutate(
    rate_per_branch = mu / branch_length,
    rate_per_element = rate_per_branch / parent_CN
  )

# Average per node type
pred_dat <- pred_dat %>% filter(parent_CN > 0)

# Plot distributions
ggplot(pred_dat, aes(rate_per_element, col=node_type)) + geom_density()


avg_rates <- pred_dat %>%
  group_by(node_type) %>%
  summarise(
    mean_rate = mean(rate_per_element),
    median_rate = median(rate_per_element)
  )

avg_rates


avg_rates_weighted <- pred_dat %>%
  group_by(node_type) %>%
  summarise(
    weighted_rate =
      sum(rate_per_element * branch_length) /
      sum(branch_length)
  )

avg_rates_weighted

```

Compare rates
```{r}
weighted_rate_terminal <- avg_rates_weighted$weighted_rate[2]
mean_CN <- mean(treedata_sub$parent_CN)

# births/copy/sub_per_genome → births/copy/sub_per_site
rate_per_copy_per_sub_per_genome <- weighted_rate_terminal / mean_CN
rate_per_copy_per_sub_per_site <- rate_per_copy_per_sub_per_genome / callable_genome_size

clock_rates      <- c(low = 1e-8, mid = 1e-7, high = 1e-6)
generations_per_year <- c(low = 300, mid = 365, high = 400)

results <- expand.grid(
    clock  = clock_rates,
    gen_yr = generations_per_year
  ) %>%
  mutate(
    # births/copy/sub/site ÷ (subs/site/year) = births/copy/year
    rate_per_copy_per_year = rate_per_copy_per_sub_per_site / clock,
    # births/copy/year ÷ (generations/year) = births/copy/generation  
    rate_per_copy_per_gen  = rate_per_copy_per_year / gen_yr
  )

results




```



### Example effects

```{r}
beta <- fixef(m5)$cond["log(parent_CN + 1)"]

# Function to get ratio
rr_cn <- function(c, beta) {
  ((c + 2) / (c + 1))^beta
}

# Confidence intervals
se <- sqrt(vcov(m5)$cond["log(parent_CN + 1)",
                         "log(parent_CN + 1)"])

# Wald for  CI
beta_ci <- beta + c(-1, 1) * 1.96 * se

# CI rate ratios
rr_ci <- function(c, beta_ci) {
  ((c + 2) / (c + 1))^beta_ci
}

rr_5_6  <- rr_cn(5, beta)
rr_5_6_ci  <- rr_ci(5, beta_ci)

rr_20_21 <- rr_cn(20, beta)
rr_20_21_ci <- rr_ci(20, beta_ci)

rr_30_31  <- rr_cn(30, beta)
rr_30_31_ci  <- rr_ci(30, beta_ci)
```




### Birth rate vs. copy number
```{r}
# Does birth rate depend on copy number?
y_raw <- treedata$birthrate.5
y_pos <- y_raw[is.finite(y_raw) & !is.na(y_raw) & (y_raw > 0)]
log_min <- floor(log10(min(y_pos)))
log_max <- ceiling(log10(max(y_pos)))
# major breaks: 10^n
major_breaks <- 10^(log_min:log_max)
# minor breaks: 1..9 * 10^n (this gives the conventional minor gridlines)
minor_breaks <- as.vector(outer(1:9, 10^(log_min:log_max), `*`))
# keep only minor breaks within the data range
minor_breaks <- minor_breaks[minor_breaks >= min(y_pos) & minor_breaks <= max(y_pos)]
  
x_levels <- levels(as.factor(treedata$copy_number))

treedata_no7 <- subset(treedata, !(lineage %in% removelineage))

figSx <- ggplot(na.omit(treedata_no7), aes(as.factor(copy_number), birthrate.5)) + 
  geom_boxplot(outlier.size = 0.3) + facet_wrap(~lineage) +
  xlab("Copy number") + ylab("IS6110 birth rate") + 
  scale_x_discrete(
    labels = ifelse(seq_along(x_levels) %% 5 == 1, x_levels, "")) + 
  scale_y_log10(breaks = major_breaks,minor_breaks = minor_breaks)+ 
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) +
  scale_y_log10()

ggsave("figures/supplement/figSx.cn_vs_birthrate.observed.png", figSx,width = 20, height = 18, units = "cm", dpi=300) 
```


### Copy number effect
```{r}
library(ggeffects)

cn_pop <- ggpredict(
  m5,
  terms = "parent_CN",
  type = "fixed", bias_correction = TRUE,
  condition = list(log_branch_length = 0))


ggplot(cn_pop, aes(x = x, y = predicted)) +
  geom_line(col='blue',size = 1) +
  geom_ribbon(aes(ymin = conf.low, ymax = conf.high), alpha = 0.2, fill = "blue") +
  labs(x = "Copy number", y = "Expected IS6110 birth rate") +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) +
  scale_y_log10()

```

Add lineage slopes and intercepts
```{r}
library(ggrepel)

# Add lineage slopes and intercepts
CN_seq <- seq(
  min(treedata_sub$parent_CN),
  max(treedata_sub$parent_CN),
  length.out = 100
)

newdat <- expand.grid(
  parent_CN = CN_seq,
  lineage = unique(treedata_sub$lineage),
  node_type = "internal",      # fix covariates
  parent = NA,
  log_branch_length = 1
)

newdat$predicted <- predict(
  m5,
  newdata = newdat,
  type = "response",
  re.form = NULL
)

library(dplyr)

line_labels <- newdat %>%
  group_by(lineage) %>%
  slice_max(parent_CN, n = 1)

panelC <- ggplot() +
  geom_line(
    data = cn_pop,
    aes(x = x, y = predicted),
    linewidth = 1
  ) +
  geom_ribbon(
    data = cn_pop,
    aes(x = x, ymin = conf.low, ymax = conf.high),
    alpha = 0.2
  ) +
  geom_line(
    data = newdat,
    aes(x = parent_CN, y = predicted, group = lineage),
    linetype = "dashed",
    alpha = 0.6
  ) +
  #geom_text_repel(
  #data = line_labels,
  #aes(x = parent_CN, y = predicted, label = lineage),
  #direction = "y",
  #hjust = 0,
  #segment.size = 0.2,
  #size = 3
  #) +
  labs(
    x = "Copy number",
    y = "Predicted IS6110 birth rate"
  ) +
  scale_y_log10() +
  coord_cartesian(clip = "off") +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12)

panelC

```


Lineage slope vs intercept
```{r}

re <- ranef(m2)$cond$lineage
re$lineage <- rownames(re)

fixef_CN <- fixef(m2)$cond["parent_CN"]
fixef_IC <- fixef(m2)$cond["(Intercept)"]

re$slope <- fixef_CN + re$parent_CN
head(re)



re <- ranef(m2)$cond$lineage
ggplot(re, aes(`(Intercept)`, parent_CN)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Lineage intercept (baseline birth rate)",
    y = "Lineage CN slope"
  ) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12)
```


### Branch type
```{r}
p_node <- ggpredict(
  m5,
  terms = "node_type",
  type = "fixed", bias_correction = TRUE,
  condition = list(log_branch_length = 0))

panelD <- ggplot(p_node, aes(x, predicted)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.15) +
  xlab("Node type") + 
  ylab("Predicted IS6110 birth rate") +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12)

panelD



```

Combine panels
```{r}
lower <- cowplot::plot_grid(panelC,panelD, nrow=1, labels=c('C', 'D'), label_fontfamily = 'CMU Sans Serif')


cowplot::plot_grid(upper, lower, nrow=2)

ggsave("figures/main/fig3.diversity_AND_birth_rate_model.png", width = 23, height = 20, units = "cm", dpi=300)



```

## Per-copy birth rate
```{r}
library(ggeffects)
library(dplyr)

# Create prediction grid: vary node_type, fix CN at representative values
pred_dat <- expand.grid(
  parent_CN   = seq(1, max(treedata_sub$parent_CN), by = 1),
  node_type   = c("internal", "terminal"),
  parent      = NA,                    # marginalise over random effect
  log_branch_length = 0                # offset = 0 → predict per unit branch length
)

pred_dat$predicted_births <- predict(
  m5,
  newdata  = pred_dat,
  type     = "response",
  re.form  = NA                        # population-level (fixed effects only)
)

# Per-copy rate = predicted births per unit branch length / CN
pred_dat <- pred_dat %>%
  mutate(per_copy_rate = predicted_births / parent_CN)

# Summarise: e.g. median per-copy rate at each CN, by node type
pred_dat %>%
  group_by(node_type) %>%
  summarise(
    median_per_copy_rate = median(per_copy_rate),
    mean_per_copy_rate   = mean(per_copy_rate)
  )

exp(fixef(m5)$cond["node_typeterminal"])

p <- ggpredict(
  m5,
  terms = c("parent_CN", "node_type"),
  type  = "fixed",
  bias_correction = TRUE,
  condition = list(log_branch_length = 0)
)

# Convert to per-copy
as.data.frame(p) %>%
  mutate(
    per_copy          = predicted / x,
    per_copy_low      = conf.low  / x,
    per_copy_high     = conf.high / x
  )



```

Visualize
```{r}
beta0 <- fixef(m5)$cond["(Intercept)"]
beta1 <- fixef(m5)$cond["log(parent_CN + 1)"]

cn_seq <- 1:max(treedata_sub$parent_CN)

analytical_df <- data.frame(
  CN = cn_seq,
  per_copy_rate = exp(beta0) * (cn_seq + 1)^beta1 / cn_seq
)

ggplot(analytical_df, aes(x = CN, y = per_copy_rate)) +
  geom_line(linewidth = 1) +
  scale_y_log10() +
  labs(
    x = "IS6110 copy number",
    y = "Per-copy IS6110 birth rate (log scale)"
  ) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12)
```


```{r}
# What is beta1 actually?
fixef(m5)$cond

# Does the within-lineage relationship look declining?
treedata_sub %>%
  filter(parent_CN > 0, births.x > 0) %>%
  group_by(lineage, parent_CN) %>%
  summarise(mean_rate = mean(birthrate.5)) %>%
  ggplot(aes(parent_CN, mean_rate)) +
    geom_point() + geom_smooth() +
    facet_wrap(~lineage, scales = "free") +
    scale_y_log10() +
    theme_bw()
```

# Test non-linear effect of CN
```{r}
treedata_l = subset(treedata, branch.length>0 & lineage %in% c('L4'))
treedata_l$log_branch_length <- log(treedata_l$branch.length.gw)

treedata_l <- treedata_l %>%
  mutate(
    parent = as.factor(parent.x),
    node_type = as.factor(node_type)
  )
treedata_l$births <- as.integer(treedata_l$births.x)

# Fit models
m2_l24 <- glmmTMB(
  births ~ parent_CN + node_type + 
    (1 | parent) + 
    offset(log_branch_length),
  data = treedata_l,
  family = nbinom2(link="log")
)

m2_l24_splines_df4 <- glmmTMB(
  births ~ ns(parent_CN, df = 4) + node_type + 
    (1 | parent) +
    offset(log_branch_length),
  data = treedata_l,
  family = nbinom2(link="log")
)

m2_l24_splines_df8 <- glmmTMB(
  births ~ ns(parent_CN, df = 8) + node_type + 
    (1 | parent.x) + 
    offset(log_branch_length),
    data = treedata_l,
  family = nbinom2(link="log")
)

AIC(m2_l24, m2_l24_splines_df4, m2_l24_splines_df8)

summary(m2_l24_splines_df4)

```

## Visualize!
```{r}
cn_splines <- ggpredict(
  m2_l24_splines_df4,
  terms="parent_CN [all]",
  type = "fixed", bias_correction = TRUE,
  condition = list(log_branch_length = 0))


# Use exposure that penalizes short branch lengths, as in the glm below 
library(dplyr)

observed_rates <- treedata_l %>%
  group_by(parent_CN) %>%
  summarise(
    total_births = sum(births),
    total_exposure = sum(branch.length.gw),
    rate = total_births / total_exposure,
    n_branches = n()
  )


# Combine with observed patterns!
fig4d <- ggplot() +
  geom_line(
    data=cn_splines, 
    aes(x = x, y = predicted), size = 1
    ) +
  geom_ribbon(
    data=cn_splines,
    aes(x = x, y = predicted, ymin = conf.low, ymax = conf.high), alpha = 0.2
    ) +
  geom_point(
    data = observed_rates,
    aes(x = parent_CN, y = rate, size = n_branches),
    color = "black", alpha = 0.5
  ) +
  geom_jitter(
    data=treedata_l, 
    aes(parent_CN, birthrate.5), alpha=0.2
    ) +
  xlab("Copy number") + ylab("Predicted IS6110 birth rate") + 
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) +
  scale_y_log10()

fig4d

```








```{r}
dat_l4 <- subset(treedata_sub, lineage == "L4")

# full linear predictor (log scale)
eta <- predict(m2_l24_splines_df4, type = "link")

# design matrix
X <- model.matrix(
  delete.response(terms(m2_l24_splines_df4)),
  data = dat_l4
)

# spline columns
spline_cols <- grep("^ns\\(parent_CN", colnames(X))

beta <- fixef(m2_l24_splines_df4)$cond
spline_lp <- as.numeric(X[, spline_cols] %*% beta[spline_cols])

# Pearson residuals
resid_p <- residuals(m2_l24_splines_df4, type = "pearson")

partial_resid <- spline_lp + resid_p

pr_df <- data.frame(
  parent_CN = dat_l4$parent_CN,
  partial_resid = partial_resid
)





cn_splines <- ggpredict(
  m2_l24_splines_df4,
  terms = "parent_CN [all]",
  type = "fixed",
  back.transform = FALSE,
  condition = list(log_branch_length = 0)
)


ggplot(pr_df, aes(x = parent_CN, y = partial_resid)) +
  geom_point(
    alpha = 0.15,
    size = 1,
    position = position_jitter(width = 0.15, height = 0)
  ) +
  geom_line(
    data = cn_splines,
    aes(x = x, y = predicted),
    linewidth = 1.2
  ) +
  geom_ribbon(
    data = cn_splines,
    aes(x = x, y = predicted, ymin = conf.low, ymax = conf.high),
    alpha = 0.25
  ) +
  scale_x_continuous(
    breaks = seq(0, 22, by = 2),
    limits = c(-0.5, 22.5)
  ) +
  labs(
    x = "IS6110 copy number (Lineage 4)",
    y = "Partial residual (log birth rate)"
  ) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12)

```

## Figure old: marginal effects
```{r}
library(ggeffects)
library(ggplot2)
library(patchwork)
library(dplyr)


# Branch length vs births
pA <- ggplot(treedata_sub, aes(branch.length, births)) + geom_point(alpha=0.2) + geom_smooth() + theme_minimal() +
  labs(
    x = "Branch length",
    y = "Number of IS6110 births"
  )


# Predict TE insertions across a range of parent_CN_s
pred_parentCN <- ggeffect(fit_offset, terms = c("parent_CN"), margin = 'marginalmeans', bias_correction = TRUE)

# Unscale CN for plotting
#mean_CN <- mean(treedata_sub$parent_CN)
#sd_CN <- sd(treedata_sub$parent_CN)
#pred_parentCN$x_unscaled <- pred_parentCN$x * sd_CN + mean_CN

# Convert predicted counts to birthrate by dividing by mean branch length and genome size to get rate per base pair
pred_parentCN$predicted_birthrate <- pred_parentCN$predicted / mean(treedata_sub$branch_length2) / genome_size
pred_parentCN$conf.low_birthrate <- pred_parentCN$conf.low / mean(treedata_sub$branch_length2) / genome_size
pred_parentCN$conf.high_birthrate <- pred_parentCN$conf.high / mean(treedata_sub$branch_length2) / genome_size


# Copy number effect
pB <- ggplot(pred_parentCN, aes(x = x, y = predicted_birthrate)) +
  geom_line(col='blue',size = 1) +
  geom_ribbon(aes(ymin = conf.low_birthrate, ymax = conf.high_birthrate), alpha = 0.2, fill = "blue") +
  labs(
    x = "Parent node copy number",
    y = "Expected IS6110 birth rate"
  ) +
  theme_minimal() 

pB


# Lineage effect
pred_lineage <- ggeffect(fit_offset, terms = "lineage", margin = 'marginalmeans', bias_correction = TRUE)

pred_lineage$predicted_birthrate <- pred_lineage$predicted / mean(treedata_sub$branch_length2) / genome_size
pred_lineage$conf.low_birthrate <- pred_lineage$conf.low / mean(treedata_sub$branch_length2) / genome_size
pred_lineage$conf.high_birthrate <- pred_lineage$conf.high / mean(treedata_sub$branch_length2) / genome_size

pC <- ggplot(pred_lineage, aes(x = x, y = predicted_birthrate)) +
  geom_point(size = 3, color = "darkgreen") +
  geom_errorbar(aes(ymin = conf.low_birthrate, ymax = conf.high_birthrate), width = 0.2) +
  labs(
    x = "Lineage",
    y = "Expected IS6110 birth rate"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))




pred_node <- ggeffect(fit_offset, terms = "node_type", margin = 'marginalmeans', bias_correction = TRUE)
pred_node$predicted_birthrate <- pred_node$predicted / mean(treedata_sub$branch_length2) / genome_size
pred_node$conf.low_birthrate <- pred_node$conf.low / mean(treedata_sub$branch_length2) / genome_size
pred_node$conf.high_birthrate <- pred_node$conf.high / mean(treedata_sub$branch_length2) / genome_size



pD <- ggplot(pred_node, aes(x = x, y = predicted_birthrate)) +
  geom_col(fill = "orange") +
  geom_errorbar(aes(ymin = conf.low_birthrate, ymax = conf.high_birthrate), width = 0.2) +
  labs(
    x = "Node type",
    y = "Expected IS6110 birth rate"
  ) +
  theme_minimal()

cowplot::plot_grid(pA,pB,pC,pD, nrow=2, labels=c('A', 'B', 'C', 'D'))

ggsave("figures/main/fig3.birthrate.png", width = 22, height = 16, units = "cm", dpi=300)    
```






## Model validation
```{r}

# Model validation
sim <- simulateResiduals(fit_offset, n = 2000)
plot(sim)

# Test overdispersion
testDispersion(sim)

# Test zero-inflation
testZeroInflation(sim)

# Check predictor residual patterns
plotResiduals(sim, treedata_sub$parent_CN_s)
plotResiduals(sim, treedata_sub$node_type)
plotResiduals(sim, treedata_sub$lineage)

# QQ to check Tweedie assumption
qqnorm(residuals(sim)); qqline(residuals(sim))
```