---
title: "IS6110 and chromosome organization"
output:
  html_document: 
    theme: simplex
date: "2025-03-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```



# Snakemake I/O
```{r}
paths <- tryCatch(
  {
    list(
      rdata = snakemake@input$rdata,
      output_html     = snakemake@output$html
    )
  },
  error = function(e){
    list(
      rdata = 'is6110.RData',
      output_html     = NULL  # or whatever default makes sense
    )
  }
)

```


# Insertion sites with mappable reference position
What is the proportion of IS that have a mappable reference position?

Do analyses with 5' matrix, as no significant differences are expected with the 3'.

Two observations to explain:
  - Reference insertions with same position but different strand
  - Anchors with multiple reference positions

```{r}
library(Biostrings)
library(dplyr)
library(ggplot2)
library(dplyr)
library(ggwordcloud)


# Number of anchors with reference position -----------------------------------
table(matrix_metadata$side)
with_ref_pos = which(!is.na(matrix_metadata$ref_position))
table(matrix_metadata[with_ref_pos,]$side)

prop_with_ref = table(matrix_metadata[with_ref_pos,]$side) / table(matrix_metadata$side)

print(paste('Proportion of insertion sites with mappable reference position: ', prop_with_ref[1], prop_with_ref[2]))
```

# Non-mappable insertion sites
Blast non-mappable flanking regions against all genic and intergenic regions in MTBCO to get an idea what they are.

```{r}
source('workflow/notebooks/utils.R')

# Get unmapped sites
nonmappable = subset(matrix_metadata,is.na(ref_position))

# Blast flanking sequences
matrix_metadata <- add_blast_info(matrix_metadata)


# Summarize by anchor and feature
hits_d = list()
feature_count = list()

for (i in 1:nrow(nonmappable)){
  row = nonmappable[i,]  
  query = paste(row$side, row$site_id, sep='_')
  
  hit = row$blasthit
  if (is.na(hit)){next}

  pref = substr(hit,1,2)
  if (pref=='mt'){categ = get_alternative_gene_name(hit, genemap)}
  else if (pref=='ig'){categ='intergenic'}
  else if (pref=='id'){categ='repeat'}
  else if (pref=='ge'){categ='pseudogene'}

  hits_d[[query]] = categ

  if (is.null(feature_count[[categ]])) {feature_count[[categ]] <- c()}
  feature_count[[categ]] <- c(feature_count[[categ]], query)
  
}

hits_d

feature_count = lapply(feature_count, unique)
lapply(feature_count, length)

word_counts = data.frame(
  'feature' = names(feature_count),
  'count' = unlist(unname(lapply(feature_count, length)))
)

head(word_counts)

# Word cloud
ggplot(subset(word_counts, count>2), aes(label = feature, size = count)) +
  geom_text_wordcloud(shape='square') +
  scale_size_area(max_size = 20) +
  theme_minimal()


```


# Circos plot: overview of insertion sites
- add positions/density of DeJesus' essential genes
- add oriC location

```{r}
library(ape)
library(magrittr)
library(ggplot2)
library(dplyr)
library(Cairo)
library(circlize)
library(ComplexHeatmap)


custom_order = c('L1', 'L2', 'L3', 'L4', 'L5', 'L6', 'L7', 'L8', 'L9', 'L10','La1', 'La2', 'La3')

# Sort according to the phylogenetic tree!
refins_ordered <- subset(refins, lineage %in% custom_order) %>% mutate(column_name = factor(strain, levels = rev(tree$tip.label))) %>% arrange(column_name)
head(refins_ordered)


# Reduce number of strains: Get n per lineage
n = 500
strains_reduced = c()
for (lineage in rev(custom_order)){
  strains_lineage = unique(refins_ordered$strain[refins_ordered$lineage==lineage])
  if (length(strains_lineage) == 0){next}
  if (length(strains_lineage) <= n){strains_reduced = c(strains_reduced, strains_lineage)}
  else {
    strains_sample = sample(strains_lineage, n)
    strains_reduced = c(strains_reduced, strains_sample)
  }
}
length(strains_reduced)

# Create list of bed files, per strain, for circos
bed_list = list()
colori = c()

refins_reduced <- subset(refins_ordered, strain %in% strains_reduced)
head(refins_reduced)

strains_ordered <- unique(refins_reduced$strain)

for (i in 1:length(strains_ordered)){
  
  df <- subset(refins_reduced, strain==strains_ordered[i])
  bed <- data.frame(
    'chr' = df$chromosome,
    'start' = df$position,
    'end' = df$position+1,
    'lineage' = df$lineage
  )
  
  bed_list[[i]] <- bed
  
  # Now the lineage color
  lineage =df$lineage[1]
  col = mtb_col_scale[lineage]
  colori <- c(colori, col)
}

length(bed_list)
length(colori)

genome = data.frame(chr  = "MTBC0", start = 1,end   = 4435783) 


# Define color legend
lineages = names(mtb_col_scale)
lineages = lineages[lineages %in% unique(refins$lineage)]
colorrs = mtb_col_scale[lineages]

lgd = Legend(
  at = lineages, type = "lines", 
  legend_gp = gpar(col = colorrs, lwd = 2), 
  title_gp = gpar(fontfamily = "CMU Sans Serif", fontsize = 10),
  labels_gp = gpar(fontfamily = "CMU Sans Serif", fontsize = 10),
  title_position = "topleft", 
  title = "Lineage", break_dist = 2)


# Plot insertions!
options(fig.width = 15, fig.height = 15, repr.plot.width = 15, repr.plot.height = 15)
par(bg = "white")

CairoPDF("figures/main/fig4a.circos.pdf",  width = 10, height = 10)

circlize_plot(genome, bed_list, colori)
draw(lgd, x = unit(10, "mm"), y = unit(20, "mm"), just = c("left", "bottom"))

dev.off()

```


# Distance effect
```{r}
library(ggplot2)

bin_sizes = c(50000, 100000, 200000, 500000)
chrom_length = 4435783

# Use positions from 5' matrix
matrix_metadata_five <- subset(matrix_metadata, side==5)
positions <- na.omit(matrix_metadata_five$ref_position)

dist_from_ori = c()

for (pos in positions){
  dist = dist_to_oriC(pos)
  dist_from_ori = c(dist_from_ori, dist)
}


bin_size = 100000
n_breaks = as.integer(chrom_length/bin_size, 0)
bins <- hist(dist_from_ori, breaks=n_breaks, plot = F)

bins_df <- data.frame(
  'mids' = bins$mids,
  'counts' = bins$counts,
  'density' = bins$density
)

cor.test(bins_df$mids, bins_df$counts, method = 'pearson')


# Plot histogram and density curve with two y axes
df <- data.frame(dists = dist_from_ori / 1000)

# parameters
bins <- 50
n <- nrow(df)
bw <- (max(df$dists) - min(df$dists)) / bins  # explicit bin width

fig4a <- ggplot(df, aes(x = dists)) +

  # histogram in counts (use explicit binwidth for precise scaling)
  geom_histogram(binwidth = bw, fill = "lightgray", color = NA) +

  # density scaled to count-scale using after_stat()
  geom_density(aes(y = after_stat(density * n * bw)),
               color = "red", linewidth = 1) +

  # left axis = counts; right axis = true density (inverse transform)
  scale_y_continuous(
    name = "Count",
    sec.axis = sec_axis(~ . / (n * bw), name = "Density")
  ) +

  labs(x = "Distance from oriC (kb)") +
  theme_bw(base_family = "CMU Sans Serif")

fig4a

```



# Strand bias
```{r}
summary(as.factor(matrix_metadata$ref_strand))

```




# Symmetry analysis
Is the distribution of IS6110 more symmetrical than expected by chance?
If yes, this might reflect large-scale structural adaptations in the chromosome
to mitigate the effects of inversions/deletions (Rocha 2004)

## Functions
```{r}
angles_from_positions <- function(positions, genome_length) {
  (2 * pi * (positions %% genome_length) / genome_length) %% (2*pi)
}


symmetry_stat_for_axis <- function(angles, weights = NULL, axis, m = 72) {
  if (is.null(weights)) weights <- rep(1, length(angles))
  
  breaks <- seq(0, 2*pi, length.out = m+1)
  
  # Count in bins
  counts1 <- hist(angles, breaks = breaks, plot = FALSE, include.lowest = TRUE,
                  right = FALSE, weights = weights)$counts
  # Count in mirrored bins
  counts2 <- hist((2*axis - angles) %% (2*pi), breaks = breaks, plot = FALSE,
                  include.lowest = TRUE, right = FALSE, weights = weights)$counts
  
  sum((counts1 - counts2)^2)
}

find_best_axis <- function(angles, weights = NULL, n_axes = 180, m = 72) {
  axes <- seq(0, 2*pi, length.out = n_axes)
  stats <- sapply(axes, function(phi) symmetry_stat_for_axis(angles, weights, phi, m))
  idx <- which.min(stats)
  list(best_axis = axes[idx], best_stat = stats[idx],
       axes = axes, stats = stats)
}
```


```{r}
library(ggplot2)
library(dplyr)

head(matrix_metadata)
chrom_length = 4435783
oriC_start=1349

# Data sets: all and minimum frequencies of 2, 10 and 100

best_axes = data.frame()

for (minimum in c(1,2,10,100)){
  for (binsize in c(50000,100000,200000)){

    sites_sub = subset(matrix_metadata, !is.na(ref_position) & nr_strains > minimum & side == 5)
    sites_sub <- sites_sub[order(sites_sub$ref_position), ]
  
    positions = sites_sub$ref_position
    weights = sites_sub$nr_strains
    angles <- angles_from_positions(positions, chrom_length)

    nbins = genome_size / binsize
    
    res = find_best_axis(angles,weights=weights, n_axes=360, m=nbins)
  
    df = data.frame(
      'condition' = as.factor(rep(minimum,length(res$stats))),
      'binsize' = as.factor(rep(binsize, length(res$stats))),
      'axis' = res$axes,
      'stats' = res$stats
    )
  
    best_axes <- rbind(best_axes, df)
  }

}

best_axes$position <- best_axes$axis * chrom_length / (2*pi)


ggplot(best_axes, aes(position, stats)) + geom_line() +
  facet_wrap(binsize~condition, scales = 'free_y')

ggplot(best_axes, aes(condition, stats, col=binsize)) + geom_boxplot() 

# How do the minima compare to the full distribution of symmetry statistics?
ggplot(best_axes, aes(stats)) + geom_density() + facet_wrap(binsize~condition, scales='free')


# Is the lowest statistic an outlier?
best_axes %>% group_by(condition, binsize) %>% summarise(MinValue = min(stats))




```


## Figure

```{r}
# Symmetry statistic scan
plot_axis_scan(res, chrom_length)

# Plot observed vs mirrored bin counts
plot_paired_bins(angles, weights, best_axis = res$best_axis, m = 36)

plot_circular_histogram(angles, weights = NULL, m = 72, res$best_axis, genome_length = chrom_length)

# Compare observed and permutation stats
perm_df = data.frame(permutations = res$perm_stats)
ggplot(perm_df, aes(permutations)) + 
  geom_density() +
  geom_vline(xintercept=res$best_stat, col='red', linetype='dashed') 
```



# Motif analysis
Create a sequence logo plot for the ~10 bases adjoining all insertions.
Also try different subsets: e.g. genic vs. intergenic.
```{r}

library(ggplot2)
library(Biostrings)
library(ggseqlogo)
library(cowplot)

fasta <- list(
  five = readDNAStringSet("workflow/results/detettore/ALL_presence-absence.5prime.fasta"),
  three = readDNAStringSet("workflow/results/detettore/ALL_presence-absence.3prime.fasta")
)

get_logo <- function(seqs, n_bases, where){
  subseqs = list()
  for (i in 1:length(seqs)) {
    seq = seqs[i]
    if (where == 'end'){
      end = width(seq)
      sseq = subseq(seq, start=end-(n_bases-1), end=end)
    }
    else if (where == 'start'){
      sseq = subseq(seq, start=1, end=n_bases)
    }
    subseqs = c(subseqs, unname(as.character(sseq)))
  }
  subseqs = unlist(subseqs)
  ggplot() + geom_logo(subseqs, method='prob') + theme_logo()

}

a <- get_logo(fasta$five, 10, 'end')
b <- get_logo(fasta$three, 10, 'start')

plot_grid(a,b)

```


# Genic vs. intergenic
Overrepresentation in intergenic as evidence for purifying selection. 
Also, a higher proportion of genic insertions are expected in singleton insertions. 

Figure: 

```{r}
library(ggplot2)

# Use 5' matrix
matrix_metadata_five <- subset(matrix_metadata, side==5)


# Add genic/intergenic category to matrix metadata (careful with NA values!!)
matrix_metadata_five$context_cat <- rep('genic', nrow(matrix_metadata_five))
matrix_metadata_five$context_cat[grepl(';', matrix_metadata_five$context)] <- 'intergenic'
matrix_metadata_five$context_cat[which(is.na(matrix_metadata_five$context))] <- NA
matrix_metadata_five$context_cat <- as.factor(matrix_metadata_five$context_cat)
table(matrix_metadata_five$context_cat)

# Assign frequency category singleton/shared
matrix_metadata_five$freqcat <- rep('singleton', nrow(matrix_metadata_five))
matrix_metadata_five$freqcat[which(matrix_metadata_five$nr_strains>1)] <- 'shared'
matrix_metadata_five$freqcat <- as.factor(matrix_metadata_five$freqcat)
table(matrix_metadata$freqcat)

# Are intergenic insertions overrepresented?
gff_path <- 'data/MTBC0/MTBC0v1.1_PGAP_annot.gff'
genic_observed <- unname(table(matrix_metadata_five$context_cat)['genic'])
intergenic_observed <- unname(table(matrix_metadata_five$context_cat)['intergenic'])
chromosome_length = 4435783

fisher_test(genic_observed, intergenic_observed,chromosome_length, gff_path)


# What about singleton vs. shared elements?
fig4b <- ggplot(na.omit(matrix_metadata_five), aes(context_cat, fill=freqcat)) + 
  geom_histogram(stat='count', position='dodge') + 
  xlab("") + ylab("Count") +
  theme_bw(base_family = "CMU Sans Serif", base_size=14) +
  scale_fill_brewer(palette='Dark2') +
  guides(fill = guide_legend(title = NULL)) +   # remove legend title
  theme(
    legend.position = c(0.98, 0.98),            # move legend inside (normalized coordinates)
    legend.justification = c("right", "top"),   # align legend anchor
    legend.background = element_rect(fill = "white", color = NA),  # optional
    legend.key.size = unit(0.5, "cm")            # optional
  )
fig4b


# Hypothesis: more singleton insertions in genes because selection has not removed them
singletons <- subset(matrix_metadata_five, freqcat=='singleton')
shared <- subset(matrix_metadata_five, freqcat=='shared')

fisher_test(
  unname(table(singletons$context_cat)['genic']),
  unname(table(singletons$context_cat)['intergenic']), 
  chromosome_length,
  gff_path
)

ratio_singletons = unname(table(singletons$context_cat)['intergenic']) /(unname(table(singletons$context_cat)['genic']) + unname(table(singletons$context_cat)['intergenic']))


fisher_test(
  unname(table(shared$context_cat)['genic']),
  unname(table(shared$context_cat)['intergenic']),
   chromosome_length,
  gff_path
)

ratio_shared = unname(table(shared$context_cat)['intergenic']) / (unname(table(shared$context_cat)['genic']) + unname(table(shared$context_cat)['intergenic']))

```



Combine panels
```{r}
library(cowplot)
library(magick)

img_circors <- image_read("figures/main/fig4a.circos.pdf", density = 600)
p_circos  <- ggdraw() + draw_image(img_circors)

plot_grid(p_circos, fig4b, 
          labels = c('A', 'B'),
          label_size = 10,
          nrow=1, 
          rel_widths = c(2,1), 
          rel_heights=c(2,1),
          label_fontfamily='CMU Sans Serif')

```









# A natural mutagenesis experiment
- add frequency aspect: shared vs. singletons
```{r}
source('workflow/notebooks/utils.R')

matrix_metadata_five <- subset(matrix_metadata, side==5)
head(matrix_metadata_five)

knockouts = subset(matrix_metadata_five, context_cat == 'genic')
intergenic = subset(matrix_metadata_five, context_cat == 'intergenic')

print(paste("Nr. of genes with insertions:", length(unique(knockouts$context))))
print(paste("Nr. of intergenic regions with insertions:", length(unique(intergenic$context))))

ko_genes = unique(knockouts$context)
ko_intergenic = unique(intergenic$context)



# How many genes/intergenic regions affected?
context_df = data.frame()
context_l = list()

for (feature in unique(matrix_metadata_five$context)){
  
  feature_df = subset(matrix_metadata_five, context == feature)
  nr_pos = length(unique(feature_df$ref_position))
  nr_strains = sum(feature_df$nr_strains)
  #nr_lineages = length(unique(feature_df$lineage))
  #resistance = feature_df$resistance[1]
  
  if (grepl(';', feature)){cat = 'intergenic'}
  else {cat = 'knockout'}

  row = data.frame(feature, nr_pos, nr_strains, cat)
  context_df = rbind(context_df, row)
}

head(context_df)
dim(context_df)

dim(subset(context_df, cat=='knockout'))
dim(subset(context_df, cat=='intergenic'))


# Frequency of gene disruption?
head(context_df)
ggplot(context_df, aes(as.factor(nr_strains))) + geom_histogram(stat='count') +
  facet_wrap(~cat)

table(context_df$nr_strains, context_df$cat)

# Table 2: the 20 most frequently disrupted regions



```

## Essentiality
How do genes never affected by IS6110 compare to in vitro essential genes, as defined by De Jesus et al. 2017?

Expectation: Many more genes should be needed in clinico!

How many of the knockouts occur only once, how many in multiple strains?

Are knockouts clade-specific, or are some genes affected throughout the MTBC?
```{r}
library(ggplot2)

# Table S3 in De Jesus et al.
orfs_dejesus <- read.csv('manuscript/literatura/DeJesus2017/orfs.csv', sep= '\t')
table(orfs_dejesus$Final.Call)

# Gene essentiality form DeJesus is present in the genemap table!



match(knockouts$context, genemap$mtbc0_gene_id)
match(knockouts$context, genemap$gene_id)




# Nr. of interrupted genes: 870
refins_dejesus <- merge(refins, orfs_dejesus, by.x='context', by.y="ORF.ID")
refins_dejesus_ne <- subset(refins_dejesus, Final.Call=='NE')

# NE in DeJesus that do not have IS6110 insertions
ne_genes = orfs_dejesus$ORF.ID[which(orfs_dejesus$Final.Call=='NE')]
length(ne_genes)



setdiff(unique(refins_ko$gene_rv), ne_genes)

table(refins_dejesus$Final.Call)

refins_dejesus_es <- subset(refins_dejesus, Final.Call=='ES')
length(unique(refins_dejesus_es$context))

# Plot essentiality categories
ggplot(refins_dejesus, aes(Final.Call)) + geom_histogram(stat='count')
```


## Accumulation analysis
Given the limited sample size, how many new genes are knocked out when considering new strains? 
Do this per lineage! And for intergenic regions.

Also: exclude singleton insertions

```{r}
library(vegan)
library(ggplot2)

knockout_m = matrix(nrow=length(ko_genes), ncol=length(metadata$GNUMBER))
rownames(knockout_m) <- ko_genes
colnames(knockout_m) <- metadata$GNUMBER

for (i in 1:length(metadata$GNUMBER)){
  s = metadata$GNUMBER[i]
  knockout_s = subset(knockouts, strain==s)
  knockout_m[,s] <- ifelse(ko_genes %in% knockout_s$gene, 1, 0)
}

# Get accumulation data separately for lineages
acc_df = data.frame()

for (lineage in lineage_order){
  
  gnrs = metadata$GNUMBER[which(metadata$LINEAGE_x==lineage)]
  if (length(gnrs) < 100){next}
  
  m_l = knockout_m[,gnrs]
  m_l <- m_l[rowSums(m_l) != 0,]
  m_lt <- t(m_l)
  
  # Species accumulation curve (random permutations)
  acc <- specaccum(m_lt, method="random", permutations=100)

  df = data.frame(
    sites = acc$sites,
    richness = acc$richness,
    sd = acc$sd,
    perm = acc$perm,
    lineage = rep(lineage,length(acc$sites))
  )
  acc_df = rbind(acc_df, df)
}

summary(as.factor(acc_df$lineage))

col_schemes <- load_color_schemes()

ggplot(subset(acc_df, lineage %in% c('L1', 'L2', 'L3', 'L4', 'L5', 'L6')), aes(x=sites, y=richness, col=lineage)) + geom_line(alpha=1, size=1)+
  geom_line(aes(y=richness-sd,col=lineage), linetype='dashed',size=0.2) +
  geom_line(aes(y=richness+sd,col=lineage), linetype='dashed', size=0.2) +
  xlab("Sampled strains") + ylab("Knocked-out genes") +
  scale_color_manual(values=col_schemes$col_scale_black) +
  theme_bw()


ggsave('/home/cristobal/TB/projects/MTBC_IS6110-poly/figures/figX.accumulaiton_curves.pdf', width=14, height = 10, units = 'cm')

# Model nr. of k.o. genes at 500 strains ~ lineage diversity * mean IS CN
subset(acc_df, sites==500)


```


