---
title: "Save genomic niches for IS6110"
output:
  html_document: 
    theme: simplex
date: "2025-03-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


# Snakemake I/O
```{r}
paths <- tryCatch(
  {
    list(
      rdata = snakemake@input$rdata,
      output_html     = snakemake@output$html
    )
  },
  error = function(e){
    list(
      rdata = 'is6110.RData',
      output_html     = NULL  # or whatever default makes sense
    )
  }
)

```

# Insertion sites with mappable reference position
What is the proportion of IS that have a mappable reference position?

Do analyses with 5' matrix, as no significant differences are expected with the 3'.

Two observations to explain:
  - Reference insertions with same position but different strand
  - Anchors with multiple reference positions

```{r}
library(Biostrings)
library(dplyr)
library(ggplot2)
library(dplyr)
library(ggwordcloud)

load(paths$rdata)

# Number of anchors with reference position -----------------------------------
table(matrix_metadata$side)
with_ref_pos = which(!is.na(matrix_metadata$ref_position))
table(matrix_metadata[with_ref_pos,]$side)

prop_with_ref = table(matrix_metadata[with_ref_pos,]$side) / table(matrix_metadata$side)

print(paste('Proportion of insertion sites with mappable reference position: ', prop_with_ref[1], prop_with_ref[2]))
```


## Explore non-mappable insertion sites
Blast non-mappable flanking regions against all genic and intergenic regions in MTBCO to get an idea what they are.

```{r}
source('workflow/notebooks/utils.R')

# Get unmapped sites
nonmappable = subset(matrix_metadata,is.na(ref_position))

# Blast flanking sequences
matrix_metadata <- add_blast_info(matrix_metadata)

# Summarize by anchor and feature
hits_d = list()
feature_count = list()

for (i in 1:nrow(nonmappable)){
  row = nonmappable[i,]  
  query = paste(row$side, row$site_id, sep='_')
  
  hit = row$blasthit
  if (is.na(hit)){next}

  pref = substr(hit,1,2)
  if (pref=='mt'){categ = get_alternative_gene_name(hit, genemap)}
  else if (pref=='ig'){categ='intergenic'}
  else if (pref=='id'){categ='repeat'}
  else if (pref=='ge'){categ='pseudogene'}

  hits_d[[query]] = categ

  if (is.null(feature_count[[categ]])) {feature_count[[categ]] <- c()}
  feature_count[[categ]] <- c(feature_count[[categ]], query)
  
}

hits_d

feature_count = lapply(feature_count, unique)
lapply(feature_count, length)

word_counts = data.frame(
  'feature' = names(feature_count),
  'count' = unlist(unname(lapply(feature_count, length)))
)

head(word_counts)

# Word cloud
ggplot(subset(word_counts, count>2), aes(label = feature, size = count)) +
  geom_text_wordcloud(shape='square') +
  scale_size_area(max_size = 20) +
  theme_minimal()


```


# Genic vs. intergenic insertions
Overrepresentation in intergenic as evidence for purifying selection. 
Also, a higher proportion of genic insertions are expected in singleton insertions. 

```{r}
library(ggplot2)

# Use 5' matrix
matrix_metadata_five <- subset(matrix_metadata, side==5)


# Add genic/intergenic category to matrix metadata (careful with NA values!!)
matrix_metadata_five$context_cat <- rep('genic', nrow(matrix_metadata_five))
matrix_metadata_five$context_cat[grepl(';', matrix_metadata_five$context)] <- 'intergenic'
matrix_metadata_five$context_cat[which(is.na(matrix_metadata_five$context))] <- NA
matrix_metadata_five$context_cat <- as.factor(matrix_metadata_five$context_cat)
table(matrix_metadata_five$context_cat)

# Assign frequency category singleton/shared
matrix_metadata_five$freqcat <- rep('singleton', nrow(matrix_metadata_five))
matrix_metadata_five$freqcat[which(matrix_metadata_five$nr_strains>1)] <- 'shared'
matrix_metadata_five$freqcat <- as.factor(matrix_metadata_five$freqcat)
table(matrix_metadata$freqcat)

# Are intergenic insertions overrepresented?
gff_path <- 'data/MTBC0/MTBC0v1.1_PGAP_annot.gff'
genic_observed <- unname(table(matrix_metadata_five$context_cat)['genic'])
intergenic_observed <- unname(table(matrix_metadata_five$context_cat)['intergenic'])
chromosome_length = 4435783

fisher_test(genic_observed, intergenic_observed,chromosome_length, gff_path)


# What about singleton vs. shared elements?
fig4b <- ggplot(na.omit(matrix_metadata_five), aes(context_cat, fill=freqcat)) + 
  geom_histogram(stat='count', position='dodge') + 
  xlab("") + ylab("Count") +
  theme_bw(base_family = "CMU Sans Serif", base_size=14) +
  scale_fill_brewer(palette='Dark2') +
  guides(fill = guide_legend(title = NULL)) +   # remove legend title
  theme(
    legend.position = c(0.98, 0.98),            # move legend inside (normalized coordinates)
    legend.justification = c("right", "top"),   # align legend anchor
    legend.background = element_rect(fill = "white", color = NA),  # optional
    legend.key.size = unit(0.5, "cm")            # optional
  )
fig4b


# Hypothesis: more singleton insertions in genes because selection has not removed them
singletons <- subset(matrix_metadata_five, freqcat=='singleton')
shared <- subset(matrix_metadata_five, freqcat=='shared')

fisher_test(
  unname(table(singletons$context_cat)['genic']),
  unname(table(singletons$context_cat)['intergenic']), 
  chromosome_length,
  gff_path
)

ratio_singletons = unname(table(singletons$context_cat)['intergenic']) /(unname(table(singletons$context_cat)['genic']) + unname(table(singletons$context_cat)['intergenic']))


fisher_test(
  unname(table(shared$context_cat)['genic']),
  unname(table(shared$context_cat)['intergenic']),
   chromosome_length,
  gff_path
)

ratio_shared = unname(table(shared$context_cat)['intergenic']) / (unname(table(shared$context_cat)['genic']) + unname(table(shared$context_cat)['intergenic']))

```


Combine panels
```{r}
library(cowplot)
library(magick)


system("pdfcrop --margins 5 figures/main/fig4a.circos.pdf figures/main/fig4a.circos.cropped.pdf")

img_circors <- image_read("figures/main/fig4a.circos.cropped.pdf", density = 600)
p_circos  <- ggdraw() + draw_image(img_circors)

left_panel <- plot_grid(
  fig4a, fig4b, 
  labels=c('B','C'),
  label_size = 10,
  ncol=2,
  label_fontfamily='CMU Sans Serif'
)

right_panel <- plot_grid(
  p_circos,
  labels=c('A'),
  label_size = 10,
  ncol=1,
  label_fontfamily='CMU Sans Serif'
)

plot_grid(right_panel, left_panel, 
          nrow=2, 
          rel_widths = c(3,1), 
          rel_heights=c(3,1))


ggsave('figures/main/fig5.circos_niches.png', width = 18, height = 22, units = "cm", dpi = 300) 


```


# Natural Mutagenesis experiment

## Gene knockouts
- add frequency aspect: shared vs. singletons
```{r}
source('workflow/notebooks/utils.R')

head(refins)

knockouts = subset(refins, context=='genic')
intergenic = subset(refins, context=='intergenic')

print(paste("Nr. of genes with insertions:", length(unique(knockouts$gene))))
print(paste("Nr. of intergenic regions with insertions:", length(unique(intergenic$gene))))

ko_genes = unique(knockouts$gene)
ko_intergenic = unique(intergenic$gene)


# How many genes/intergenic regions affected?
context_df = data.frame()
context_l = list()

for (feature in unique(refins$gene)){
  
  feature_df = subset(refins, gene == feature)
  nr_pos = length(unique(feature_df$position))
  nr_lineages = length(unique(feature_df$lineage))
  nr_strains = length(unique(feature_df$strain))
  #resistance = feature_df$resistance[1]
  
  if (grepl(';', feature)){cat = 'intergenic'}
  else {cat = 'knockout'}

  row = data.frame(feature, nr_pos, nr_lineages, nr_strains, cat)
  context_df = rbind(context_df, row)
}

head(context_df)
dim(context_df)

dim(subset(context_df, cat=='knockout'))
dim(subset(context_df, cat=='intergenic'))

```


## Accumulation analysis
Given the limited sample size, how many new genes are knocked out when considering new strains? 
Do this per lineage! And for intergenic regions.

Also: exclude singleton insertions

```{r}
library(vegan)
library(ggplot2)

knockout_m = matrix(nrow=length(ko_genes), ncol=length(metadata$GNUMBER))
rownames(knockout_m) <- ko_genes
colnames(knockout_m) <- metadata$GNUMBER

for (i in 1:length(metadata$GNUMBER)){
  s = metadata$GNUMBER[i]
  knockout_s = subset(knockouts, strain==s)
  knockout_m[,s] <- ifelse(ko_genes %in% knockout_s$gene, 1, 0)
}

# Get accumulation data separately for lineages
acc_df = data.frame()

for (lineage in lineage_order){
  
  gnrs = metadata$GNUMBER[which(metadata$LINEAGE_x==lineage)]
  if (length(gnrs) < 100){next}
  
  m_l = knockout_m[,gnrs]
  m_l <- m_l[rowSums(m_l) != 0,]
  m_lt <- t(m_l)
  
  # Species accumulation curve (random permutations)
  acc <- specaccum(m_lt, method="random", permutations=100)

  df = data.frame(
    sites = acc$sites,
    richness = acc$richness,
    sd = acc$sd,
    perm = acc$perm,
    lineage = rep(lineage,length(acc$sites))
  )
  acc_df = rbind(acc_df, df)
}

summary(as.factor(acc_df$lineage))

col_schemes <- load_color_schemes()

ggplot(subset(acc_df, lineage %in% c('L1', 'L2', 'L3', 'L4', 'L5', 'L6')), aes(x=sites, y=richness, col=lineage)) + geom_line(alpha=1, size=1)+
  geom_line(aes(y=richness-sd,col=lineage), linetype='dashed',size=0.2) +
  geom_line(aes(y=richness+sd,col=lineage), linetype='dashed', size=0.2) +
  xlab("Sampled strains") + ylab("Knocked-out genes") +
  scale_color_manual(values=col_schemes$col_scale_black) +
  theme_bw()


ggsave('/home/cristobal/TB/projects/MTBC_IS6110-poly/figures/figX.accumulaiton_curves.pdf', width=14, height = 10, units = 'cm')

# Model nr. of k.o. genes at 500 strains ~ lineage diversity * mean IS CN
subset(acc_df, sites==500)


```



## Essentiality
How do genes never affected by IS6110 compare to in vitro essential genes, as defined by De Jesus et al. 2017?

Expectation: Many more genes should be needed in clinico!

How many of the knockouts occur only once, how many in multiple strains?

Are knockouts clade-specific, or are some genes affected throughout the MTBC?
```{r}
library(ggplot2)

# Table S3 in De Jesus et al.
orfs_dejesus <- read.csv('manuscript/literatura/DeJesus2017/orfs.csv', sep= '\t')


# Nr. of interrupted genes: 870
refins_dejesus <- merge(refins, orfs_dejesus, by.x='context', by.y="ORF.ID")
refins_dejesus_ne <- subset(refins_dejesus, Final.Call=='NE')

# NE in DeJesus that do not have IS6110 insertions
ne_genes = orfs_dejesus$ORF.ID[which(orfs_dejesus$Final.Call=='NE')]
length(ne_genes)



setdiff(unique(refins_ko$gene_rv), ne_genes)

table(refins_dejesus$Final.Call)

refins_dejesus_es <- subset(refins_dejesus, Final.Call=='ES')
length(unique(refins_dejesus_es$context))

# Plot essentiality categories
ggplot(refins_dejesus, aes(Final.Call)) + geom_histogram(stat='count')
```



# Hotspots
Count how many times genic and intergenic regions were hit. 
For the niche model: can we obtain an estimate of how much of the genome is permissible?

Supplement: 
  - figure with frequency of independend insertion sites/births per region
  - supplementary table with hotspot summary
Main:
  - figure 6


## Extend region annotation
For the regions to be comparable to previously described hotspots and H37Rv orthologs, we generate a region table that indicates orthologs as well as insertion sequence annotations.

Detettore6110 reports the gene name rather than the gene ID for facilitating exploration. 
Not ideal for parsing, this will be changed in a future version. 

```{r}
head(genemap)


regionmap = read.csv("data/MTBC0/MTBC.genomic_regions.tsv", sep='\t')
head(regionmap)


get_alt_gene_name <- function(gene_name, genemap){
  " Get gene ID if present else mtbc0_gene_id
  "
  row = genemap[genemap$mtbc0_gene_id==gene_name, ]
  if (is.na(row$gene_id)){return(gene_name)}
  else if (nchar(row$gene_id)==0){return(gene_name)}
  else {return(row$gene_id)}
}

detettore_name = c()

for (i in 1:nrow(regionmap)){
  region_name = regionmap$region_name[i]
  typus = regionmap$region_type[i]

  if (typus != 'intergenic'){detettore_name = c(detettore_name, region_name)}

  else {
    region_split = strsplit(region_name, ';') %>% unlist()
    geneA = region_split[1]
    geneB = region_split[2]

    geneA_new = get_alt_gene_name(geneA, genemap)
    geneB_new = get_alt_gene_name(geneB, genemap)

    detettore_name = c(detettore_name, paste(geneA_new, geneB_new, sep=';'))
  }

}

regionmap$detettore_name = detettore_name


write.table(regionmap,"data/MTBC0/MTBC.genomic_regions.extended.tsv", sep='\t', quote=F, row.names=F)

```



## Summarize hotspots
Approach: create strainXregion matrix, reconstruct ancestral states, and count 0->presence transitions. 

```{r}
load(paths$rdata)

strains = tree$tip.label
regions = unique(refins$gene)

# Create empty matrix with strains in rows and regions in columns
regionmat = matrix(
  nrow=length(strains), 
  ncol=length(regions),
  dimnames=list(strains, regions)
  )

# Collect information for hotspot summary
npos = list()

# Fill matrix
for (i in 1:nrow(refins)){

  if (i%%1000==0){print(i)}

  row = refins[i,]
  strain = row$strain
  region = row$gene
  position = row$position
  strand = row$strand

  regionmat[strain,region] = 1

  # GEt the corresponding uFS for a reference insertion
  #if (strand == '+'){ua = subset(anchor_map, V1==row$strain & #V2==row$anchor_5)$V3}
  #else {ua = subset(anchor_map, V1==row$strain & V2==row$anchor_3)$V3}
  #if (length(ua) == 0){next}

  # Add uFS id to matrix
  #if (is.na(regionmat[strain,region])){regionmat[strain,region] = ua}
  #else {
  #  existing = regionmat[strain,region]
  #  regionmat[strain,region] = paste(existing, ua, sep='_')
  #}

  # Collect different positions per region
  if (!(region %in% names(npos))){
    npos[[region]] = c(position)
    }
  else {
    npos[[region]] = c(npos[[region]], position)
    }
}

# Convert NA to 0
regionmat[is.na(regionmat)] <- 0

```


## Reconstruct ancestral states and count number of births.
Summary structure: region, nstrains, nbirths, ndeaths, npositions.
Here one could also look at transitions from one uFS to another, which could indicate errors in detettore due to mutations happening after the insertion of IS6110. 

```{r}
library(ape)
library(castor)
library(magrittr)

hotspot_summary = data.frame()

for (i in 1:ncol(regionmat)){

  if (i%%100==0){print(i)}
  
  site_vec <- regionmat[, i]
  site_id = colnames(regionmat)[i]

  # Skip constant or NA-only sites
  if (any(is.na(site_vec)) || length(unique(site_vec)) == 1) next
      
  # Convert to numeric starting with 1
  site_vec_mapped <- map_to_state_space(site_vec)
  index2name = names(site_vec_mapped$name2index)
  index2name <- setNames(index2name, unname(site_vec_mapped$name2index))

  # Reconstruct ancestral states
  asr_result <- asr_max_parsimony(tree, tip_states = site_vec_mapped$mapped_states)

  # Go through tree and record transitions
  ntips = length(tree$tip.label)
  # Nr. of strains that contain an insertion in the region
  nstrains = sum(site_vec_mapped$mapped_states!=1) 
  # Number of different uFS
  #nstates = site_vec_mapped$Nstates - 1

  # For each node, store likelihood of presence
  lh_presence <- asr_result$ancestral_likelihoods[,2]
  # Combine tip and internal node states
  full_states <- c(unname(site_vec), lh_presence)

  # Count number of independent birth/insertion events
  nbirths = 0
  ndeaths = 0

  for (j in 1:nrow(tree$edge)) {
        
    parent = tree$edge[j, 1]
    child  = tree$edge[j, 2]

    parent_state = full_states[parent]
    child_state = full_states[child]
    
    # Get ancestral states at child node
    #if (child <= ntips){
    #  child_node = site_vec_mapped$mapped_states[child]
    #  child_state = ifelse(child_node==1, 'absent', 'present')
    #  }
    #else{
    #  child_node = asr_result$ancestral_likelihoods[child-ntips,]
    #  child_state = ifelse(child_node[1]>0.5, 'absent', 'present')    
    #  }
    # At parent node
    #parent_node = asr_result$ancestral_likelihoods[parent-ntips,]
    #parent_state = ifelse(parent_node[1]>0.5, 'absent', 'present')

    # No change
    if (parent_state == child_state) next
    # Birth
    if (parent_state==0 & child_state==1){nbirths = nbirths+1}
    # Death
    else if (parent_state==1 & child_state==0){ndeaths = ndeaths+1}
  }

  row = data.frame(
    'region' = site_id,
    'nstrains' = nstrains,
    'nbirths' = nbirths,
    'ndeaths' = ndeaths
  )
  print(row)
  hotspot_summary = rbind(hotspot_summary, row)
}

write.table(hotspot_summary, "workflow/results/hotspot_summary.simplified.tsv", sep='\t', quote=F, row.names = F)

```


## Add Rv ortholog names
If there is a gene ID, use it, else use Rv ortholog name. 
If there is no Rv ortholog, use mtbc0 locus tag.

```{r}
library(magrittr)

load(paths$rdata)

hotspot_summary = read.csv("workflow/results/hotspot_summary.simplified.tsv", sep='\t')
mtbc0_regions = read.csv('data/MTBC0/MTBC.genomic_regions.tsv', sep='\t')

head(hotspot_summary)


get_alternative_gene_name <- function(genename, genemap){
  
  pref = substr(genename, 1,5)
  
  if (pref=='mtbc0'){
    if (!(genename %in% genemap$mtbc0_gene_id)){return(genename)}
    else {gene_id = genemap$gene_id[which(genemap$mtbc0_gene_id==genename)]}
    
    if (nchar(gene_id) == 0){
      alt_name = genemap$h37rv_id[which(genemap$mtbc0_gene_id==genename)]
    }
    else {alt_name = gene_id}
  }
  else {
    alt_name = genename
    #rv = genename
    #rv = genemap$h37rv_id[which(genemap$gene_id==genename)]
  }
  
  #if (length(rv) > 1){
  #  rv = rv[[1]]
  #  print(paste('Two genemap entries for ', genename, sep=" "))
  #}
  if (length(alt_name) > 0){return(alt_name)}
  else if (length(alt_name) == 0){return('None')}
}


alt_gene_names = c()

for (region in hotspot_summary$region){
  
  if (grepl(';', region)){ # intergenic
    
    region_split = strsplit(region, ';') %>% unlist()
    
    geneA = region_split[1]
    geneB = region_split[2]
    
    geneA_new = get_alternative_gene_name(geneA, genemap)
    geneB_new = get_alternative_gene_name(geneB, genemap)
    region_new = paste(geneA_new, geneB_new, sep = ';')
    alt_gene_names = c(alt_gene_names, region_new)
  }
  
  else { # genic
    gene_new = get_alternative_gene_name(region, genemap)
    alt_gene_names = c(alt_gene_names, gene_new)
  }
}

hotspot_summary$region_rv = alt_gene_names
write.table(hotspot_summary, "workflow/results/hotspot_summary.tsv", sep='\t', quote=F, row.names = F)


```


## FIGURE 4
```{r}
## Figure: hotspot overview + exemplar trees

library(ggplot2)
library(ggrepel)
library(magrittr)
library(ggtree)
library(tidytree)
library(cowplot)
library(magick)
library(Cairo)
library(dplyr)

load(paths$rdata)

hotspot_summary = read.csv("workflow/results/hotspot_summary.tsv", sep='\t')
regionmap = read.csv("data/MTBC0/MTBC.genomic_regions.extended.tsv", sep='\t')


head(hotspot_summary)
head(regionmap)


# Detettore regions that are not in the region map
setdiff(hotspot_summary$region, regionmap$detettore_name)

# Regions that are not in detettore output
setdiff(regionmap$detettore_name,hotspot_summary$region)


# Load ancestral states for tree visualization
anc_states_five <- read.csv(
  'workflow/results/detettore/AS_5prime.ancestral_likelihoods.tsv', sep = '\t')
names(anc_states_five) <- gsub('X', '', names(anc_states_five))


# ── Panel A: linear genome × nbirths ─────────────────────────────────────────

# Attach reference position to hotspots_summary via hotspots
region_pos <- hotspots %>%
  group_by(region) %>%
  summarise(ref_position = median(as.numeric(position), na.rm = TRUE), .groups = 'drop')

hs <- hotspot_summary %>%
  left_join(region_pos, by = 'region') %>%
  arrange(desc(nbirths))

top50 <- slice_head(hs, n = 20)

panelAx <- ggplot(
  subset(hs, nbirths>10), 
  aes(x = ref_position / 1e6, y = nbirths, size=nstrains)
  ) +
  geom_point(alpha = 0.4, colour = 'grey40') +
  geom_label_repel(
    data       = top50,
    aes(label  = region_rv),
    size       = 3,
    label.size = 0,
    max.overlaps = 40,
    segment.colour = 'grey60',
    segment.size   = 0.3
  ) +
  scale_x_continuous(
    name   = 'Genomic position (Mb)',
    limits = c(0, 4.44),
    breaks = seq(0, 4, 0.5)
  ) +
  ylab('Inferred births') +
  theme_bw(base_family = 'CMU Sans Serif', base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = c(0.98, 0.98),
    legend.justification = c(1, 1),
    strip.background = element_blank()
  )

panelAx


# More informative: nstrains vs. nbirths!
panelA <- ggplot() + 
  geom_jitter(
    data=subset(hotspot_summary, nbirths>1), 
    aes(nbirths,nstrains), 
    alpha=0.2, size=2) +
  geom_label_repel(
    data=subset(hotspot_summary, nstrains>500 | nbirths>100), 
    aes(nbirths, nstrains, label=region_rv), 
    label.size=0, 
    size = 3) + 
  theme_bw(base_family = "CMU Sans Serif", base_size = 14) +
  ylab("Nr. of strains") + xlab("Inferred births (log scale)") +
  scale_x_log10()

panelA





# ── Panel B: exemplar hotspot trees ──────────────────────────────────────────

# Choose which hotspots to show: top N by nbirths (skip any missing from ASR)
n_trees   <- 12          # set to 6 or 9
top_spots <- hs %>%
  filter(region %in% colnames(anc_states_five) |
           region %in% hotspots$region) %>%
  slice_head(n = n_trees)

p_base <- ggtree(tree, layout = 'circular', col = 'grey90',
                 size = 0.1, alpha = 0.8) %<+% metadata

tree_pdfs <- character(n_trees)

for (i in seq_len(nrow(top_spots))) {

  focal      <- top_spots$region[i]
  label_name <- if (!is.na(top_spots$region_rv[i])) top_spots$region_rv[i] else focal
  nb         <- top_spots$nbirths[i]

  hotspots_sub <- subset(hotspots, region == focal)
  site_ids     <- hotspots_sub$site_id[hotspots_sub$site_id %in% names(anc_states_five)]
  if (length(site_ids) == 0) next

  anc_states_sub <- anc_states_five[, c('node', site_ids), drop = FALSE]
  prob_mat       <- as.matrix(anc_states_sub[, -1, drop = FALSE])

  max_prob  <- apply(prob_mat, 1, max)
  hit_col   <- apply(prob_mat, 1, which.max)

  anc_state_compressed <- data.frame(
    node           = anc_states_sub$node,
    insertion_site = ifelse(max_prob > 0.5,
                            colnames(prob_mat)[hit_col],
                            NA_character_)
  )

  pd <- p_base$data
  pd$insertion_site <- anc_state_compressed$insertion_site[
    match(pd$node, anc_state_compressed$node)]

  n_sites <- length(unique(na.omit(pd$insertion_site)))
  pal     <- scales::hue_pal()(max(n_sites, 1))

  p_tree <- p_base +
    geom_tree(
      data = subset(pd, !is.na(insertion_site)),
      #aes(colour = insertion_site),
      size = 0.35
    ) +
    #scale_colour_manual(values = pal) +
    labs(caption = sprintf('%s', label_name)) +
    theme(
      legend.position  = 'none',
      plot.caption = element_text(size = 7, hjust = 0.5, margin = margin(t = -35)),
      plot.caption.position = 'plot', 
      plot.title      = element_text(size = 7, hjust = 0.5),
      plot.background = element_rect(fill = NA, color = NA),
      panel.background = element_rect(fill = NA, color = NA)
    )

  # Save individual tree to temp PDF, then crop
  tmp_raw    <- tempfile(fileext = '.pdf')
  tmp_crop   <- tempfile(fileext = '.pdf')

  ggsave(tmp_raw, plot = p_tree,
         width = 6, height = 6, units = 'cm', device = cairo_pdf)
  system(paste('pdfcrop', tmp_raw, tmp_crop))

  tree_pdfs[i] <- tmp_crop
}

# Read cropped PDFs back as ggdraw panels
tree_panels <- lapply(tree_pdfs[nzchar(tree_pdfs)], function(f) {
  img <- image_read(f, density = 250)
  ggdraw() + draw_image(img)
})

n_cols   <- 4
n_rows   <- ceiling(length(tree_panels) / n_cols)
panelB   <- plot_grid(plotlist = tree_panels, ncol = n_cols, nrow = n_rows)
panelB

# ── Combine and save ──────────────────────────────────────────────────────────

full_fig <- plot_grid(
  panelA, panelB,
  nrow        = 2,
  rel_heights = c(1.8, n_rows * 0.9),
  labels      = c('A', 'B'),
  label_fontfamily = 'CMU Sans Serif',
  label_size  = 12
)

ggsave('figures/main/fig_hotspots.png',
       plot   = full_fig,
       width  = 30, height = 10 + n_rows * 6,
       units  = 'cm',
       dpi = 300)

full_fig
```


## TABLE 2: top hotspots
Table header: hotspot name, nbirths, nstrains, Rv orthologs, Products

```{r}
library(gt)
library(magrittr)
library(tidyverse)
library(ggplot2)


head(hotspot_summary)

# Sort table and pick top 12 hotspots
toptwelve <- hotspot_summary[order(hotspot_summary$nbirths, decreasing = T),][1:12,]


```

## Characterize hotspot regions
```{r}
library(ggplot2)

head(hotspot_summary2)

hotspot_summary2$type = rep('genic', nrow(hotspot_summary2))
hotspot_summary2$type [grepl(';', hotspot_summary2$region)] <- 'intergenic'


dim(hotspot_summary2)
table(hotspot_summary2$nbirths)
table(hotspot_summary2$nstrains)

# Check single disruptions
ggplot(subset(hotspot_summary2, nbirths==1), aes(nstrains)) + 
  geom_histogram()

table(subset(hotspot_summary2, nbirths==1)$nstrains)

# Proportion of regions with < 10 disruptions
nrow(subset(hotspot_summary2, nbirths<10))

nrow(subset(hotspot_summary2, nbirths>20))
nrow(subset(hotspot_summary2, nbirths>50))


# Frequency of disruptions
ggplot(hotspot_summary, aes(nbirths)) + geom_histogram()
ggplot(hotspot_summary, aes(nbirths, nstrains)) + geom_point() + geom_smooth()

cor(hotspot_summary$nbirths, hotspot_summary$nstrains)


```







```{r}


summary(subset(hotspots_summary, nbirths>1))
summary(subset(hotspots_summary, nbirths>10))
summary(subset(hotspots_summary, nbirths>20))
summary(subset(hotspots_summary, nbirths>50))


ggplot(subset(hotspots_summary, nbirths>1),aes(x=nbirths)) + geom_step(aes(y=..y..),stat="ecdf")

ggplot(subset(hotspots_summary, nbirths>1),aes(x=nbirths)) + geom_density()


```


## Gene set enrichment

```{r}
# ============================================================
# GSEA of IS6110 insertion hotspots using GFF GO annotations
# ============================================================
library(ggplot2)
library(dplyr)
library(tidyr)
library(stringr)
library(clusterProfiler)  # or fgsea

# ============================================================
# 1. Parse GO terms from GFF
# ============================================================

parse_gff_go <- function(gff_path) {
  lines <- readLines(gff_path)
  # Keep only CDS lines
  cds_lines <- lines[grepl("\tCDS\t", lines)]
  
  result <- lapply(cds_lines, function(line) {
    fields <- strsplit(line, "\t")[[1]]
    attr  <- fields[9]
    
    # Extract locus_tag
    locus <- str_match(attr, "locus_tag=([^;]+)")[,2]
    
    # Extract GO terms from Ontology_term
    go_raw <- str_match(attr, "Ontology_term=([^;]+)")[,2]
    if (is.na(go_raw)) return(NULL)
    
    go_terms <- str_split(go_raw, ",")[[1]]
    go_terms <- go_terms[grepl("^GO:", go_terms)]
    
    if (length(go_terms) == 0) return(NULL)
    
    data.frame(gene = locus, go_id = go_terms, stringsAsFactors = FALSE)
  })
  
  bind_rows(result)
}

go_annotations <- parse_gff_go("data/MTBC0/MTBC0v1.1_PGAP_annot.gff")

# Build term2gene and term2name for clusterProfiler
term2gene <- go_annotations[, c("go_id", "gene")]

# Optional: fetch GO descriptions (requires AnnotationDbi / GO.db)
# library(GO.db)
# term2name <- AnnotationDbi::select(GO.db, keys=unique(term2gene$go_id),
#                                    columns="TERM", keytype="GOID") %>%
#   rename(go_id = GOID, name = TERM)

# ============================================================
# 2. Prepare hotspot ranking
# ============================================================
# Assumes hotspots_summary has columns: region, nbirths
# 'region' is a gene locus_tag (genic) or "geneA;geneB" (intergenic)

hotspots_summary <- read.csv("workflow/results/hotspot_summary.tsv", sep = "\t")

# Keep only genic hotspots (intergenic regions have ';')
genic_hotspots <- hotspots_summary %>%
  filter(!grepl(";", region)) %>%
  arrange(desc(nbirths))


ggplot(genic_hotspots, aes(nbirths)) + geom_histogram() + scale_x_log10()





# ============================================================
# 3a. Over-representation analysis (ORA) on top hotspots
# ============================================================

hotspot_threshold = 10

gene_list_top  <- subset(genic_hotspots, nbirths >= hotspot_threshold)$region
gene_universe  <- genic_hotspots$region  # all tested genes

ora_result <- enricher(
  gene      = gene_list_top,
  universe  = gene_universe,
  TERM2GENE = term2gene,
  # TERM2NAME = term2name,   # uncomment if you have GO names
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  qvalueCutoff  = 0.2
)

summary(ora_result)
dotplot(ora_result)

# ============================================================
# 3b. GSEA (ranked, uses all genes — more powerful)
# ============================================================

ranked_vec <- setNames(genic_hotspots$nbirths, genic_hotspots$region)

gsea_result <- GSEA(
  geneList      = sort(ranked_vec, decreasing = TRUE),
  TERM2GENE     = term2gene,
  # TERM2NAME   = term2name,
  pAdjustMethod = "BH",
  pvalueCutoff  = 0.05,
  minGSSize     = 5,
  maxGSSize     = 500,
  seed          = 42
)

dotplot(gsea_result)
gseaplot2(gsea_result, geneSetID = 1)  # plot top hit

# ============================================================
# 4. Alternative: use go_function / go_process free-text field
#    (richer but needs mapping to GO IDs)
# ============================================================

parse_gff_go_verbose <- function(gff_path) {
  lines <- readLines(gff_path)
  cds_lines <- lines[grepl("\tCDS\t", lines)]
  
  lapply(cds_lines, function(line) {
    attr  <- strsplit(line, "\t")[[1]][9]
    locus <- str_match(attr, "locus_tag=([^;]+)")[,2]
    
    go_fn  <- str_match(attr, "go_function=([^;]+)")[,2]
    go_pro <- str_match(attr, "go_process=([^;]+)")[,2]
    
    # Parse: "description|GOID||evidence"
    parse_go_field <- function(field) {
      if (is.na(field)) return(NULL)
      entries <- str_split(field, ",")[[1]]
      go_ids  <- str_match(entries, "\\|(GO:\\d+)\\|")[,2]
      go_ids[!is.na(go_ids)]
    }
    
    go_terms <- c(parse_go_field(go_fn), parse_go_field(go_pro))
    if (length(go_terms) == 0) return(NULL)
    data.frame(gene = locus, go_id = go_terms, stringsAsFactors = FALSE)
  }) %>% bind_rows()
}

# Use this instead of (or combined with) parse_gff_go() above
go_annotations_v2 <- parse_gff_go_verbose("MTBC0v1.1_PGAP_annot.gff")
```



## Plot hotspot summary
```{r}
# How do measures of parallelism compare?
ggplot(subset(hotspots_summary, nbirths>1), aes(nsites, API)) + geom_point(alpha=0.6) +
  geom_label_repel(data=subset(hotspots_summary, API>20), aes(label=region), label.size=0, size = 3) +
  theme_bw() + 
  xlab("Number of independent insertion sites") + ylab("Parallelism index") +
  scale_color_viridis(option='A', end = 0.8)

ggsave('/home/cristobal/TB/projects/MTBC_IS6110-poly/presentations/labmeeting.21-10-2025/figures/adaptive_parallelism.pdf', width=22, height = 17, units = 'cm')



# Nr independent insertion sites vs. nr independent insertions
panelA <- ggplot(subset(hotspots_summary), aes(strains_total, nbirths)) + geom_point(alpha=0.4) +
  geom_label_repel(data=subset(hotspots_summary, nbirths>100 | strains_total>1500), aes(label=region), label.size=0, size = 3) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 14) +
  xlab("Nr. of strains") + ylab("Nr. of independent births") +
  theme(
    legend.position = c(1, 0),              # Bottom right inside the plot
    legend.justification = c("right", "bottom"),
    legend.background = element_rect(fill = NA, color = NA),
    legend.key.height = unit(1, "cm"),          # Smaller legend key
    legend.key.width = unit(2, "cm"),
    legend.text = element_text(size = 12),         # Smaller text
    legend.title = element_text(size = 12)
  ) +
    scale_size(name = "Nr. of strains")

panelA

# Distribution of PI/nsites etc
table(hotspots_summary$nsites)

ggplot(hotspots_summary, aes(nsites)) + geom_histogram() +  theme_bw() + xlab('Number of independent insertion sites') + ylab("Count")

ggplot(hotspots_summary, aes(nbirths)) + geom_histogram() +  theme_bw() + xlab('Number of independent insertion events') + ylab("Count")


# Show both in one
library(reshape2)
melt(hotspots_summary, id.vars=c('region'), measure.vars = c('nsites', 'nbirths')) %>%
  ggplot(aes(value, fill=variable)) + geom_histogram(position='dodge', bins=20) + 
  theme_bw(base_family = "CMU Sans Serif", base_size = 14) +
  scale_fill_brewer(palette='Dark2') +
  theme(
    legend.position = c(1, 1),              # Bottom right inside the plot
    legend.justification = c("right", "top"),
    legend.background = element_rect(fill = NA, color = NA),
    legend.key.height = unit(1, "cm"),          # Smaller legend key
    legend.key.width = unit(1, "cm"),
    legend.text = element_text(size = 12),         # Smaller text
    legend.title = element_text(size = 12)
  )


ggsave('/home/cristobal/TB/projects/MTBC_IS6110-poly/presentations/labmeeting.21-10-2025/figures/n_indepentend_sites_dist.pdf', width=15, height = 10, units = 'cm')
```
 

## Plot hotspot on tree
```{r}
library(ggplot2)
library(magrittr)
library(ggtree)
library(tidytree)
library(RColorBrewer)
library(ggrepel)
library(viridis)

hotspot_cnt = as.data.frame(table(hotspots$region))

# Get ancestral states. For region of interest, add states to tree
anc_states_five = read.csv('workflow/results/detettore/AS_5prime.ancestral_likelihoods.tsv', sep='\t')
names(anc_states_five) <- gsub('X', '', names(anc_states_five))

p <- ggtree(tree, layout="circular", col='grey90',size=0.1, alpha=0.8) %<+% metadata


# example: mtbc0_001868 (plcD)
hotspots_of_interest = c('dnaA;dnaN', 'mtbc0_000805;phoP', 'mtbc0_001868','mtbc0_001879;mtbc0_001880','mtbc0_002031')


focal = 'ctpD'
focal='mtbc0_003539;mtbc0_003540'
focal='mtbc0_000845;mtbc0_000846'
hotspots_sub = subset(hotspots, region == focal)  
anc_states_sub = anc_states_five[,c('node', hotspots_sub$site_id)]

# Are there strains with multiple insertions in the same region? Yes
table(rowSums(anc_states_sub[,-1]))


# Create two new columns for the tree: insertion_site, births
anc_state_compressed <- data.frame(
  'node' = anc_states_sub$node,
  'insertion_site' = rep(NA, nrow(anc_states_sub))
)

for (site_id in hotspots_sub$site_id){
  anc_state_compressed$insertion_site[which(anc_states_sub[[site_id]]>0.5)] <- site_id
}

p$data$insertion_site = anc_state_compressed$insertion_site[match(p$data$node, anc_state_compressed$node)]

p + geom_tree(data=subset(p$data, !(is.na(insertion_site))), size=0.3) + theme(legend.position = "none")

ggsave("figures/main/fig6b.plcD_hotspot.pdf", 
       width = 13, height = 13, units = "cm",
       device = cairo_pdf)      # full-vector

system("pdfcrop --margins 5 figures/main/fig6b.plcD_hotspot.pdf figures/main/fig6b.plcD_hotspot.cropped.pdf")


img_tree <- image_read("figures/main/fig6b.plcD_hotspot.cropped.pdf", density = 300)
panelB  <- ggdraw() + draw_image(img_tree)

lower = cowplot::plot_grid(panelB, panelB)


cowplot::plot_grid(panelA, lower, nrow=2, rel_heights = c(2,1))


# Save tree to file and crop white space
#dest = paste("presentations/labmeeting.21-10-2025/figures/", focal, '.pdf')
#ggsave(dest, width = 15, height = 15, units = "cm", device = cairo_pdf) 


```


### Write hotspot trees to pdf
```{r}
library(ggplot2)
library(magrittr)
library(ggtree)
library(tidytree)
library(cowplot)
library(Cairo)
library(dplyr)

load(paths$rdata)

# Load ancestral states
anc_states_five <- read.csv(
  'workflow/results/detettore/AS_5prime.ancestral_likelihoods.tsv', sep = '\t')
names(anc_states_five) <- gsub('X', '', names(anc_states_five))

# Pick top 20 hotspots by nbirths
top20 <- hotspot_summary2 %>%
  arrange(desc(nbirths)) %>%
  slice_head(n = 40)

# Base tree (circular, tip labels off)
p_base <- ggtree(tree, layout = "circular", col = 'grey90', size = 0.1, alpha = 0.8) %<+% metadata

# Build one plot per hotspot
tree_plots <- lapply(seq_len(nrow(top20)), function(i) {
  
  focal       <- top20$region[i]
  label_name  <- if (!is.na(top20$region[i])) top20$region[i] else focal
  nb          <- top20$nbirths[i]
  
  hotspots_sub <- subset(hotspots, region == focal)
  
  # Guard: skip if no site columns found
  site_ids <- hotspots_sub$site_id
  site_ids <- site_ids[site_ids %in% names(anc_states_five)]
  if (length(site_ids) == 0) return(NULL)
  
  anc_states_sub <- anc_states_five[, c('node', site_ids), drop = FALSE]
  prob_mat       <- as.matrix(anc_states_sub[, -1, drop = FALSE])
  
  # Assign each node to the highest-probability site (if any site > 0.5)
  max_prob  <- apply(prob_mat, 1, max)
  hit_col   <- apply(prob_mat, 1, which.max)
  
  anc_state_compressed <- data.frame(
    node           = anc_states_sub$node,
    insertion_site = ifelse(
      max_prob > 0.5,
      colnames(prob_mat)[hit_col],
      NA_character_
    )
  )
  
  # Merge into tree data
  pd <- p_base$data
  pd$insertion_site <- anc_state_compressed$insertion_site[
    match(pd$node, anc_state_compressed$node)]
  
  n_sites <- length(unique(na.omit(pd$insertion_site)))
  pal     <- scales::hue_pal()(max(n_sites, 1))
  
  p_out <- p_base +
    geom_tree(
      data   = subset(pd, !is.na(insertion_site)),
      #aes(col = insertion_site),
      size   = 0.4
    ) +
    #scale_color_manual(values = pal) +
    ggtitle(sprintf("%s\n(births = %f)", label_name, nb)) +
    theme(
      legend.position  = "none",
      plot.title       = element_text(
        size   = 7,
        hjust  = 0.5,
        margin = margin(b = 2)
      ),
      plot.margin = margin(3, 3, 3, 3)
    )
  
  p_out
})

# Remove NULLs (sites with no ASR data)
tree_plots <- Filter(Negate(is.null), tree_plots)

# Arrange into pages of 6 (2 cols × 3 rows)
plots_per_page <- 6
n_pages        <- ceiling(length(tree_plots) / plots_per_page)

CairoPDF("figures/supplement/figSx.top20_hotspots_on_tree.pdf",
         width = 14, height = 21)   # A4-ish portrait

for (pg in seq_len(n_pages)) {
  
  idx   <- ((pg - 1) * plots_per_page + 1) : min(pg * plots_per_page, length(tree_plots))
  page_plots <- tree_plots[idx]
  
  # Pad to full grid if last page is short
  while (length(page_plots) < plots_per_page) {
    page_plots <- c(page_plots, list(ggplot() + theme_void()))
  }
  
  grid <- plot_grid(
    plotlist = page_plots,
    ncol     = 2,
    nrow     = 3
  )
  
  print(grid)
}

dev.off()
```




## Subtree for exploration: PPE34 insertions in L2
Check out hotspot regions where there might be issues with uFS multiplication 
through post-insertion mutations.
```{r}
load(paths$rdata)

# Get ancestral states. For region of interest, add states to tree
anc_states_five = read.csv('workflow/results/detettore/AS_5prime.ancestral_likelihoods.tsv', sep='\t')
names(anc_states_five) <- gsub('X', '', names(anc_states_five))
```


```{r}
library(ape)
library(ggtree)
library(tidytree)
library(treeio)

lineage = 'L4'
r = 'mtbc0_002993;cas2'
#r = 'mtbc0_002031'

mrca_node = mrcas$node[which(mrcas$lineage==lineage)]
strains = metadata$GNUMBER[which(metadata$lineage==lineage)]

p <- ggtree(tree, col='grey80',size=0.5, alpha=0.8) %<+% metadata

# Add region info vectorized
hotspots_sub = subset(hotspots, region == r)
anc_states_sub = anc_states_five[,c('node', hotspots_sub$site_id)]
prob_mat <- as.matrix(anc_states_sub[, -1])

# find column index of first site > 0.5 per row
hit_index <- max.col(prob_mat > 0.5, ties.method = "first")

# rows with no hits
no_hit <- rowSums(prob_mat > 0.5) == 0

# build compressed dataframe
anc_state_compressed <- data.frame(
  node = anc_states_sub$node,
  insertion_site = colnames(prob_mat)[hit_index]
)

anc_state_compressed$insertion_site[no_hit] <- NA

# Are there strains with multiple insertions in the same region? Yes
table(rowSums(anc_states_sub[,-1]))

anc_state_long <- as.data.frame(which(prob_mat > 0.5, arr.ind = TRUE))
anc_state_long$node <- anc_states_sub$node[anc_state_long$row]
anc_state_long$insertion_site <- colnames(prob_mat)[anc_state_long$col]
anc_state_long <- anc_state_long[, c("node", "insertion_site")]


p$data$insertion_site = anc_state_compressed$insertion_site[match(p$data$node, anc_state_compressed$node)]

p +
  geom_tree(data=subset(p$data, !(is.na(insertion_site))), aes(col=insertion_site),size=0.3) +
  geom_zoom_clade(node = mrca_node) +
  theme(legend.position = "none")
  #geom_hilight(node = mrca_node)

# Which uFS are these?
sort(table(p$data$insertion_site[which(p$data$label %in% strains)]))

```

Extract uFS from fasta in order to align them
```{r}
library(Biostrings)
ufs_fasta = 'workflow/results/detettore/ALL_unique_anchors.fasta'

fasta_file <- readDNAStringSet(ufs_fasta)



# List of patterns to grep for
patterns = unique(p$data$insertion_site[which(p$data$label %in% strains)])[-2]

# Subset the sequences based on the patterns
subset_seqs <- lapply(patterns, function(pattern) {
  fasta_file[grep(pattern = pattern, x = names(fasta_file))]
})

# Concatenate the subset sequences into a single FASTA file
combined_fasta <- DNAStringSet()
for (i in seq_along(subset_seqs)) {
  combined_fasta <- c(combined_fasta, subset_seqs[[i]])
}

# Write the combined sequences to a single FASTA file
writeXStringSet(combined_fasta, "explorations/detettore/breakpoint_drift/5.uFS.mtbc0_002993-cas2.fasta")


```



## Plot gene models and mutations
```{r}
library(ggplot2)

head(genemap)
head(hotspot)

focal = 'mtbc0_000805;phoP'

# Get mutations
hotspot_sub = subset(hotspots, region == focal)
hotspot_sub$position <- as.integer(hotspot_sub$position)

# Create data frame with gene coordinates
focal_splt = strsplit(focal, ';') %>% unlist()
geneA = focal_splt[1]
geneB = focal_splt[2]

genemap_sub = subset(genemap, mtbc0_gene_id == geneA | gene_id == geneA | mtbc0_gene_id == geneB| gene_id == geneB)
genemap_sub$length = genemap_sub$end - genemap_sub$start
genemap_sub$mid = genemap_sub$start + (genemap_sub$length/2)

# --- Compute x-axis limits based on mutation positions ---
x_min <- min(hotspot_sub$position) - 100
x_max <- max(hotspot_sub$position) + 100

# --- Define y positions for layers ---
gene_ymin <- 0
gene_ymax <- 0.02   # thin rectangles
label_y   <- 0.01   # vertically centered labels
mut_y     <- 0.025   # triangles above genes

# --- Plot ---
ggplot() +
  
  geom_rect(
    data = genemap_sub,
    aes(xmin = start, xmax = end, ymin = gene_ymin, ymax = gene_ymax),
    color = "black", fill = "steelblue", alpha = 0.6
  ) +
  geom_text(
    data = genemap_sub,
    aes(x = mid, y = label_y, label = gene_id),
    size = 5, vjust = 0.5, color = "black"
  ) +
  geom_point(
    data = hotspot_sub,
    aes(x = as.integer(position), fill = births, y = mut_y),
    shape = 25, size = 6, alpha = 0.4, color = "black"
  ) +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  xlab("Genomic position") +
  coord_cartesian(xlim = c(x_min, x_max), ylim = c(0, 0.05), expand = FALSE) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.title       = element_blank(),
    axis.title.y       = element_blank(),
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    axis.line.y        = element_blank()
  )
  
```




## Genic vs- intergenic regions

```{r}
regions <- read.csv("./data/MTBC0/MTBC.genomic_regions.tsv", sep='\t')
head(regions)

tapply(regions$length, regions$region_type, summary)

ggplot(regions, aes(length,col=region_type)) + geom_density() +
  scale_x_log10()

```


# Target site preferences

```{r}




```


# The breakpoint drift problem
Detect and collapse breakpoint-drift uFS
For each region with >1 uFS, test all pairs for phylogenetic nesting.
Nested pairs (one clade contained within the other, zero co-occurrence) are merged into a single representative uFS before birth counting.

```{r}
library(ape)
library(dplyr)

# --- 1. Build strain × uFS presence-absence matrix from ancestral likelihoods tip rows ---
# anc_states_five rows 1:Ntips correspond to tree tips in tree$tip.label order
ntips <- length(tree$tip.label)
tip_rows <- anc_states_five[seq_len(ntips), ]          # tip states only
rownames(tip_rows) <- tree$tip.label

site_cols <- setdiff(names(tip_rows), "node")
ufs_mat   <- (as.matrix(tip_rows[, site_cols]) > 0.5) * 1L
# ufs_mat: rows = strains, cols = site_ids (uFS)

# --- 2. Map each uFS to its region ---
ufs_region <- setNames(hotspots$region, hotspots$site_id)

# --- 3. Test all within-region pairs for nesting ---
regions_multi <- names(which(table(hotspots$region) > 1))

nesting_results <- lapply(regions_multi, function(r) {
  
  sites <- hotspots$site_id[hotspots$region == r]
  # Keep only sites present in matrix
  sites <- sites[sites %in% colnames(ufs_mat)]
  if (length(sites) < 2) return(NULL)
  
  pairs <- combn(sites, 2, simplify = FALSE)
  
  res <- lapply(pairs, function(p) {
    test_phylo_nesting(p[1], p[2], ufs_mat, tree)
  })
  
  do.call(rbind, Filter(Negate(is.null), res))
})

nesting_df <- do.call(rbind, Filter(Negate(is.null), nesting_results))

# How many nested pairs?
cat("Total within-region pairs tested:", nrow(nesting_df), "\n")
cat("Nested pairs (breakpoint-drift candidates):", sum(nesting_df$nested), "\n")
cat("Of which with zero co-occurrence:", 
    sum(nesting_df$nested & nesting_df$cooccurring == 0), "\n")

# --- 4. Build merge map: collapse nested, non-co-occurring uFS ---
# Strategy: within each connected component of nested pairs, keep the uFS
# with the highest nstrains as the representative.

# Filter to high-confidence drift: nested AND mutually exclusive
drift_pairs <- subset(nesting_df, nested & cooccurring == 0)

# Build connected components via union-find
merge_map <- setNames(hotspots$site_id, hotspots$site_id)   # identity map

if (nrow(drift_pairs) > 0) {
  
  find <- function(x, parent) {
    while (parent[x] != x) x <- parent[x] <- parent[parent[x]]
    x
  }
  union <- function(a, b, parent) {
    ra <- find(a, parent); rb <- find(b, parent)
    if (ra != rb) parent[ra] <- rb
    parent
  }
  
  all_sites <- unique(c(drift_pairs$siteA, drift_pairs$siteB))
  parent    <- setNames(all_sites, all_sites)
  
  for (k in seq_len(nrow(drift_pairs))) {
    parent <- union(drift_pairs$siteA[k], drift_pairs$siteB[k], parent)
  }
  
  # For each component, elect the representative = highest nstrains
  components <- tapply(
    all_sites,
    sapply(all_sites, find, parent = parent),
    identity
  )
  
  for (comp in components) {
    rep_site <- comp[which.max(hotspots$nstrains[match(comp, hotspots$site_id)])]
    for (s in comp) merge_map[s] <- rep_site
  }
}

n_merged <- sum(merge_map != names(merge_map))
cat("uFS collapsed into representatives:", n_merged, "\n")
cat("Unique representatives elected:", length(unique(merge_map[merge_map != names(merge_map)])), "\n")

# --- 5. Apply merge map to hotspots ---
hotspots$site_id_original <- hotspots$site_id          # keep original for traceability
hotspots$site_id_merged   <- merge_map[hotspots$site_id]
hotspots$is_drift         <- hotspots$site_id_merged != hotspots$site_id

# Remove the non-representative members of each merged group
hotspots_corrected <- hotspots %>%
  filter(!is_drift) %>%
  select(-is_drift)

cat("hotspots rows before correction:", nrow(hotspots), "\n")
cat("hotspots rows after  correction:", nrow(hotspots_corrected), "\n")

# Overwrite hotspots for all downstream use
hotspots <- hotspots_corrected
```



# L2 vs La3 hotspots
```{r}

```





# Supplement

## Circos plot: overview of insertion sites
- add positions/density of DeJesus' essential genes
- add oriC location

```{r}
library(ape)
library(magrittr)
library(ggplot2)
library(dplyr)
library(Cairo)
library(circlize)
library(ComplexHeatmap)


custom_order = c('L1', 'L2', 'L3', 'L4', 'L5', 'L6', 'L7', 'L8', 'L9', 'L10','La1', 'La2', 'La3')

# Sort according to the phylogenetic tree!
refins_ordered <- subset(refins, lineage %in% custom_order) %>% mutate(column_name = factor(strain, levels = rev(tree$tip.label))) %>% arrange(column_name)
head(refins_ordered)

# Reduce number of strains: Get n per lineage
n = 500
strains_reduced = c()
for (lineage in rev(custom_order)){
  strains_lineage = unique(refins_ordered$strain[refins_ordered$lineage==lineage])
  if (length(strains_lineage) == 0){next}
  if (length(strains_lineage) <= n){strains_reduced = c(strains_reduced, strains_lineage)}
  else {
    strains_sample = sample(strains_lineage, n)
    strains_reduced = c(strains_reduced, strains_sample)
  }
}
length(strains_reduced)

# Create list of bed files, per strain, for circos
bed_list = list()
colori = c()

refins_reduced <- subset(refins_ordered, strain %in% strains_reduced)
head(refins_reduced)

strains_ordered <- unique(refins_reduced$strain)

for (i in 1:length(strains_ordered)){
  
  df <- subset(refins_reduced, strain==strains_ordered[i])
  bed <- data.frame(
    'chr' = df$chromosome,
    'start' = df$position,
    'end' = df$position+1,
    'lineage' = df$lineage
  )
  
  bed_list[[i]] <- bed
  
  # Now the lineage color
  lineage =df$lineage[1]
  col = mtb_col_scale[lineage]
  colori <- c(colori, col)
}

length(bed_list)
length(colori)

genome = data.frame(chr  = "MTBC0", start = 1,end   = 4435783) 

# Define color legend
lineages = names(mtb_col_scale)
lineages = lineages[lineages %in% unique(refins$lineage)]
colorrs = mtb_col_scale[lineages]



```

Plot the circle
```{r}
lgd = Legend(
  at = lineages, type = "lines", 
  legend_gp = gpar(col = colorrs, lwd = 2), 
  title_gp = gpar(fontfamily = "CMU Sans Serif", fontsize = 10),
  labels_gp = gpar(fontfamily = "CMU Sans Serif", fontsize = 10),
  title_position = "topleft", 
  title = "Lineage", break_dist = 2)

# Plot insertions!
options(fig.width = 15, fig.height = 15, repr.plot.width = 15, repr.plot.height = 15)
par(bg = "white")

CairoPDF("figures/main/fig4a.circos.pdf",  width = 10, height = 10)

circos.clear()

circos.par(start.degree=90)

circos.genomicInitialize(
  genome, 
  plotType = c("axis"),
  major.by = 250000,
  axis.labels.cex = 0.5,
  sector.names = NULL,
  labels.cex = 1
)

#circos.genomicDensity(bed_list, track.height = 0.05, col="grey", window.size = 2e4)

circos.genomicTrack(
  bed_list,
  stack=TRUE,
  track.height = 0.7,
  bg.border = NA,
  panel.fun = function(region, value, ...) {
    i = getI(...)
    circos.genomicRect(region, value, col = colori[i], border = colori[i], ...)
  }
)

draw(lgd, x = unit(10, "mm"), y = unit(20, "mm"), just = c("left", "bottom"))

dev.off()

```

## Distance effect
```{r}
library(ggplot2)

bin_sizes = c(50000, 100000, 200000, 500000)
chrom_length = 4435783

# Use positions from 5' matrix
matrix_metadata_five <- subset(matrix_metadata, side==5)
positions <- na.omit(matrix_metadata_five$ref_position)

dist_from_ori = c()

for (pos in positions){
  dist = dist_to_oriC(pos)
  dist_from_ori = c(dist_from_ori, dist)
}


bin_size = 100000
n_breaks = as.integer(chrom_length/bin_size, 0)
bins <- hist(dist_from_ori, breaks=n_breaks, plot = F)

bins_df <- data.frame(
  'mids' = bins$mids,
  'counts' = bins$counts,
  'density' = bins$density
)

cor.test(bins_df$mids, bins_df$counts, method = 'pearson')


# Plot histogram and density curve with two y axes
df <- data.frame(dists = dist_from_ori / 1000)

# parameters
bins <- 50
n <- nrow(df)
bw <- (max(df$dists) - min(df$dists)) / bins  # explicit bin width

fig4a <- ggplot(df, aes(x = dists)) +

  # histogram in counts (use explicit binwidth for precise scaling)
  geom_histogram(binwidth = bw, fill = "lightgray", color = NA) +

  # density scaled to count-scale using after_stat()
  geom_density(aes(y = after_stat(density * n * bw)),linewidth = 1) +

  # left axis = counts; right axis = true density (inverse transform)
  #scale_y_continuous(
  #  name = "Count",
  #  sec.axis = sec_axis(~ . / (n * bw), name = "Density")
  #) +
  ylab("Count") +

  labs(x = "Distance from oriC (kb)") +
  theme_bw(base_family = "CMU Sans Serif")

fig4a

```

# Not used

## Burden analysis 
```{r}
library(ggplot2)
library(magrittr)
library(ggrepel)
library(viridis)
library(cowplot)

burden_is6110 <- read.csv('workflow/results/gwas/burden_is6110.tsv', sep='\t')
burden_is6110 <- burden_is6110[,-c(8,9 )]
burden_is6110$typus <- rep('is6110', nrow(burden_is6110)) 
burden_is6110$index <- 1:nrow(burden_is6110)
head(burden_is6110)


load_permutation_res <- function(permutation_results){
  # Load permutation results
  permut <- data.frame()
  for (f in permutation_results){
    df <- read.csv(f, sep='\t')
    df <- df[,-c(8,9,10,11)]
    bn <- basename(f)
    rep <- strsplit(bn, '\\.')%>% unlist()
    n=rep[2]
    df$n <- rep(n, nrow(df))
    permut <- rbind(permut, df)
  }
  permut$logp <- -log10(permut$lrt.pvalue)
  return(permut)
}

permutations_is6110 <- load_permutation_res(list.files('workflow/results/gwas/permutations', pattern="is6110*", full.names = T))



# Get threshold from permutations: 0.1, 1 and 5 percentiles
threshoolds_is6110 <-  quantile(permutations_is6110$lrt.pvalue, probs=c(0.001, 0.01, 0.05))
burden_is6110$logp <- -log10(burden_is6110$lrt.pvalue)

# Add alternative gene names




# Manhatten plot
a_r3 <- ggplot(burden_is6110, aes(x=index, y=logp)) + 
  geom_point() + theme_bw(base_family = "CMU Sans Serif") + ylab('-Log10(P)') + xlab('') +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=8)) +
  geom_hline(aes(yintercept=-log10(threshoolds_is6110[[1]])), col='red', linetype='dashed') +
  geom_label_repel(data=subset(burden_is6110, logp>-log10(threshoolds_is6110[[1]])), aes(label=variant), label.size=0, size = 3)

      #geom_boxplot(data=permut, aes(x=variant, y=logp), outlier.shape = NA, size=0.1)+

# Volcano plot
b_r3 <- ggplot(burden_is6110, aes(y=logp, x=beta, col=100*variant_h2)) + geom_point(alpha=1) +
  geom_hline(aes(yintercept=-log10(threshoolds_is6110[[1]])), col='red', linetype='dashed') +
  theme_bw(base_family = "CMU Sans Serif") + ylab('-Log10(P)') + xlab('Effect size (beta)') +
  scale_color_viridis(option='turbo', name='Var. expl.') +
  theme(
    legend.position = c(1, 1),              # Bottom right inside the plot
    legend.justification = c("right", "top"),
    legend.background = element_rect(fill = NA, color = NA),
    legend.key.height = unit(0.5, "cm"),          # Smaller legend key
    legend.key.width = unit(0.5, "cm"),
    legend.text = element_text(size = 8),         # Smaller text
    legend.title = element_text(size = 8)
  ) +
    geom_label_repel(data=subset(burden_is6110, logp>-log10(threshoolds_is6110[[1]])), aes(label=variant), label.size=0, size = 3)

fig_gwas_r3 <- plot_grid(
  a_r3,b_r3, 
  labels = c('A', 'B'), 
  rel_widths = c(1.5,1),
  label_fontfamily='CMU Sans Serif')

fig_gwas_r3
```

## Test phylogenetic nesting of uFS
Nested uFS indicate single insertion events that where split into multiple because if mutations occurring in the flanking sequence after the insertion!

-> Identify breakpoint-drif candidates and merge them!

```{r}
test_phylo_nesting <- function(siteA, siteB, ins_mat, tree) {
  
  SA <- rownames(ins_mat)[ins_mat[, siteA] == 1]
  SB <- rownames(ins_mat)[ins_mat[, siteB] == 1]
  
  # skip trivial cases
  if (length(SA) < 2 || length(SB) < 2)
    return(NULL)
  
  # co-occurrence
  cooccur <- length(intersect(SA, SB))
  
  # MRCA nodes
  mrcaA <- getMRCA(tree, SA)
  mrcaB <- getMRCA(tree, SB)
  
  if (is.null(mrcaA) || is.null(mrcaB))
    return(NULL)
  
  # tips descending from each MRCA
  cladeA <- get_descendant_tips(tree, mrcaA)
  cladeB <- get_descendant_tips(tree, mrcaB)
  
  # nesting tests
  B_within_A <- all(SB %in% cladeA)
  A_within_B <- all(SA %in% cladeB)
  
  nested <- (B_within_A || A_within_B)
  
  data.frame(
    siteA = siteA,
    siteB = siteB,
    nA = length(SA),
    nB = length(SB),
    cooccurring = cooccur,
    nested = nested,
    B_within_A = B_within_A,
    A_within_B = A_within_B
  )
}
```

## Adaptive parallelism index
```{r}
library(ggplot2)
library(ggrepel)
library(viridis)
library(magrittr)

load(paths$rdata)

# Plot: nr of independent insertions vs. median size of clusters with a specific insertion!
head(hotspots)

regions = unique(hotspots$region)

hotspots_summary <- data.frame(
  region = regions,
  nsites = rep(0, length(regions)),
  nbirths = rep(0, length(regions)),
  ndeaths = rep(0, length(regions)),
  strains_min = rep(0, length(regions)),
  strains_max = rep(0, length(regions)),
  strains_mean = rep(0, length(regions)), 
  strains_total = rep(0, length(regions))
)



for (i in 1:length(regions)){
  r = regions[i]
  hotspots_sub = subset(hotspots, region == r)
  hotspots_summary$nsites[i] = nrow(hotspots_sub)
  hotspots_summary$nbirths[i] = sum(hotspots_sub$births)
  hotspots_summary$ndeaths[i] = sum(hotspots_sub$deaths)
  hotspots_summary$strains_min[i] = min(hotspots_sub$nstrains)
  hotspots_summary$strains_max[i] = max(hotspots_sub$nstrains)
  hotspots_summary$strains_mean[i] = mean(hotspots_sub$nstrains)
  hotspots_summary$strains_total[i] = sum(hotspots_sub$nstrains)
}


# Adaptive parallelism index API
hotspots_summary$API = hotspots_summary$nsites * log10(hotspots_summary$strains_mean + 1)
ggplot(hotspots_summary, aes(API)) + geom_density()

```


## Compare the 2 hotspot summaries
Mostly as expected, when summarized by region, fewer birth events are inferred. Few exceptions, especially two intergenic regions with way more births:
  - mtbc0_003539;mtbc0_003540
  - mtbc0_000845;mtbc0_000846

```{r}
library(ggrepel)

head(hotspots_summary)
head(hotspot_summary2)

hotspot_summary_merged = merge(hotspots_summary, hotspot_summary2, by='region')

head(hotspot_summary_merged)


ggplot(hotspot_summary_merged, aes(nbirths.x, nbirths.y)) + 
  geom_point() +
  geom_abline(intercept = 0, slope=1) +
  geom_label_repel(
    data=subset(
      hotspot_summary_merged, 
      nbirths.x>100 | nbirths.y>100), 
    aes(label=region), 
    label.size=0, 
    size = 3)

```

