---
title: "A niche constraints model for the evolutionary dynamics of IS6110"
output:
  html_document: 
    theme: simplex
date: "2025-03-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```


# Snakemake I/O
```{r}

if (exists("snakemake")) {
  paths <- list(
      subtree = snakemake@input$subtree,
      abc_params = snakemake@input$abc_params,
      abc_summary = snakemake@input$abc_summary,
      rdata = snakemake@input$rdata,
      output_html     = snakemake@output[[1]]
    )
} else {
  message("No snakemake object found — using default paths for interactive testing")
  paths <- list(
      subtree = "workflow/results/niche_model/subsampled_tree.rooted.nex",
      abc_params = "workflow/results/niche_model/1e6_v2/abc_params.tsv",
      abc_summary = "workflow/results/niche_model/1e6_v2/abc_summaries.tsv",
      obs_stats = "workflow/results/niche_model/obs_stats.yaml",
      rdata = "workflow/results/is6110.RData",
      output_html     = NULL
    )
}

library(abc)
library(ape)
library(castor)
library(cowplot)
library(dplyr)
library(GGally)
library(ggplot2)
library(ggtree)
library(magrittr)
library(ranger)
library(reshape2)
library(tidyr)
library(viridis)

```


# Summary statistics

## Load observed summary statistics in the subtree
```{r}
library(yaml)

obs_stats <- read_yaml(paths$obs_stats)
obs_df <- as.data.frame(obs_stats)
```


## Load simulation results
```{r}
library(magrittr)
library(ggplot2)
library(reshape2)

params_mat <- read.csv(paths$abc_params, sep='\t') %>% as.matrix()
#colnames(params_mat) <- c('p_lethal', 'p_deleterious', 'r_birth', 'r_death', 'r_death_mult')
dim(params_mat)
summary(params_mat)

summaries_mat <- read.csv(paths$abc_summary, sep='\t') %>% as.matrix()
dim(summaries_mat)
colnames(summaries_mat)
summary(summaries_mat)

# Plot distributions of all summary stats
as.data.frame(summaries_mat) %>% melt() %>%
  ggplot(aes(value)) + geom_density() + facet_wrap(~variable, scales='free')

```


# Find informative summary stats
The full set of summary stats is redundant, as shown by the correlogram. Use random forest to pick stats that are actuall informative. 

```{r}
library(ggplot2)
library(GGally)

# Plot correlogram
panelA <- ggcorr(summaries_mat, label=TRUE)
panelA
```

Estimate variable importance with random forests
```{r}
library(ranger)
library(ggplot2)

params <- c('p_neutral', 'gamma_shape', 'gamma_scale', 'r_birth', 'r_loss')

dim(summaries_mat)
subsamp_n = 10000
subsamp_i = sample(1:nrow(summaries_mat), subsamp_n)

importanzia <- data.frame()

for (p in params){

  reference_table <- cbind(
    p = params_mat[subsamp_i,p],
    summaries_mat[subsamp_i,]
  )

  rf <- ranger(
    p ~ .,
    data = reference_table,
    importance = "permutation",
    num.threads=6
  )
  
  imp = importance(rf)
  
  df <- data.frame(
    var = rep(p, ncol(summaries_mat)),
    stat = names(imp),
    value = unname(imp)
  )
  importanzia <- rbind(importanzia, df)
}

importanzia %>%
    group_by(stat) %>%
    summarise(
      importance_sum = sum(value),
      importance_mean = mean(value),
      importance_sd = sd(value),
      importance_min = min(value),
      importance_max = max(value),
      .groups = "drop"
    ) %>%
    arrange(desc(importance_sum))
  

# Plot variable importance
panelB <- ggplot(importanzia, aes(y=stat, x=value)) + 
  geom_point() + facet_wrap(~var, scales='free_x') +
  theme_bw(base_family='CMU Sans Serif', base_size=12) + 
  xlab("Variable importance") + ylab("")

panelB
```

Save correlogram and variable importance for the supplement.
```{r}

ggsave('figures/supplement/figS12.summary_stats_correlogram.png', panelA, width = 45, height =45, units = "cm", dpi = 300) 

ggsave('figures/supplement/figS13.summary_stats_variable_importance.png', panelB,  width = 20, height = 20, units = "cm", dpi = 300) 
```


## Select informative core stats
```{r}
core_stats <- c(
  'sfs_1',              
  'sfs_5',
  'sfs_gt25pct',
  'gini_occupancy',   
  'max_occupancy',
  'median_cn',
  'lineage_sd'
)

obs_stats_core <- obs_stats[core_stats]
summaries_mat_core <- summaries_mat[,core_stats]

```

# ABC
```{r}
library(abc)


# Run abc (rejection followed by local linear regression)
abc_res <- abc(target = obs_stats_core,
               param = params_mat,
               sumstat = summaries_mat_core,
               tol = 0.001,   # keep top 1% closest sims
               method = "neuralnet")

summary(abc_res)

```



# Posterior vs. prior distributions
```{r}
library(tidyr)
library(ggplot2)

posterior <- as.data.frame(abc_res$unadj.values)
#colnames(posterior) <- c('p_neutral', 'gamma_shape', 'gamma_scale', 'r_birth', 'r_loss')

# Add prior samples for comparison
prior <- as.data.frame(params_mat)
#colnames(prior) <- c('p_essential', 'p_tolerated', 'r_birth', 'r_purge')

posterior$distribution <- 'posterior'
prior$distribution     <- 'prior'

panelA <- 
bind_rows(posterior, prior) %>%
  pivot_longer(-distribution, names_to='parameter', values_to='value') %>%
  ggplot(aes(value, fill=distribution, colour=distribution)) +
  geom_density(alpha=0.4) +
  facet_wrap(~parameter, scales='free') +
  scale_fill_manual(values=c(posterior='steelblue', prior='grey60')) +
  scale_colour_manual(values=c(posterior='steelblue', prior='grey40')) +
  theme_bw(base_family='CMU Sans Serif', base_size=12) +
  xlab('') + ylab('Density')

panelA

```


# Posterior predicive check
```{r}
# ---------------------------------------------------------------------------
# Panel D: posterior predictive check
# ---------------------------------------------------------------------------

# OR, more simply, if you want to be explicit about the tolerance:
# Extract which simulations were accepted
accepted_idx <- abc_res$region  # logical vector of length nrow(summaries_mat)
accepted_sums <- summaries_mat[accepted_idx, ]

#used_stats <- names(obs_stats)
used_stats <- core_stats

ppc_long <- as.data.frame(accepted_sums[, used_stats]) %>%
  pivot_longer(everything(), names_to = 'statistic', values_to = 'value') 
  
obs_ppc <- data.frame(
  statistic = factor(used_stats, levels = used_stats),
  xintercept = as.numeric(obs_stats[used_stats])
)

pD <- ggplot(ppc_long, aes(value)) +
  geom_density(fill = 'steelblue', alpha = 0.5, colour = NA) +
  geom_vline(data = obs_ppc, aes(xintercept = xintercept),
             colour = 'red', linewidth = 0.7) +
  facet_wrap(~ statistic, scales = 'free', ncol = 3) +
  theme_bw(base_family = 'CMU Sans Serif', base_size = 10) +
  theme(strip.background = element_blank()) +
  xlab('') + ylab('Density')

pD
```


# Best model vs. observed data
```{r}
# Best model
best_i = which(abc_res$dist==min(abc_res$dist))
params_mat[best_i,]
comp_df <- cbind(obs_stats, summaries_mat[best_i,])
colnames(comp_df) <- c('Observed', 'Simulated')
comp_df



```




# Simulate data with best parameters
```{r}


args_best <- c(
  "workflow/scripts/niche_model_numba.py",
  "--tree", "workflow/results/niche_model/subsampled_tree.rooted.nex",
  "--p_essential", params_mat[best_i,'p_essential'],
  "--p_tolerated", params_mat[best_i,'p_tolerated'],
  "--r_birth", params_mat[best_i, 'r_birth'],
  "--r_purge", params_mat[best_i, 'r_purge'],
  "--metadata", paths$metadata,
  "--niches 7362",
  "--seed", "42"
)

stdout_best = system2("/home/cristobal/anaconda3/envs/ete3/bin/python", args_best, stdout=TRUE)
cn_sim_best = read.delim(text=stdout_best, sep='\t')
summary(cn_sim_best)


# Use weighted medians of the posteriors
abc_sum <- summary(abc_res)
str(abc_sum)
abc_sum

args_wm <- c(
  "workflow/scripts/niche_model_numba.py",
  "--tree", "workflow/results/niche_model/subsampled_tree.rooted.nex",
  "--p_essential", abc_sum['Weighted Median:','p_essential'],
  "--p_tolerated", abc_sum['Weighted Median:','p_tolerated'],
  "--r_birth", abc_sum['Weighted Median:','r_birth'],
  "--r_purge", abc_sum['Weighted Median:','r_purge'],
  "--metadata", paths$metadata,
  "--niches 7362",
  "--seed", "10"
)

stdout_wm = system2("/home/cristobal/anaconda3/envs/ete3/bin/python", args_wm, stdout=TRUE)
cn_sim_wm = read.delim(text=stdout_wm, sep='\t')
summary(cn_sim_wm$copy_number)
```


## Copy number distribution
```{r}
# Compare observed vs simulated CN distribution
obs_cn <- copy_numbers[[2]]
sim_cn <- cn_sim_best$copy_number

data.frame(
  quantile = c('25%', '50%', '75%'),
  observed = quantile(obs_cn, c(0.25, 0.5, 0.75)),
  simulated = quantile(sim_cn, c(0.25, 0.5, 0.75))
)

# Plot distributions
bind_rows(
  data.frame(CN = obs_cn, source = 'observed'),
  data.frame(CN = sim_cn, source = 'simulated')
) %>%
  ggplot(aes(CN, fill = source)) +
  geom_density(alpha = 0.4) +
  theme_bw()
```

## Figure
A) Observed and simulated copy numbers with parameter values sampled from the posterior. B) Posterior distributions for birth rate, death rate, and the proportion of suitable niches in the genome.

```{r}
library(ggtree)
library(ggplot2)
library(ape)
library(viridis)


tree <- read.tree('workflow/results/niche_model/subsampled_tree.rooted.nex')

p <- ggtree(tree, alpha=0.1, size=0) 
# Add observed values to tree
p$data$copy_number <- copy_numbers$CN[match(p$data$label, copy_numbers$strain)]
# Add simulated values to tree
p$data$copy_number_simulated <- cn_sim_best$copy_number[match(p$data$label, cn_sim_best$strain)]

# Plot trees
## Observed
a <- p + geom_tree(aes(col=copy_number)) + scale_color_viridis(option = 'turbo', guide="none") 

## Simulated
b <- p + geom_tree(aes(col=copy_number_simulated)) + scale_color_viridis(option = 'turbo', guide="none")

## Combine
lower <- cowplot::plot_grid(a,b, labels=c('B', 'C'))
cowplot::plot_grid(panelA, lower, nrow=2, labels=c('A', ''))

#ggsave('figures/main/fig6.nichemodel.png', width = 22, height = 20, units = "cm", dpi = 300) 


```


# Cross-validation
```{r}
# Basic summary of posteriors
summary(abc_res)

# Goodness of fit: cross-validation
cv_res <- cv4abc(
  param    = params_mat,
  sumstat  = summaries_mat,
  nval     = 100,
  tols     = c(0.01, 0.05),
  method   = "neuralnet"
)
summary(cv_res)
plot(cv_res)  # predicted vs true for each parameter
```



# Adjusted vs. unadjusted values
```{r}
library(abc)
library(ggplot2)
library(cowplot)
library(dplyr)
library(tidyr)

# Extract unadjusted and adjusted parameters
unadj <- as.data.frame(abc_res$unadj.values)
adj   <- as.data.frame(abc_res$adj.values)

colnames(unadj) <- c('p_lethal', 'r_birth', 'r_death')
colnames(adj)   <- c('p_lethal', 'r_birth', 'r_death')

# Add iteration index for pairing
unadj$iter <- 1:nrow(unadj)
adj$iter   <- 1:nrow(adj)

# ---------------------------------------------------------------------------
# Panel A: Unadjusted vs adjusted parameters (scatter + arrows)
# ---------------------------------------------------------------------------

# For visualization, focus on two key parameters
param_pairs <- list(
  list(x = 'p_lethal', y = 'r_birth'),
  list(x = 'p_lethal', y = 'r_death'),
  list(x = 'r_birth',  y = 'r_death')
)

plot_pair <- function(xvar, yvar) {
  df <- data.frame(
    iter = unadj$iter,
    x_unadj = unadj[[xvar]],
    y_unadj = unadj[[yvar]],
    x_adj   = adj[[xvar]],
    y_adj   = adj[[yvar]]
  )
  
  ggplot(df, aes(x = x_unadj, y = y_unadj)) +
    geom_point(size = 0.8, colour = 'grey60', alpha = 0.6) +
    geom_segment(
      aes(xend = x_adj, yend = y_adj),
      arrow = arrow(length = unit(1.5, 'mm'), type = 'closed'),
      colour = 'steelblue', alpha = 0.3, linewidth = 0.3
    ) +
    geom_point(aes(x = x_adj, y = y_adj), size = 0.8, colour = 'steelblue', alpha = 0.8) +
    theme_bw(base_family = 'CMU Sans Serif', base_size = 11) +
    labs(
      x = xvar,
      y = yvar,
      title = paste(xvar, 'vs', yvar)
    ) +
    theme(plot.title = element_text(size = 10, face = 'bold'))
}

pA1 <- plot_pair('p_lethal', 'r_birth')
pA2 <- plot_pair('p_lethal', 'r_death')
pA3 <- plot_pair('r_birth', 'r_death')

pA <- plot_grid(pA1, pA2, pA3, ncol = 3, labels = c('A1', 'A2', 'A3'),
                label_fontfamily = 'CMU Sans Serif')

# ---------------------------------------------------------------------------
# Panel B: Marginal adjustment per parameter (density + shift)
# ---------------------------------------------------------------------------

unadj_long <- unadj %>%
  select(-iter) %>%
  pivot_longer(everything(), names_to = 'parameter', values_to = 'value') %>%
  mutate(adjustment = 'Unadjusted (accepted)')

adj_long <- adj %>%
  select(-iter) %>%
  pivot_longer(everything(), names_to = 'parameter', values_to = 'value') %>%
  mutate(adjustment = 'Ridge regression\nadjusted')

bind_rows(unadj_long, adj_long) -> dens_df

pB <- ggplot(dens_df, aes(value, fill = adjustment, colour = adjustment)) +
  geom_density(alpha = 0.4, linewidth = 0.5) +
  facet_wrap(~ parameter, scales = 'free', ncol = 1) +
  scale_fill_manual(values = c(
    'Unadjusted (accepted)'    = 'grey60',
    'Ridge regression\nadjusted' = 'steelblue'
  )) +
  scale_colour_manual(values = c(
    'Unadjusted (accepted)'    = 'grey40',
    'Ridge regression\nadjusted' = 'steelblue4'
  )) +
  theme_bw(base_family = 'CMU Sans Serif', base_size = 11) +
  theme(
    legend.position = c(0.98, 0.98),
    legend.justification = c('right', 'top'),
    legend.title = element_blank(),
    legend.key.size = unit(0.4, 'cm'),
    strip.background = element_blank()
  ) +
  xlab('') + ylab('Density')

pB
# ---------------------------------------------------------------------------
# Panel C: Adjustment magnitude per parameter (violin plot)
# ---------------------------------------------------------------------------

# Compute relative shift for each parameter
adjustment_df <- data.frame(
  p_lethal = (adj$p_lethal - unadj$p_lethal) / (max(unadj$p_lethal) - min(unadj$p_lethal)),
  r_birth  = (adj$r_birth - unadj$r_birth) / (max(unadj$r_birth) - min(unadj$r_birth)),
  r_death  = (adj$r_death - unadj$r_death) / (max(unadj$r_death) - min(unadj$r_death))
) %>%
  mutate(iter = 1:nrow(.)) %>%
  pivot_longer(-iter, names_to = 'parameter', values_to = 'shift')

pC <- ggplot(adjustment_df, aes(parameter, shift, fill = parameter)) +
  geom_violin(alpha = 0.6, colour = NA) +
  geom_hline(yintercept = 0, linetype = 'dashed', colour = 'grey50', linewidth = 0.5) +
  theme_bw(base_family = 'CMU Sans Serif', base_size = 11) +
  theme(legend.position = 'none', strip.background = element_blank()) +
  xlab('') + ylab('Relative shift (standardised)') +
  scale_fill_manual(values = c(p_lethal = '#1f77b4', r_birth = '#ff7f0e', r_death = '#2ca02c'))


```




# Narrow down priors
```{r}
library(dplyr)

posterior_fresh <- as.data.frame(abc_res$adj.values)
colnames(posterior_fresh) <- c('p_lethal', 'r_birth', 'r_death')

posterior_summary <- posterior_fresh %>%
  pivot_longer(everything(), names_to = 'parameter', values_to = 'value') %>%
  group_by(parameter) %>%
  summarise(
    q025 = quantile(value, 0.025),
    q975 = quantile(value, 0.975),
    median = median(value),
    .groups = 'drop'
  )

print(posterior_summary)
```


# Not used
## Use region-wise summary stats instead of full matrix?
```{r}
library(ape)
library(castor)
library(magrittr)

# Load phylogeny
tree <- read.tree(paths$subtree)

# Five prime presence-absence matrix
matrix5 <- read.csv(args[2], sep='\t')
rownames(matrix5) <- matrix5[,1]
matrix5 <- matrix5[,-1]
colnames(matrix5) <- gsub('X', '', colnames(matrix5))

# Matrix metadata
matrix_metadata <- read.csv(args[3], sep='\t')

# Output path
outpath <- args[4]

# Disrupted regions
regions = unique(matrix_metadata$context) %>% na.omit()

# Reorder matrix rows so they match the tree!
phyl_order <- match(tree$tip.label, rownames(matrix5))
matrix5 <- matrix5[phyl_order,]
all(rownames(matrix5) == tree$tip.label)

hotspot_summary = data.frame()

for (region in regions){

  ufs = subset(matrix_metadata, side==5 & context == region)$site_id
  if (length(ufs)==0){next}

  regionmat = matrix5[,ufs]

  if (length(ufs) == 1){regionmat_vec = regionmat}
  else {regionmat_vec = as.integer(rowSums(regionmat) > 0)}
  nstrains = sum(regionmat_vec)

  # Convert to numeric starting with 1
  regionmat_vec_mapped <- map_to_state_space(regionmat_vec)

  # Reconstruct ancestral states
  asr_result <- asr_max_parsimony(tree, tip_states = regionmat_vec_mapped$mapped_states)

  # Which column in ancestral_likelihoods corresponds to presence (1)?
  presence_state_index <- regionmat_vec_mapped$name2index[["1"]]
  # For each node, store likelihood of presence
  lh_presence <- asr_result$ancestral_likelihoods[, presence_state_index]

  # Combine tip and internal node states
  # Use mapped tip states (1=absent, 2=present) converted to 0/1
  tip_states_01 <- ifelse(regionmat_vec_mapped$mapped_states == 1, 0, 1)
  full_states <- c(tip_states_01, lh_presence)
  full_states_binary <- ifelse(full_states > 0.5, 1, 0)

  # Count number of independent birth/insertion events
  nbirths = 0
  ndeaths = 0

  for (j in 1:nrow(tree$edge)) {
        
    parent = tree$edge[j, 1]
    child  = tree$edge[j, 2]

    parent_state = full_states_binary[parent]
    child_state = full_states_binary[child]

    # No change
    if (parent_state == child_state) next
    # Birth
    if (parent_state==0 & child_state==1){nbirths = nbirths+1}
    # Death
    else if (parent_state==1 & child_state==0){ndeaths = ndeaths+1}
  }

  row = data.frame(
    'region' = region,
    'nstrains' = nstrains,
    'nbirths' = nbirths,
    'ndeaths' = ndeaths
  )
  #print(row)
  hotspot_summary = rbind(hot

```

## Internal vs. terminal branch lengths
```{r}
library(ape)
library(ggtree)
library(ggplot2)

tree = read.tree(paths$subtree)

tree$edge.length[1:tree$Nnode]

branch_lengths <- data.frame(
  length = treesim$edge.length,
  type = c(
    rep('terminal', treesim$Nnode),
    rep('internal',length(treesim$edge.length)-treesim$Nnode)
  )
)
head(branch_lengths)

ggplot(branch_lengths, aes(x=type, y=length)) + geom_boxplot()
ggplot(branch_lengths, aes(length, col=type)) + geom_density()

branch_lengths %>%
  group_by(type) %>%
  summarize(
    mean = mean(length),
    median = median(length)
    )

```

## Prior diagnostics
Check whether the priors used actually produce meaningful results. I had issues with "pathological regimes", where a large proportion of simulations produced maximal or 0 copy numbers...
```{r}

# Calculate Euclidean distance
obs <- unlist(obs_stats_core)

d <- sqrt(rowSums((sweep(as.matrix(summaries_mat_core), 2, obs))^2))
quantile(d, c(.001,.01,.05,.1))
keep <- order(d)[1:20]

cbind(d[keep], summaries_mat_core[keep, ])

# Steep jump in distances between 0 and max. CN
plot(sort(d))

# Plot parameters against core summary stats
subsamp_n = 100000
subsamp_i = sample(1:nrow(summaries_mat), subsamp_n)
simdf <- cbind(params_mat[subsamp_i,], summaries_mat[subsamp_i,]) %>% as.data.frame()

# r_birth
ggplot(simdf, aes(r_birth, mean_cn)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(r_birth, gini_occupancy)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(r_birth, n_rare)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(r_birth, lineage_sd)) + geom_point(alpha=0.1) + geom_smooth()

# r_loss
ggplot(simdf, aes(r_loss, mean_cn)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(r_loss, gini_occupancy)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(r_loss, n_rare)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(r_loss, lineage_sd)) + geom_point(alpha=0.1) + geom_smooth()

# p_neutral
ggplot(simdf, aes(p_neutral, mean_cn)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(p_neutral, gini_occupancy)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(p_neutral, n_rare)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(p_neutral, lineage_sd)) + geom_point(alpha=0.1) + geom_smooth()

# gamma_shape
ggplot(simdf, aes(gamma_shape, mean_cn)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(gamma_shape, gini_occupancy)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(gamma_shape, n_rare)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(gamma_shape, lineage_sd)) + geom_point(alpha=0.1) + geom_smooth()

# gamma_scale
ggplot(simdf, aes(gamma_scale, mean_cn)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(gamma_scale, gini_occupancy)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(gamma_scale, n_rare)) + geom_point(alpha=0.1) + geom_smooth()
ggplot(simdf, aes(gamma_scale, lineage_sd)) + geom_point(alpha=0.1) + geom_smooth()


# How many simulations are pathological?
regime <- ifelse(
    summaries_mat[,"mean_cn"] == 0,
    "extinct",
    ifelse(summaries_mat[,"mean_cn"] > 50,
           "exploding",
           "reasonable")
)
table(regime)


boxplot(params_mat[,"p_neutral"] ~ regime)
boxplot(params_mat[,"r_birth"] ~ regime)
boxplot(params_mat[,"r_loss"] ~ regime)
boxplot(params_mat[,"gamma_shape"] ~ regime)
boxplot(params_mat[,"gamma_scale"] ~ regime)


# Tighten posteriors
patholog_i = which(summaries_mat[,"mean_cn"]==0 | summaries_mat[,"mean_cn"] > 100)
print(paste(length(patholog_i), "observations out of", nrow(summaries_mat), "produce unrealistic results"))

params_mat_tight <- params_mat[-patholog_i, ]
summaries_mat_tight <- summaries_mat[-patholog_i,]
summaries_mat_tight_core <- summaries_mat_tight[,core_stats]


# Multiple criteria
summary(as.data.frame(summaries_mat))

realistic_ <- with(
  as.data.frame(summaries_mat),
  mean_cn > 0 & mean_cn < 30 &
  max_cn < 50
  )
table(realistic_)

summary(params_mat[realistic_, ])
obs_stats
summary(summaries_mat[realistic_,])

```