---
title: "A niche constraints model for the evolutionary dynamics of IS6110"
output:
  html_document: 
    theme: simplex
date: "2025-03-19"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```



# Snakemake I/O
```{r}
paths <- tryCatch(
  {
    list(
      branchsummary_5 = snakemake@input[[1]],
      output_html     = snakemake@output[[1]]
    )
  },
  error = function(e){
    list(
      branchsummary_5 = 'workflow/results/detettore/AS_5prime.branch_summary.tsv',
      output_html     = NULL  # or whatever default makes sense
    )
  }
)

```

# Constrained niche model
A simple model to explain IS copy number variation, hotspots, symmetry etc.:
Few loci in the genome where IS can insert safely. This are hit by chance, and once hit,
additional copies increase the birth rate. Model implemented in contrained_niche_evolution.py.

Use downsampled tree, e.g. 500 strains.

Fit model to data using approximate Bayesian computation. Summary statistics: median CN, var(CN)_clades, ...
Chose initial distributions according to real data (birth rate, death rate, max_number ...). 
Fixed: initial_copies = 1, max_niches = 38
Fitted: p_lethal, r_birth, r_death

p_lethal = 0.95
r_transpose = 4000
r_delete = 0
max_niches = 40
initial_copies = 1



## Approximate Bayesian computation

Simulation test run
```{r}
args <- c(
  "workflow/scripts/niche_model.py",
  "--tree", "workflow/results/niche_model/subsampled_tree.rooted.nex",
  "--p_lethal",  "0.6621324",
  "--r_birth", "4507.4047070",
  "--r_death", "249.1706106",
  "--initial_copies", "1",
  "--max_copies", "40"
)


stdout = system2("/home/cristobal/anaconda3/envs/ete3/bin/python", args, stdout=TRUE)
cn_sim = read.delim(text=stdout, sep='\t')
```


Test run with the spatially explicit model
```{r}
args <- c(
  "workflow/scripts/niche_model.py",
  "--tree", "workflow/results/niche_model/subsampled_tree.rooted.nex",
  "--p_lethal",  "0.6621324",
  "--r_birth", "4507.4047070",
  "--r_death", "249.1706106",
  "--initial_copies", "1",
  "--max_copies", "40",
  "--spatial",
  "--niches", "10000"
)


stdout = system2("/home/cristobal/anaconda3/envs/ete3/bin/python", args, stdout=TRUE)
cn_sim = read.delim(text=stdout, sep='\t')
```


Summary statistics
```{r}

library(ape)
library(ggplot2)
library(ggtree)
library(viridis)
library(dplyr)

load('is6110.RData')

get_summary_stats <- function(copy_numbers, metadata){
  
  copy_numbers$lineage = metadata$lineage[match(copy_numbers[,1], metadata$GNUMBER)]
  
  cladevar = copy_numbers %>% 
    group_by(lineage) %>%
    summarize(Mean = mean(CN), SD = sd(.[[2]]))
  
  summdata = list(
    'Mean' = mean(copy_numbers[,2]),
    'SD' = sd(copy_numbers[,2]),
    'min' = min(copy_numbers[,2]),
    '25%' = quantile(copy_numbers[,2], 0.25),
    '50%' = quantile(copy_numbers[,2], 0.5),
    '75%' = quantile(copy_numbers[,2], 0.75),
    'max' = max(copy_numbers[,2]),
    'cladevar' = var(cladevar$Mean),
    'cladesd' = sd(cladevar$Mean)
  ) 
  return(summdata)
}


treesim = read.tree('workflow/results/niche_model/subsampled_tree.rooted.nex')
plot(treesim, show.tip.label = FALSE)
copy_numbers_sub <- subset(copy_numbers, strain %in% tree$tip.label)
head(copy_numbers_sub)
stats_obs = get_summary_stats(copy_numbers_sub,metadata)
stats_obs

```


## Get posterior distributions
```{r}
library(abc)
library(magrittr)


params_mat <- read.csv('workflow/results/niche_model/abc_params.tsv', sep='\t') %>% as.matrix()
colnames(params_mat) <- c('p_lethal', 'r_birth', 'r_death')
dim(params_mat)
summary(params_mat)



summaries_mat <- read.csv('workflow/results/niche_model/abc_summaries.tsv', sep='\t') %>% as.matrix()
colnames(summaries_mat) <- c('Mean', 'SD', 'min', '25%', '50%',  '75%', 'max', 'cladevar', 'cladesd')
dim(summaries_mat)

# Run abc (rejection followed by local linear regression)
abc_res <- abc(target = stats_obs,
               param = params_mat,
               sumstat = summaries_mat,
               tol = 0.001,   # keep top 1% closest sims
               method = "neuralnet")

summary(abc_res)
abc_res
plot(abc_res, params_mat)
hist(abc_res)

plot(density(abc_res$dist))

```

Plot parameter combinations vs distance to visualize peaks
```{r}
library(plotly)

str(abc_res)

abc_res_best = abc_res$region

abc_dists <- data.frame(
  p_lethal =  params_mat[,'p_lethal'][abc_res_best],
  r_birth = params_mat[,'r_birth'][abc_res_best],
  r_death = params_mat[,'r_death'][abc_res_best],
  p_lethal_adj = abc_res$adj.values[,'p_lethal'],
  r_birth_adj = abc_res$adj.values[,'r_birth'],
  r_death_adj = abc_res$adj.values[,'r_death'],
  dist = abc_res$dist[abc_res_best]
)

head(abc_dists)

# r_birth vs p_lethal
ggplot(abc_dists, aes(r_birth, p_lethal, col=dist)) + geom_point(size=2, alpha=0.5) +
  scale_color_viridis(option = 'turbo')

# with adjusted values
ggplot(abc_dists, aes(r_birth_adj, p_lethal_adj, col=dist)) + geom_point(size=3, alpha=0.5) +
  scale_color_viridis(option = 'turbo') + theme_bw() + xlab('Birth rate') + ylab('Proportion of inaccessible sites')

# How do adjusted and unadjusted values compare?
ggplot(abc_dists, aes(r_birth_adj, r_birth, col=dist)) + geom_point(size=2, alpha=0.5) +
  scale_color_viridis(option = 'turbo')

ggplot(abc_dists, aes(p_lethal_adj, p_lethal, col=dist)) + geom_point(size=2, alpha=0.5) +
  scale_color_viridis(option = 'turbo')


# 3d plot
plot_ly(
  abc_dists,
  x = ~r_birth, y = ~p_lethal, z = ~-dist,
  intensity = ~-dist,
  type = "mesh3d",
  colors= colorRamp(rainbow(5))
)


```

Cross validation
```{r}
cv = cv4abc(params_mat,summaries_mat, abc_res, 100, c(0.01, 0.001), 'mean')

```




Get best model and re-run to get copy numbers
```{r}
best_i = which(abc_res$dist==min(abc_res$dist))
params_mat[best_i,]
summaries_mat[best_i,]
stats_obs


args <- c(
  "workflow/scripts/niche_model.py",
  "--tree", "workflow/results/niche_model/subsampled_tree.rooted.nex",
  "--p_lethal",  "0.6621324",
  "--r_birth", "4507.4047070",
  "--r_death", "249.1706106",
  "--initial_copies", "1",
  "--max_copies", "40"
)


stdout = system2("/home/cristobal/anaconda3/envs/ete3/bin/python", args, stdout=TRUE)
cn_sim = read.delim(text=stdout, sep='\t')
head(cn_sim)
```


## Figure

A) Observed and simulated copy numbers with parameter values sampled from the posterior. B) Posterior distributions for birth rate, death rate, and the proportion of suitable niches in the genome.

```{r}
library(ggtree)
library(ggplot2)
library(ape)
library(viridis)


tree <- read.tree('workflow/results/niche_model/subsampled_tree.rooted.nex')

p <- ggtree(tree, alpha=0.1, size=0) 
# Add observed values to tree
p$data$copy_number <- copy_numbers$CN[match(p$data$label, copy_numbers$strain)]
# Add simulated values to tree
p$data$copy_number_simulated <- cn_sim$copies[match(p$data$label, cn_sim$tip)]

# Plot trees
## Observed
a <- p + geom_tree(aes(col=copy_number)) + scale_color_viridis(option = 'turbo', guide="none") + 
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5, linetype='dashed'))

## Simulated
b <- p + geom_tree(aes(col=copy_number_simulated)) + scale_color_viridis(option = 'turbo', guide="none")

## Combine
cowplot::plot_grid(a,b)

# Plot posterior distributions



```


## Posterior predictive checks

```{r}
post_params <- post$adj.values    # or post$unadj.values

# Draw posterior samples
n_sim <- 1000       # how many posterior predictive simulations  
idx <- sample(1:nrow(post_params), n_sim, replace = TRUE)
pp_params <- post_params[idx, ]

# Simulate data
pp_sims = 

# Compute summary stats
pp_sumstats <- t(sapply(pp_sims, compute_sumstats)) 


# Compare predictive distribution vs observed summary stats
pp_check <- abc::gfit(target = obs_stats,
                      sumstat = sumstats_matrix,
                      nb.repl = 1000)  # posterior predictive sims
plot(pp_check)

# Posterior predictive P values
pp_pvals <- colMeans(pp_sumstats >= matrix(obs_stats,
                                           nrow=n_sim,
                                           ncol=length(obs_stats),
                                           byrow=TRUE))
pp_pvals



```

## Figure

