# Prelude: Strand bias issue
There seems to be a potential strand bias issue: when inferring births/deaths through ancestral state reconstruction, there is an imbalance of deaths inferred from the 5' and 3' matrices: 4008/3584 deaths. Does this arrise from false negatives in the 5' or false positives in the 3' matrix? 

- compare distribution of 5' and 3' support
- ratio of 5' vs 3' support per strain

```{r}
anchors <- read.csv('workflow/results/detettore/ALL_anchors.tsv', sep='\t')
head(anchors)

sb = data.frame()

for (s in unique(anchors$strain)){

  anchors_strain = subset(anchors, strain == s)

  x = by(anchors_strain$num_reads, anchors_strain$side, median)

  five = ifelse('5' %in% names(x), x[['5']], NA)
  three = ifelse('3' %in% names(x), x[['3']], NA)
  
  row = data.frame(
    strain = s,
    median_reads_five = five,
    median_reads_three = three
  )
  sb = rbind(sb,row)
}

sb$diff = sb$median_reads_five-sb$median_reads_three


ggplot(sb, aes(diff)) + geom_histogram()
summary(sb$diff)


ggplot(sb, aes(median_reads_five, median_reads_three)) + 
  geom_point() + 
  geom_smooth() + geom_abline(slope=1, intercept=0, col='red')

wilcox.test(sb$median_reads_five, sb$median_reads_three, paired = TRUE, alternative = "two.sided")

```

## GC bias?
Is there a GC difference in the 5' and 3' context? -> Parallels with motif analysis
```{r}


```

Issue further explored in the birth/death section, where per branch birth/death events are inferred. 
