---
title: "detettore figures"
output: html_document
date: "2025-04-14"
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Prologue
```{r}
source('workflow/notebooks/utils.R')

# Load results
if (!exists('matrices', envir = .GlobalEnv)){
  out <- load_results()
  tree = out$tree
  metadata = out$metadata
  matrices = list('5' = out$matrix5, '3' = out$matrix3)
  matrix_metadata = out$matrix_meta
  refins = out$reference_insertions
  anchor_map = out$anchor_map
  copy_numbers = out$copy_numbers
  rm(out)
  save.image()
}

# Set plotting theme (not useful, ggsave onyl renders correct font when specified in the plot itself)
library(ggplot2)
plot_theme = theme_bw(base_family = "CMU Sans Serif", base_size=12)
theme_set(plot_theme)


# Load color schemes and define lineage order
schemata <- load_color_schemes()
lineage_order = c('L1', 'L2', 'L3', 'L4', 'L5', 'L6', 'L7', 'L8', 'L9', 'L10','La1', 'La2', 'La3', 'LaX', 'Dassie bacillus')


# Change names of lineages C and D
metadata$lineage = metadata$LINEAGE_x
metadata$sublineage = metadata$SUBLINEAGE_x
metadata$lineage[which(metadata$lineage=='C')] <- 'LaX'
metadata$lineage[which(metadata$lineage=='D')] <- 'Dassie bacillus'


# Add lineage to copy_numbers
copy_numbers$lineage = metadata$lineage[match(copy_numbers$strain, metadata$GNUMBER)]
copy_numbers$sublineage = metadata$sublineage[match(copy_numbers$strain, metadata$GNUMBER)]


# Add information to gene metadata: essentiality, distance to origin, resistance
genemap = read.delim("data/MTBC0/mtbc0_to_h37rv.genemap.tsv", sep='\t')
genemap = augment_genemap(
  genemap,
  'manuscript/literatura/DeJesus2017/orfs.csv',
  'data/who_resistance/WHO-UCN-TB-2023.7-eng_genomic_coordinates.txt'
)

# Add blast info to matrix_metadata
matrix_metadata <- add_blast_info(matrix_metadata)

save.image('is6110.RData')

```

## Parse reference insertions
- add unique positional ID position_strand
- flag insertions where the proportion of reads NOT supporting the insertion is higher than 0.1
- add Rv names if there is an H37Rv ortholog
- add context category 'genic/intergenic'
```{r}
# Add lineage and sublineage
refins$lineage = metadata$lineage[match(refins$strain,metadata$GNUMBER)]
refins$sublineage = metadata$sublineage[match(refins$strain,metadata$GNUMBER)]

# Nr of unique insertion sites: some sites with same position, different strand!
refins$unique_id <- paste(refins$position, refins$strand, sep='-')

# Reference support vs. 5' support + 3' support
refsupp_threshold = 0.1
refins$support_ref_ratio = refins$support_ref / (refins$support_5+refins$support_3)
refins$filter <- rep('PASS', nrow(refins))
refins$filter[which(refins$support_ref_ratio > refsupp_threshold)] <- 'FAILED'

ggplot(subset(refins, support_ref_ratio > 0), aes(support_ref_ratio)) + geom_density() + theme_bw() + scale_x_log10() + geom_vline(aes(xintercept = refsupp_threshold), col='red', linetype ='dashed')

# Add Rv gene names
refins$rv_ortholog <- add_alternative_gene_names(refins$gene, genemap)

# Add genic/intergenic 
refins$context <- rep('genic', nrow(refins))
refins$context[grepl(';', refins$gene)] <- 'intergenic'
refins$context <- as.factor(refins$context)

save.image('is6110.RData')
```

## Rescale branch lengths to account for non-variable positions
Information from https://github.com/cstritt/large_variable_alignment/
```{r}
genome_size = 4411532  
excluded_positions = 673325
var_positions = 548052 # from workflow/results/phylogeny/snp_alignment.uniqueseq.phy

scaling_factor <- (genome_size - excluded_positions)/ var_positions
tree$edge.length = tree$edge.length * scaling_factor

save.image('is6110.RData')
```

## Get MRCA of lineages
```{r}
library(ape)
library(reshape2)
# Check monophyly and get mrca
mrcas <- list()
nomono <- list()

for (l in lineage_order){
  md_l <- subset(metadata, lineage==l)
  gnrs <- md_l$GNUMBER[which(md_l$GNUMBER %in% tree$tip.label)]
  
  is_mono <- is.monophyletic(tree, gnrs)
  mrca_node <- getMRCA(tree, gnrs)
  
  if (is_mono){
    mrcas[[l]] <- mrca_node
  }
  else {
    print(paste("Not monophyletic: ", l))
    all_descendants <- extract.clade(tree, mrca_node)$tip.label
    subtree <- keep.tip(tree, all_descendants)
    nomono[[l]] <- subtree
  }
}
mrcas
mrcas = as.data.frame(mrcas) %>% melt()
names(mrcas) <- c("lineage", "node")
mrcas


# Lineages 8, 10, and D are missing as they have a single sample
table(metadata$lineage)
subset(metadata, lineage == 'L8')
subset(metadata, lineage == 'L9')
subset(metadata, lineage == 'L10')
subset(metadata, lineage == 'Dassie bacillus')

# Use L9  mrca to label L9, 10 and Dassie bacillus, which are too close anyway but not monophyletic
singles <- data.frame(
  'lineage' = c('L8', 'L10', 'Dassie bacillus', 'L9/L10/Dassie'),
  'node' = c(
    which(tree$tip.label=='G26689'), 
    which(tree$tip.label=='G50409'),
    which(tree$tip.label=='G01494'),
    18206)
)

mrcas <- rbind(mrcas, singles)
mrcas

save.image('is6110.RData')

```

## Summarize hotspots
```{r}
site_summary <- read.csv('workflow/results/detettore/AS_5prime.site_summary.tsv', sep='\t')
site_summary$site_id <- gsub("X", "", site_summary$site_id)

hotspots <- data.frame()
refins_absent = data.frame()  # to further investigate: why are some refins not mapped to the matrix?


# Only consider 5' reads
for (i in 1:nrow(refins)){

  if (i%%1000 == 0){print(i)}

  row = refins[i,]
  region = row$gene
  
  if (row$strand == '+'){ua = subset(anchor_map, V1==row$strain & V2==row$anchor_5)$V3}
  else {ua = subset(anchor_map, V1==row$strain & V2==row$anchor_3)$V3}

  if (length(ua) == 0){
    refins_absent = rbind(refins_absent, row)
    next 
    }  

  if (ua %in% hotspots$site_id){next}

  ua_asr = subset(site_summary, site_id == ua)
  ua_meta = subset(matrix_metadata, site_id == ua)
  
  dfrow = data.frame(
    'region' = region,
    'site_id' = ua,
    'position' = ua_meta$ref_position,
    'births' = ua_asr$births,
    'deaths' = ua_asr$deaths,
    'nstrains' = ua_meta$nr_strains,
    'blasthit' = ua_meta$blasthit,
    'rv_homolog' = row$rv_ortholog,
    'cluster' = ua_meta$cluster
  )
  hotspots = rbind(hotspots, dfrow)
}

save.image('is6110.RData')
```

## Clean up Rdata

```{r}
rm(singles, row, md_l, ua_asr, ua_meta)
rm(site_summary)
rm(dfrow))
save.image('is6110.RData')
```

# A: method & benchmarking

## Create fake tree and matrix for method figure
```{r}
library(ape)
library(phytools)
library(ggtree)
library(ggplot2)


# 1. Simulate a random phylogenetic tree with 20 samples
set.seed(123)  # for reproducibility
n_taxa <- 50
tree <- rtree(n_taxa)

# Plot the tree
plot(tree, main = "Simulated Phylogenetic Tree")
tree <- root(tree, 't1')

# Simulate 10 traits evolving on the tree
Q <- matrix(c(-0.1, 0.1,
              0.1, -0.1), 
            nrow = 2, byrow = TRUE)

rownames(Q) <- colnames(Q) <- c("0", "1")

n_traits <- 40
trait_matrix <- matrix(NA, nrow = n_taxa, ncol = n_traits)

for (i in 1:n_traits) {
  trait <- sim.Mk(tree, Q = Q, anc = "0")
  trait_matrix[, i] <- as.numeric(trait)
}
rownames(trait_matrix) <- tree$tip.label
colnames(trait_matrix) <- paste0("Trait_", 1:n_traits)

trait_matrix <- apply(trait_matrix, c(1, 2), as.character)

# Plot
p <- ggtree(tree)

pm <- gheatmap(p, trait_matrix, colnames = FALSE) +  
  scale_fill_manual(values=c('2'='black', '1'='white')) +
  theme(legend.position = "none") +
  scale_x_reverse()

pm

```



## Benchmarking: calculate sensitivity and precision
```{r}
library(ggplot2)
library(magrittr)
library(betareg)
library(margins)
library(cowplot)
library(RColorBrewer)
library(wesanderson)
library(viridis)

#input_table <- snakemake@input[[1]]
#output_html <- snakemake@output[[1]]

input_table <- 'workflow/results/benchmarking/detettore/benchmark_summary.tsv'

bm <- read.table(input_table, sep='\t', header=T)

bm$assembly <- as.factor(bm$assembly)
bm$coverage <- as.factor(bm$coverage)
bm$readlength <- as.factor(bm$readlength)
bm$side <- as.factor(bm$side)

# Recall/sensitivity and precision, false discovery rate
bm$recall = bm$true_positives / (bm$true_positives + bm$false_negatives)
bm$precision = bm$true_positives / (bm$true_positives + bm$false_positives)
bm$fdr = bm$false_positives / (bm$true_positives + bm$false_positives)

# Number of unpaired inverted repeats
bm$unpaired = bm$unpaired_IRL + bm$unpaired_IRR
```


## Check quality of IS annotation
Only include assemblies where the IS copy number equals the number of paired IRs. 
This excludes the 3 assemblies where IS copies are fragmented. I'm not aware of
biological processes that would lead to the presence of fragmented, active (!) IS 
in the genome; the other assemblies don't show this pattern; and, finally, the
three affected assemblies are all from the same source and based on Ion Torrent 
short reads.

```{r}
# Copy nr vs. nr of unpaired IRs
ggplot(bm, aes(copy_number, unpaired)) + geom_point() + theme_bw()
bm[which(bm$unpaired>0), 1] %>% unique() %>% as.character()

# Copy nr vs. nr of paired IS
figS1 <- ggplot(bm, aes(copy_number, paired_IRs)) + 
  geom_point() + theme_bw(base_family = "CMU Sans Serif", base_size=12) +
  xlab("Number of annotated intact IS6110 copies") + 
  ylab("Number of paired inverted repeats")

ggsave("figures/supplement/figS1.fragmented_copies.pdf", figS1, width = 12, height = 10, units = "cm", device = cairo_pdf)     


# Exclude 3 assemblies
exclude <- bm[which(bm$paired_IRs != bm$copy_number), 1] %>% unique() %>% as.character()
exclude
bm_filt <- subset(bm, !(assembly %in% exclude)) %>% droplevels()

```

## Describe benchmarking data set
```{r}
# How many IS copies in total?
bm_copynr <- bm_filt[!duplicated(bm_filt$assembly), ]
print(paste('Number of IS6110 elements analyzed for benchmarking: ', sum(bm_copynr$copy_number)))

```


## Figure S2: sensitivity and precision

```{r}
library(ggplot2)
library(cowplot)
library(reshape2)

head(bm_filt)

bm_filt_fig <- melt(bm_filt, id.vars=c('coverage', 'side','readlength'), measure.vars=c('precision', 'recall'))

bm_filt_fig$variable = as.character(bm_filt_fig$variable)
bm_filt_fig$variable[which(bm_filt_fig$variable=='recall')] <- 'Sensitivity'
bm_filt_fig$variable[which(bm_filt_fig$variable=='precision')] <- 'Precision'
bm_filt_fig$variable = factor(bm_filt_fig$variable, levels=c('Sensitivity', 'Precision'))


bm_filt_fig$readlength = as.character(bm_filt_fig$readlength)
bm_filt_fig$readlength[which(bm_filt_fig$readlength=='100')] <- '100 bp reads'
bm_filt_fig$readlength[which(bm_filt_fig$readlength=='150')] <- '150 bp reads'
bm_filt_fig$readlength = as.factor(bm_filt_fig$readlength)

# Reorder factor levels
bm_filt_fig$side = factor(bm_filt_fig$side, levels=c('5', '3'))
str(bm_filt_fig)


figS2 <- ggplot(bm_filt_fig, aes(y=value, x=coverage, fill=side)) +
  geom_boxplot(outlier.size = 0.5) + facet_grid(variable~readlength) +
  theme_bw(base_family = "CMU Sans Serif", base_size=12) +
  xlab('Coverage') + ylab('') + labs(fill = "Side")

figS2

ggsave("figures/supplement/figS2.benchmarking.pdf", figS2, width = 15, height = 12, units = "cm",device = cairo_pdf)     

```



## Sensitivity
Plot sensitivity against coverage. Show mean recall for all parameter combinations. Fit GLM to see which parameters affect recall.
```{r}
# Line plot
bm_filt %>% 
  ggplot(aes(y=recall, x=coverage, col=side,group=interaction(assembly,side))) + 
  geom_point(alpha=0.5) + geom_line(alpha=0.5) + 
  facet_wrap(~readlength) + 
  theme(text = element_text(size = 12)) +
  theme_bw() + ylab('Sensitivity') + xlab('Coverage') +
  scale_color_brewer(palette='Dark2')

# Boxplot
bm_filt  %>% 
  ggplot(aes(y=recall, x=coverage, fill=side)) + 
  geom_boxplot(alpha=0.5) + 
  facet_wrap(~readlength) +
  theme(text = element_text(size = 12)) +
  theme_bw() + ylab('Sensitivity') + xlab('Coverage') +
  scale_fill_brewer(palette='Dark2')

aggregate(recall~coverage+readlength+side, bm_filt, mean)

```

## How many unpaired IRs are there?
False positives are cause by IS6110-specific IRs that are not part of a full element.
```{r}
bm_filt_hq <- subset(bm_filt, coverage==50 & readlength == 150 & side == 3)
bm_filt_hq

unpaired_IRs <- sum(bm_filt_hq$unpaired_IRL) + sum(bm_filt_hq$unpaired_IRR)
print(paste('Number of unpaired inverted repeats in the assemblies used for benchmarking: ', unpaired_IRs))

```


## What affects sensitivity?
Use beta regression to model a proportion:

- slightly change 0s and 1s to avoid boundary issues
- weight by copy number to give less weight to low CN strains
- coefficients from beta regression are log odds. Exponentiate and get marginal effects for interpretation

Results: coverage of 20 can produce quite some false negatives, leading to very low
sensitivity in low-CN strains. Otherwise there seem to be single elements in single
assemblies that are occasionally missed, but overall sensitivity is high!

```{r}
# adjust 0s and 1s
bm_filt$recall_adj <- (bm_filt$recall * (nrow(bm_filt) - 1) + 0.5) / nrow(bm_filt)

m_recall_beta <- betareg(recall_adj ~ assembly + coverage + readlength + side, data = bm_filt, weights=bm_filt$copy_number)

summary(m_recall_beta)

# Get odds from log odds ration
exp(coef(m_recall_beta))

# Marginal effects
m_recall_beta_margins <- margins(m_recall_beta)
summary(m_recall_beta_margins)

```


## Same for precision
```{r}

# Line plot
figS2b <- bm_filt %>% 
  ggplot(aes(y=precision, x=coverage, col=side,group=interaction(assembly,side))) + 
  geom_point(alpha=0.5) + geom_line(alpha=0.5) + 
  facet_wrap(~readlength) + 
  theme(text = element_text(size = 12)) +
  theme_bw() + ylab('Precision') + xlab('Coverage') +
  scale_color_brewer(palette='Dark2')

figS2c

# Boxplot
figS1d <- bm_filt %>% 
  ggplot(aes(y=precision, x=coverage, fill=side)) + 
  geom_boxplot(alpha=0.5) + 
  facet_wrap(~readlength) +
  theme(text = element_text(size = 12)) +
  theme_bw() + ylab('Precision') + xlab('Coverage') +
  scale_fill_brewer(palette='Dark2')

figS2d

aggregate(precision~coverage+readlength+side, t_filt, mean)


# Single assemblies affect precision, especially ASM25337
subset(bm_filt, side =='5' & false_positives > 0) %>% 
  ggplot(aes(y=as.factor(false_positives), x=copy_number, col=assembly)) +
  geom_jitter(width=0.05, height = 0.05, alpha=0.8) + xlab('Copy number') + ylab('False positives') +
  scale_x_continuous(breaks = seq(0, 25, by = 2)) +
  scale_color_brewer(palette='Set1') +
  theme_bw()


```


## What affects precision?

Results: there is a single case of 3 false positives, else max FP is 2. The assembly is the 
main factor affecting precision, read length and coverage have nothing to do with it. In these
few assemblies with false positives, the 3' side is for some reason less precise. 

Whether these effects are due to misassembly or IS copies that have diverged from the 
target IS through some biological process is unclear. These are exceptions, however; mean precision
is overall high at 97%. 

```{r}
#adjust 0s and 1s
t_filt$precision_adj <- (t_filt$precision * (nrow(t_filt) - 1) + 0.5) / nrow(t_filt)

m_precision_beta <- betareg(precision_adj ~ assembly + coverage + readlength + side, data = t_filt, weights=t_filt$copy_number)

summary(m_precision_beta)

# Get odds from log odds ration
exp(coef(m_precision_beta))

# Marginal effects!
m_precision_beta_margins <- margins(m_precision_beta)
summary(m_precision_beta_margins)

```

## Are false positives due to unpaired IRs?
Only in one assembly (ASM25337). This strain has 2 IS6110 copies and a pair of dispersed IRL and IRR. 

```{r}
subset(t_filt, side == '5' & (false_positives > 0 | unpaired > 0)) %>% 
  ggplot(aes(x=false_positives, y=unpaired, col=assembly)) + geom_jitter(width = 0.1, height = 0.1) 

```


## Number of supporting reads

- How many supporting reads can we expect at different coverages?
- How do z-scores behave when there are few observations?

```{r}
library(ggplot2)

anchors <- read.csv('workflow/results/benchmarking/detettore/ALL_anchors.tsv', sep='\t')
head(anchors)

# Split sample name
split_col <- strsplit(anchors$strain, "_", fixed = TRUE)
split_df <- as.data.frame(do.call(rbind, split_col))
colnames(split_df) <- c("ASM", "COV", "RL")
anchors <- cbind(anchors, split_df)

# Plot distributions of supporting reads
ggplot(anchors, aes(x=COV, y=num_reads, col=RL)) + 
  geom_boxplot(size=1)  + geom_jitter(alpha=0.1) +
  theme_bw()


# Plot distributions of z-scores
ggplot(anchors, aes(x=COV, y=mod_z_score,col=RL)) + 
  geom_violin(size=1)  + geom_jitter(alpha=0.1) +
  theme_bw()


# Classifier to check if num_reads and/or z-score can identify false positives
library(randomForest)
model_rf <- randomForest(as.factor(is_false_positive) ~ num_support + z_score, data=df, ntree=500)
print(model_rf)

```






# B: Results overview




## Table 1: lineagae summary
- add diversity (calculated from tree, e.g. recycle pyseer similiarity matrix)
```{r}
library(gt)
library(magrittr)
library(tidyverse)
library(ape)
library(ggplot2)

table1 <- copy_numbers %>%
  group_by(lineage) %>%
  summarise(
    n = n(),
    mean = mean(CN),
    min = min(CN),
    max = max(CN),
    var = var(CN),
    sd = sd(CN)
  )

table1


# Add number of insertion events
table1$nr_events <- rep(0, nrow(table1))
table1$nr_insertions <- rep(0, nrow(table1))

for (i in 1:nrow(table1)){
  
  lineage = table1$lineage[i] %>% as.character()
    
  gnrs = metadata$GNUMBER[metadata$lineage == lineage]
  m_lineage = matrices$`5`[which(rownames(matrices$`5`) %in% gnrs),] 
  dim(m_lineage)
  
  # Remove columns with sum == 0
  if (length(gnrs) > 1){
    m_lineage <- m_lineage[, colSums(m_lineage) != 0]
    dim(m_lineage)
    
    if (is.null(ncol(m_lineage))){table1$nr_events[i] <- 1}
    else {table1$nr_events[i] <- ncol(m_lineage)}  
    table1$nr_insertions[i] <- sum(m_lineage)
  }
  
  else {
    table1$nr_events[i] <- sum(m_lineage)
    table1$nr_insertions[i] <- sum(m_lineage)
  }
}


# Add lineage SNP diversity
dm <- cophenetic(tree)
dim(dm)
dm_smry <- summary(dm[upper.tri(dm)])

gnrs = rownames(dm)

dists_lineages = data.frame()

for (l in lineage_order){
  gnrs_l = subset(metadata, lineage == l)$GNUMBER
  dm_li = match(gnrs_l, gnrs)
  dm_l = dm[dm_li, dm_li]
  pwd_l = dm_l[upper.tri(dm_l)]
  df_l = data.frame(
    dists = pwd_l,
    lineage = rep(l, length(pwd_l))
  )
  dists_lineages = rbind(dists_lineages, df_l)
}

rm(dm)

ggplot(dists_lineages, aes(x=lineage, y=dists)) + 
  geom_boxplot(outliers=F) +
  xlab("") + ylab('Pairwise phylogenetic distance')

median_pwd <- tapply(dists_lineages$dists, dists_lineages$lineage, median)
table1$diversity <- round(median_pwd[table1$lineage], 5)



# Arrange table
table1$nr_events_per_strain <- table1$nr_events/table1$n
table1 <- table1 %>%  arrange(factor(lineage, levels=lineage_order))

# Write to file
write.table(as.data.frame(table1), file = "workflow/results/table1.tsv", quote = FALSE, sep = "\t", row.names = F)

#table1 <- table1[-c(14, 15),]
table1
table1 <- table1[,c("lineage", 'n', 'diversity', 'nr_events', 'nr_insertions', 'mean',  'sd','min', 'max')]

names(table1) <- c('Lineage', 'Strains', 'Diversity', 'Insertion sites', 'Total insertions', 'Mean', 'SD', 'Min', 'Max')

gt_tbl <- 
  gt(table1) |>
  fmt_number(columns= c(Strains, `Insertion sites`, `Total insertions`, Min, Max), decimals=0) |>
  fmt_number(columns= c(Mean, SD), decimals=1) |>
  tab_header(
    title = md("**Table 1**. IS6110 insertions in 10,000 strains")) |>
  tab_spanner(label = "Copy number",columns = c(Mean, SD, Min, Max)) |>
  tab_options(table_body.hlines.style = "none") |>
  opt_table_font(font='CMU Sans Serif', stack = "transitional")  |>
  opt_align_table_header("left") |>
  tab_footnote(
    footnote = "Diversity is calculated as the median pairwise phylogenetic distance within lineages.",
    locations = cells_column_labels(columns = "Diversity")) |>
  tab_footnote(
    footnote = md("This clade comprises *M. microti* and *M. pinnipedii*."),
    locations = cells_body(
      rows = table1$Lineage == "LaX",  # Replace "X" with the specific value
      columns = vars(Lineage)
    ))
  
gt_tbl

gt_tbl_latex <- as_latex(gt_tbl)
as.character(gt_tbl_latex) %>% cat(file="figures/main/table1.tex")

```

## Compare 5' and 3' matrices

```{r}

```







# C: copy numbers

Fig. 2. a) Phylogeny of the 10,000 selected strains, estimated with ... The colors of the
tips indicated the copy number. Lineage labelling follows the 'modern' terminology, 
as summarized in ..., b) copy number distributions by lineage 

## Ancestral state reconstruction with copy numbers!
```{r}
library(castor)
library(treeio)
library(ggtree)
library(viridis)

head(copy_numbers)
summary(copy_numbers)

# Reorder matrix rows so they match the tree!
phyl_order <- match(tree$tip.label, copy_numbers$strain)
copy_numbers_ordered <- as.integer(copy_numbers[phyl_order,]$CN)
copy_numbers_ordered <- copy_numbers_ordered + 1


# Convert to numeric starting with 1
site_vec_mapped <- map_to_state_space(copy_numbers_ordered)

plot(density(site_vec_mapped$mapped_states))  

# Reconstruct ancestral states
asr_result <- asr_max_parsimony(
  tree, 
  tip_states = copy_numbers_ordered
  )

head(asr_result$ancestral_likelihoods)
summary(asr_result$ancestral_states)

ancestral_states = c()
for (i in nrow(asr_result$ancestral_likelihoods)){
  lhs = asr_result$ancestral_likelihoods[i,]
  max_i = which(lhs==max(lhs))
  if (length(max_i) > 1){anc_state = NA}
  else {anc_state = max_i -1 }
  ancestral_states = c(ancestral_states, anc_state)
}
ancestral_states


# Plot on tree!
p <- ggtree(tree, layout='circular',size=0.1, alpha=0.5)  %<+% metadata
p$data$copy_number <- c(copy_numbers_ordered, asr_result$ancestral_states)
#p + geom_tree(aes(col=copy_number),size=0.1, alpha=0.5) +  scale_color_viridis(option='turbo')

write.table(p$data, 'workflow/results/phylogeny/treedata.CN_ASR.tsv', sep='\t')

td <- as.treedata(p)
write.beast(td, 'workflow/results/phylogeny/snp_alignment.rooted.treefile.annotated') 


```


## Tree and boxplot
```{r}
library(ape)
library(ggtree)
library(RColorBrewer)
library(viridis)
library(reshape2)
library(cowplot)
library(magick)
library(ggplot2)

# Fig. 2a: tree -----------------------------------------------------------


# Color copy number, label lineages
metadata$cn <- as.numeric(copy_numbers$CN[match(metadata$GNUMBER, copy_numbers$strain)])

t <- ggtree(tree, layout="circular",size=0, alpha=0) %<+% metadata
# Add CN from ancestral state reconstruction above!
t$data$copy_number = c(copy_numbers_ordered, asr_result$ancestral_states)

t$data$lineage[which(t$data$lineage %in% c('L9', 'L10', 'Dassie bacillus'))] <- 'L9/L10/Dassie'

fig2a <- t + geom_tree(aes(col=copy_number),size=0.1, alpha=0.8) +
  #t + geom_tippoint(aes(col=cn), size=0.7, alpha=0.4, stroke=NA) +
  theme_void(base_family = "CMU Sans Serif", base_line_size = 10) + 
  theme(
    legend.position=c(0.88, 0.35), 
    legend.key.width = unit(2, "mm"),
    legend.title = element_text(size = 4),
    legend.text = element_text(size = 4),
    legend.title.position='bottom',
    legend.key.size = unit(3, "mm")) +
  geom_cladelab(
    data = subset(mrcas, !(lineage %in% c('L9', 'L10', 'Dassie bacillus'))), 
    mapping=aes(node = node, label = lineage),
    fontsize = 1.5, barsize=0.1, offset.text=0.0008,color='grey20') +
  scale_color_viridis(option = 'turbo') +
  labs(col = "IS6110 copy number")

fig2a


# Crop white space
ggsave("figures/main/fig2a.pdf", fig2a,
       width = 13, height = 13, units = "cm",
       device = cairo_pdf)      # full-vector

system("pdfcrop --margins 5 figures/main/fig2a.pdf figures/main/fig2a_crop.pdf")





# 2b: copy number distribution ----------------------------------------------
brewer.pal(3, "Dark2")
brewer.pal(3, "Set1")

copy_numbers$lineage <- factor(copy_numbers$lineage, levels = lineage_order)

copy_numbers$lineage_new <- as.character(copy_numbers$lineage)
copy_numbers$lineage_new[which(copy_numbers$lineage=='Dassie bacillus')] <- 'Db'
copy_numbers$lineage_new <- factor(copy_numbers$lineage_new, levels = c(lineage_order, 'Db'))



# Replace Dassie bacillus with Db
#copy_numbers$lineage[which(copy_numbers$lineage=='Dassie bacillus')] <- 'Db'


fig2b <- ggplot(copy_numbers, aes(y=CN,x=lineage_new)) + 
  geom_jitter(alpha=0.3, size=1, color="grey50", stroke=NA) + 
  geom_boxplot(alpha=0,outlier.shape = NA) +
  annotate("text",
           x = 1:length(table(copy_numbers$lineage)),
           y = 38,
           label = table(copy_numbers$lineage),
           col = "grey50", size=2) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 10) +
  xlab("Lineage") + ylab("IS6110 copy nr.")

fig2b


# 2c new: frequency distribution of CN
fig2c <- ggplot(copy_numbers, aes(CN)) + geom_histogram() +
  theme_bw(base_family = "CMU Sans Serif", base_size = 10) + 
  xlab('Copy number') + ylab('Nr. of strains')

fig2c



# Combine panels --------------------------------------------------------------
img_tree <- image_read("figures/main/fig2a_crop.pdf", density = 600)
p_tree  <- ggdraw() + draw_image(img_tree)

lower_panel <- plot_grid(
  fig2c, fig2b,nrow = 1,
  labels=c('B', 'C'),
  rel_widths = c(1,2),
  label_size = 10,
  label_fontfamily='CMU Sans Serif'
  )

plot_grid(p_tree, lower_panel, 
          labels = c('A', ''),
          label_size = 10,
          ncol=1, 
          rel_heights=c(3,1),
          label_fontfamily='CMU Sans Serif')


ggsave("figures/main/fig2.tree_copy-numbers.pdf",
       width = 15, height = 17, units = "cm",
       device = cairo_pdf)      # full-vector

```


## Sublineage copy numbers 
```{r}
library(dplyr)

l='L1'
cn_subl <- subset(copy_numbers, lineage==l) %>% 
  group_by(sublineage) %>% 
  summarise(mean_value = mean(CN))

print(cn_subl, n=nrow(cn_subl))
```


## Do generalists have higher CNs?
Plot CN in lineage 4 sublineages!
```{r}
l4_cn <- subset(out$metadata, LINEAGE_x=='L4')

ggplot(l4_cn,aes(y=cn,x=SUBLINEAGE_x)) + 
  geom_jitter(alpha=0.3, size=1, color="#377EB8", stroke=NA) + 
  geom_boxplot(alpha=0,outlier.shape = NA) +
  annotate("text",
           x = 1:length(table(l4_cn$SUBLINEAGE_x)),
           y = 38,
           label = table(l4_cn$SUBLINEAGE_x),
           col = "grey50", size=2) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 10) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  xlab("Sublineage") + ylab("IS6110 copy number")
```


## Burden analysis 
```{r}
library(ggplot2)
library(magrittr)
library(ggrepel)
library(viridis)
library(cowplot)

burden_snps_indels <- read.csv('workflow/results/gwas/burden_snps_indels.tsv', sep='\t')
burden_snps_indels <- burden_snps_indels[,-c(8,9 )]
burden_snps_indels$typus <- rep('snps_indels', nrow(burden_snps_indels)) 
burden_snps_indels$index <- 1:nrow(burden_snps_indels)
head(burden_snps_indels)

burden_is6110 <- read.csv('workflow/results/gwas/burden_is6110.tsv', sep='\t')
burden_is6110 <- burden_is6110[,-c(8,9 )]
burden_is6110$typus <- rep('is6110', nrow(burden_is6110)) 
burden_is6110$index <- 1:nrow(burden_is6110)
head(burden_is6110)


load_permutation_res <- function(permutation_results){
  # Load permutation results
  permut <- data.frame()
  for (f in permutation_results){
    df <- read.csv(f, sep='\t')
    df <- df[,-c(8,9,10,11)]
    bn <- basename(f)
    rep <- strsplit(bn, '\\.')%>% unlist()
    n=rep[2]
    df$n <- rep(n, nrow(df))
    permut <- rbind(permut, df)
  }
  permut$logp <- -log10(permut$lrt.pvalue)
  return(permut)
}

permutations_snps_indels <- load_permutation_res(list.files('workflow/results/gwas/permutations', pattern="snps_indels*", full.names = T))
permutations_is6110 <- load_permutation_res(list.files('workflow/results/gwas/permutations', pattern="is6110*", full.names = T))



# Get threshold from permutations: 0.1, 1 and 5 percentiles
threshoolds_snps_indels <-  quantile(permutations_snps_indels$lrt.pvalue, probs=c(0.001, 0.01, 0.05))
threshoolds_is6110 <-  quantile(permutations_is6110$lrt.pvalue, probs=c(0.001, 0.01, 0.05))


burden_comb <- rbind(burden_snps_indels, burden_is6110)
burden_comb$logp <- -log10(burden_comb$lrt.pvalue)


# Manhatten plot
a_r3 <- ggplot(burden_comb, aes(x=index, y=logp)) + 
  geom_point() + theme_bw(base_family = "CMU Sans Serif") + ylab('-Log10(P)') + xlab('') +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1, size=8)) +
  facet_wrap(~typus, scales='free_x') +
  geom_hline(aes(yintercept=-log10(threshoolds_is6110[[1]])), col='red', linetype='dashed')

      #geom_boxplot(data=permut, aes(x=variant, y=logp), outlier.shape = NA, size=0.1)+

# Volcano plot
b_r3 <- ggplot(burden_comb, aes(y=logp, x=beta, col=100*variant_h2)) + geom_point(alpha=1) +
  geom_hline(aes(yintercept=-log10(threshoolds_is6110[[1]])), col='red', linetype='dashed') +
  theme_bw(base_family = "CMU Sans Serif") + ylab('-Log10(P)') + xlab('Effect size (beta)') +
  scale_color_viridis(option='turbo', name='Var. expl.') +
  theme(
    legend.position = c(1, 0),              # Bottom right inside the plot
    legend.justification = c("right", "bottom"),
    legend.background = element_rect(fill = NA, color = NA),
    legend.key.height = unit(0.5, "cm"),          # Smaller legend key
    legend.key.width = unit(0.5, "cm"),
    legend.text = element_text(size = 8),         # Smaller text
    legend.title = element_text(size = 8)
  )

fig_gwas_r3 <- plot_grid(
  a_r3,b_r3, 
  labels = c('A', 'B'), 
  rel_widths = c(1.5,1),
  label_fontfamily='CMU Sans Serif')

fig_gwas_r3
```






## Burden analysis with all genes
```{r}
library(ggplot2)
library(magrittr)
library(ggrepel)
library(viridis)
library(cowplot)

burden <- read.csv('workflow/results/gwas/burden_all_genes.tsv', sep='\t')
burden <- burden[,-c(8,9 )]

permutation_results_all <- list.files('workflow/results/gwas/permutations', pattern="burden_all*", full.names = T)

# Load permutation results
permut_all <- data.frame()
for (f in permutation_results_all){
  df <- read.csv(f, sep='\t')
  df <- df[,-c(8,9,10,11)]
  bn <- basename(f)
  rep <- strsplit(bn, '\\.')%>% unlist()
  n=rep[2]
  df$n <- rep(n, nrow(df))
  permut_all <- rbind(permut_all, df)
}

permut_all$logp <- -log10(permut_all$lrt.pvalue)


# Get threshold from permutations: 1 and 5 percentiles
thresholds_burden = quantile(permut_all$lrt.pvalue, probs=c(0.01, 0.05))
burden$logp <- -log10(burden$lrt.pvalue)
sign_burden <- -log10(thresholds_burden[[1]])

burden$i <- 1:nrow(burden)

# Manhatten plot
a_burden <- ggplot(burden, aes(x=i, y=logp)) + 
  geom_point(alpha=0.3) + theme_bw(base_family = "CMU Sans Serif") +
  #geom_point(data=subset(burden, variant %in% burden_r3$variant), aes(col='R3 genes')) +
  geom_hline(aes(yintercept=sign_burden), col='red', linetype='dashed') +
  scale_color_manual(values = c("R3 genes" = "red"),name=NA) +
  ylab('-Log10(P)') + xlab('') +
  theme(
    legend.position = c(0.95, 0.95),             
    legend.justification = c("right", "top"),
    legend.background = element_rect(fill = 'grey89', color = NA),
    legend.title=element_blank(),
    legend.key = element_rect(fill = 'grey89', color = NA)
    )

a_burden

# Volcano plot
b_burden <- ggplot(burden, aes(y=logp, x=beta, col=100*variant_h2)) + geom_point(alpha=1) +
  geom_hline(aes(yintercept=sign_burden), col='red', linetype='dashed') +
  theme_bw(base_family = "CMU Sans Serif") + ylab('-Log10(P)') + xlab('Effect size (beta)') +
  geom_label_repel(data=subset(burden, logp>sign_burden),aes(label=variant), label.size=0, size = 2,col='black') +
  scale_color_viridis(option='turbo', name='Var. expl.') +
  theme(
    legend.position = c(1, 0),             
    legend.justification = c("right", "bottom"),
    legend.background = element_rect(fill = NA, color = NA),
    legend.key.height = unit(0.5, "cm"), 
    legend.key.width = unit(0.5, "cm"),
    legend.text = element_text(size = 8),
    legend.title = element_text(size = 8)
  )

fig_gwas_all <- plot_grid(
  a_burden,b_burden,
  labels = c('A', 'B'), 
  rel_widths = c(1.5,1))

fig_gwas_all

ggsave("figures/fig3.gwas_burden_all_genes.pdf",
       width = 20, height = 10, units = "cm",
       device = cairo_pdf)      # full-vector

burden <- burden[with(burden, order(-logp)),]
head(burden)

burden[which(burden$logp > sign_burden & burden$beta > 1), ]$variant



```

## GWAS with all non-synonymous variants
```{r}
gwas_snps <- read.csv('workflow/results/gwas/burden_snps_indels.tsv', sep='\t')
gwas_snps <- gwas_snps[,-c(8,9)]
gwas_snps$logp <- -log10(gwas_snps$lrt.pvalue)
gwas_snps$pos <- as.numeric(gsub(".*_([0-9]+).*", "\\1", gwas_snps$variant))


gwas_is6110 <- read.csv('workflow/results/gwas/burden_is6110.tsv', sep='\t')
gwas_is6110 <- gwas_is6110[,-c(8,9)]
gwas_is6110$logp <- -log10(gwas_is6110$lrt.pvalue)
gwas_is6110$pos <- as.numeric(gsub(".*_([0-9]+).*", "\\1", gwas_is6110$variant))



sign_snps = 0.01
sign_corr = sign_snps/nrow(gwas_snps)
sign_corr_logp = -log10(sign_corr)


# Load variant metadata
var_dict = read.csv("data/major_variants.tsv", sep='\t')
head(var_dict)
var_dict$pyseer_id = paste(var_dict$chromosome, var_dict$position, var_dict$REF, var_dict$ALT, sep='_')
gwas_snps$variant %in% var_dict$pyseer_id
gwas_snps$gene = var_dict$gene[match(gwas_snps$variant, var_dict$pyseer_id)]

gene_split <- strsplit(gwas_snps$gene, '_')
gwas_snps$gene_rv <- sapply(gene_split, `[`, 2)




# Remove duplicate frequencies
#gwas_snps <- gwas_snps[!duplicated(gwas_snps$lrt.pvalue), ]

a_snps <- ggplot(gwas_snps, aes(x=pos, y=logp)) + 
  geom_point(alpha=0.2) +
  theme_bw(base_family = "CMU Sans Serif") + 
  ylab('-Log10(P)') + xlab('Position') + 
  geom_hline(aes(yintercept=sign_corr_logp), col='red', linetype='dashed')

# volcano plot
b_snps <- ggplot(gwas_snps, aes(y=logp, x=beta, col=100*variant_h2)) + 
  geom_point(alpha=0.5) +
  geom_hline(aes(yintercept=sign_corr_logp), col='red', linetype='dashed') +
  theme_bw(base_family = "CMU Sans Serif") + ylab('-Log10(P)') + xlab('Effect size (beta)') +
  scale_color_viridis(option='turbo', name='Var. expl.') +
  theme(
    legend.position = c(1, 0),              # Bottom right inside the plot
    legend.justification = c("right", "bottom"),
    legend.background = element_rect(fill = NA, color = NA),
    legend.key.height = unit(0.5, "cm"),          # Smaller legend key
    legend.key.width = unit(0.5, "cm"),
    legend.text = element_text(size = 8),         # Smaller text
    legend.title = element_text(size = 8)
  )

plot_grid(
  a_snps,b_snps, 
  labels = c('A', 'B'), 
  rel_widths = c(1.5,1))

# Check top 10 SNPs
gwas_snps <- gwas_snps[with(gwas_snps, order(-logp)),]
head(gwas_snps,10)
topgenes <- subset(gwas_snps, logp>10)$gene_rv






# Plot hit on tree
#library(magrittr)
#gwas_snps <- read.csv('workflow/results/gwas/all_variants.tsv', sep='\t')
#x <- hit_on_tree(gwas_snps, out$tree, out$copy_numbers,out$metadata,'MTB_anc_1107917_G_T')

```


# D: birth and death

## Frequency distribution
```{r}
library(ggplot2)
library(reshape2)
library(magrittr)

counts <- colSums(matrices$`5`)
sfs5 <- data.frame(table(counts))

counts <- colSums(matrices$`3`)
sfs3 <- data.frame(table(counts))

sfs <- merge(sfs5, sfs3, by = "counts", all.x = TRUE, all.y=TRUE)
names(sfs) <- c('counts', 'fiveprime' , 'threeprime')
summary(sfs)

sfs$counts <- as.numeric(sfs$counts)

# Correlation 5' and 3'
ggplot(sfs, aes(fiveprime, threeprime)) + geom_point() +
  geom_abline(slope=1, intercept=0) +
  theme_bw()


# SFS plot
melt(sfs, id.vars = c('counts')) %>% 
  ggplot(aes(x=counts, y=value)) + geom_bar(stat='identity') +
  ylab('Number of sites') + xlab('Number of strains')  +
  theme_minimal() + facet_wrap(~variable)

ggplot(sfs, aes(x=counts, y=threeprime)) + geom_bar(stat='identity') +
  ylab('Number of sites') + xlab('Number of strains')  +
  theme_minimal() + scale_y_log10()

```

## IS6110 diversity
```{r}
library(ggplot2)
library(magrittr)
library(ape)
library(reshape2)
library(Matrix)

table1 <- read.csv('workflow/results/table1.tsv', sep='\t')
table1$IS_diversity <- rep(NA, nrow(table1))

add_snp_dist = FALSE  # OOM!
subsample_n = 0

is_allele_spec <- function(mat){
  " Get binned allele frequencies from a matrix
  "
  nstrains = nrow(mat)
  
  # only consider shared alleles to make patterns more clear
  #mat <- mat[,colSums(mat) > 1]
  mat <- as.matrix(mat)
  nsites = ncol(mat)

  allelecounts <- colSums(mat)

  low =  table(allelecounts <= nstrains*0.25)
  lowmid = table(nstrains*0.25 < allelecounts & allelecounts <= nstrains*0.5)
  midhigh = table(nstrains*0.5 < allelecounts & allelecounts <= nstrains*0.75)
  high = table(nstrains*0.75 < allelecounts & allelecounts <= nstrains)

  prop_out = data.frame(
    'low' = tryCatch({low[['TRUE']]}, error = function(e){0})/nsites,
    'lowmid' =  tryCatch({lowmid[['TRUE']]}, error = function(e){0})/nsites,
    'midhigh' =  tryCatch({midhigh[['TRUE']]}, error = function(e){0})/nsites,
    'high' =  tryCatch({high[['TRUE']]}, error = function(e){0})/nsites,
    'nsites' = nsites
  )
  return(prop_out)
}


pairwise_diff_raw <- function(mat) {
  mat <- as.matrix(mat) * 1  # ensure numeric 0/1
  L <- ncol(mat)
  
  # Number of shared 1s between pairs
  shared_ones <- mat %*% t(mat)
  
  # Number of 1s in each strain
  ones_per_strain <- rowSums(mat)
  
  # Using: differences = ones_i + ones_j - 2 * shared_ones
  diff_mat <- outer(ones_per_strain, ones_per_strain, "+") - 2 * shared_ones
  
  return(diff_mat)  # full symmetric matrix of raw differences
}

to_table_tri <- function(mat, value_name="value") {
  ij <- which(upper.tri(mat), arr.ind = TRUE)
  data.frame(
    sample1 = rownames(mat)[ij[,1]],
    sample2 = rownames(mat)[ij[,2]],
    !!value_name := mat[ij]
  )
}

to_table <- function(mat, value_name = "value") {
  rn <- rownames(mat)
  
  # Upper triangle WITHOUT diagonal
  idx <- which(upper.tri(mat), arr.ind = TRUE)
  
  data.frame(
    sample1 = rn[idx[, 1]],
    sample2 = rn[idx[, 2]],
    value = mat[idx],
    row.names = NULL,
    check.names = FALSE
  ) |>
    dplyr::rename(!!value_name := value)
}

# Optionally add SNP distances (may required subsampling)
if (add_snp_dist){
  dm <- cophenetic(tree)
  gnrs = rownames(dm)
}

# Create data frame with pairwise IS6110 and SNP distances!
pairwise_differences <- data.frame()
shared_allele_proportions <- data.frame()


for (l in lineage_order){
  if (l %in% c('L8', 'L10', 'Dassie bacillus')){next}

  tm_sub <- subset_tree_matrix(tree, matrices$`5`, metadata, l, 'lineage') 
  mat_sub = tm_sub[[2]]
  diff_mat = pairwise_diff_raw(mat_sub)

  # Median pairwise IS differences
  median_pwd = median(diff_mat)
  table1$IS_diversity[which(table1$Lineage==l)] <- median_pwd
  print(paste(l, round(median_pwd,4)))

  # Frequency spectrum for shared insertions
  aspec <- is_allele_spec(mat_sub)
  aspec$lineage <- l
  shared_allele_proportions <- rbind(shared_allele_proportions, aspec)

  # Subsample if combining with SNP distances
  nstrains = length(tm_sub[[1]]$tip.label)
  if (subsample_n > 0){nsubsample = ifelse(nstrains<subsample_n, nstrains, subsample_n)}

  # Convert to table
  diff_df <- to_table(diff_mat, 'is6110_distance')
  if (subsample_n > 0){diff_df <- diff_df[sample(nrow(diff_df), nsubsample), ]}

  # Optionally add SNP distances
  if (add_snp_dist){
    # Add phylogenetic distance
    gnrs_l = unique(diff_df$sample1, diff_df$sample2)
    dm_li = match(gnrs_l, gnrs)
    dm_l = dm[dm_li, dm_li]  
    diff_df_snps <- to_table(dm_l, 'snp_distance')
    diff_df <- merge(diff_df, diff_df_snps, by = c('sample1', 'sample2'))
  }

  # Add lineage 
  diff_df$lineage = rep(l, nrow(diff_df))
  pairwise_differences <- rbind(pairwise_differences, diff_df)
}

by(pairwise_differences$is6110_distance, pairwise_differences$lineage, summary)

# Overall differences
#diff_mat_full = pairwise_diff_raw(matrices$`5`)
#median(diff_mat_full)

```


Figure X. A) Pairwise IS6110 diversity. B) Proportion and count of high frequency insertion sites. 


```{r}
library(ggrepel)


a <- ggplot(pairwise_differences, aes(lineage, is6110_distance)) + geom_boxplot(outlier.size=0.5, size=0.3) +
  theme_bw() + xlab("Lineage") + ylab("IS6110 pairwise differences") 


# Allele frequency spectrum
focal = c('L1', 'L2', 'L3', 'L4','L5', 'L6', 'La1', 'La2', 'La3')

b <- subset(shared_allele_proportions, lineage %in% focal) %>%
  melt(measure.vars = c('low', 'lowmid', 'midhigh', 'high'), id.vars='lineage') %>%
  ggplot(aes(x=variable, y=value, fill=lineage)) + geom_histogram(stat='identity', position='dodge') +
  scale_fill_manual(values=schemata$col_scale) + theme_bw()


# Shared proportion vs count
shared_allele_proportions$high_nsites <- shared_allele_proportions$high * shared_allele_proportions$nsites

c <- ggplot(subset(shared_allele_proportions, lineage %in% focal), aes(high_nsites, high)) + geom_point() +
  theme_bw() + geom_text_repel(aes(label=lineage), nudge_x = 1, size=2) +
  xlab("Number of high frequency sites") + ylab("Proportion of high frequency sites")


cowplot::plot_grid(a,c, ncol=2, labels=c('A', 'B'), label_fontfamily = 'CMU Sans Serif', rel_widths = c(3,2))

ggsave('figures/main/figX.IS_diversity.pdf', width = 20, height = 8, units = "cm", device = cairo_pdf) 





```



## Birth and death rates
How does IS6110 activity vary across the MTBC? 
Do more copies imply more transposition? Or is this impossible to observe due to the effects of strong purifying selection?

Compare the distribution of birth and death rates among lineages! For this, assign lineage to internal nodes using the MRCA information.

```{r}
library(ggtree)
library(ggplot2)
library(dplyr)
library(viridis)
library(reshape2)
library(RColorBrewer)


branch_smry <- merge(
  read.csv('workflow/results/detettore/AS_5prime.branch_summary.tsv', sep='\t'),
  read.csv('workflow/results/detettore/AS_3prime.branch_summary.tsv', sep='\t'),
  by =c('parent', 'child')
)

# Add copy numbers and birth/death rates to tree
metadata$copy_number <- copy_numbers$CN[match(metadata$GNUMBER, copy_numbers$strain)]

# Assign lineage to internal nodes
node_lineage = data.frame()

for (i in 1:nrow(mrcas)){
  row = mrcas[i,]
  l = row$lineage
  n = row$node

  desc = get_descendants(tree, node=n)
  desc_df = data.frame(
    node = desc,
    lineage = rep(l, length(desc))
  )
  node_lineage = rbind(node_lineage, desc_df)
}

# Load tree data with CN ancestral states
treedata = read.csv('workflow/results/phylogeny/treedata.CN_ASR.tsv', sep='\t')
summary(treedata)

# Add births and deaths
treedata <- left_join(treedata, branch_smry, by=c("node" = "child"))

treedata$lineage = node_lineage$lineage[match(treedata$node, node_lineage$node)]

names(treedata)[1] <- 'parent'

# Assign node type terminal/internal
treedata$node_type = rep('internal', nrow(treedata))
treedata$node_type[which(treedata$isTip)] <- 'terminal'
treedata$node_type <- as.factor(treedata$node_type)

summary(treedata)
```


Correlation between 5' and 3' estimates
```{r}
# Correlation between 5' and 3' estimates
a <- ggplot(branch_smry, aes(births.x, births.y)) + geom_point() + geom_smooth() + theme_bw() + 
  xlab('Births inferred from 5-prime') + ylab('Births inferred from 3-prime')
b <- ggplot(branch_smry, aes(deaths.x, deaths.y)) + geom_point() + geom_smooth() + theme_bw() +
  xlab('Deaths inferred from 5-prime') + ylab('Deaths inferred from 3-prime')

cowplot::plot_grid(a,b,ncol=2,labels=c('A', 'B'), label_fontfamily = 'CMU Sans Serif')

cor(branch_smry$births.x,branch_smry$births.y)
cor(branch_smry$deaths.x,branch_smry$deaths.y)

```

Estimate birth and death rates!
```{r}
# Remove 0-length branches
treedata_zero <- subset(treedata, branch.length==0)
sum(treedata_zero$births.x, na.rm = TRUE)

# Ignore 0-length branches, they only have few births
treedata_noZero <- subset(treedata, branch.length>0)
summary(treedata_noZero)

# Get birth and death rates
treedata_noZero$birthrate_genome = treedata_noZero$births.x/treedata_noZero$branch.length
treedata_noZero$deathrate_genome = treedata_noZero$deaths.x/treedata_noZero$branch.length

# Divide by genome size to get rate per position
genome_size = 4435783
treedata_noZero$birthrate = treedata_noZero$birthrate_genome / genome_size
treedata_noZero$deathrate = treedata_noZero$deathrate_genome / genome_size

summary(treedata_noZero$birthrate)
summary(treedata_noZero$deathrate)

```

Basic summary information
```{r}
# How many births and deaths in total?
sum(treedata_noZero$births.x)
sum(treedata_noZero$births.y)
sum(treedata_noZero$deaths.x)
sum(treedata_noZero$deaths.y)

# How many branches with 0 births/deaths
length(which(treedata$births.x==0 & treedata$isTip)) / nrow(treedata[which(treedata$isTip),])
length(which(treedata$births.x==0 & !treedata$isTip)) / nrow(treedata[-which(treedata$isTip),])

# Distributions of birth and death rates
ggplot(treedata_noZero, aes(x=isTip, y=birthrate)) + geom_boxplot()

# Terminal vs. internal branches
by(treedata_noZero$birthrate, treedata_noZero$isTip, summary)

# OVerall mean birth and death rate
summary(treedata_noZero)

# Number of births terminal vs. internal branches
by(treedata_noZero$births.x, treedata_noZero$isTip, sum)
by(treedata_noZero$births.y, treedata_noZero$isTip, sum)
by(treedata_noZero$deaths.x, treedata_noZero$isTip, sum)
by(treedata_noZero$deaths.y, treedata_noZero$isTip, sum)

by(treedata_noZero$birthrate, treedata_noZero$isTip, summary)

```


## Model birth and death rates

```{r}
library(glmmTMB)

head(treedata)
summary(treedata$birthrate)





fit_raw <- glmmTMB(birthrate ~ copy_number + node_type + lineage_mod + (1 | lineage_mod),
                   data = treedata, family = tweedie(link = "log"))

summary(fit_raw)
# Interpret: exp(coef) = multiplicative change per +1 copy
exp(coef(summary(fit_raw))$cond[, "Estimate"])


newdata <- expand.grid(
  copy_number = seq(min(treedata$copy_number), max(treedata$copy_number), length.out = 100),
  node_type = "internal",      # choose reference or loop over types
  lineage_mod = levels(treedata$lineage)[1]
)
pred <- predict(fit_raw, newdata = newdata, type = "response", se.fit = TRUE)
newdata$pred <- pred$fit
newdata$lower <- pred$fit - 1.96 * pred$se.fit
newdata$upper <- pred$fit + 1.96 * pred$se.fit

ggplot(newdata, aes(x=copy_number, y=pred)) + geom_point() + geom_smooth() +
  theme_bw() + xlab('Copy number') + ylab('Predicted birth rate') 

ggplot(newdata, aes(x=copy_number, y=pred, ymin=lower, ymax=upper)) + geom_errorbar() + geom_smooth() +
  theme_bw() + xlab('Copy number') + ylab('Predicted birth rate')










# Tweedie (compound Poisson-gamma) is good for continuous positives with point-mass zeros
fit_tweedie <- glmmTMB(birthrate ~ copy_number_z + node_type + lineage + (1 | lineage),
                       data = treedata, family = tweedie(link = "log"))


summary(fit_tweedie)

coefs <- summary(fit_tweedie)$coefficients$cond

# exponentiate fixed effects with CIs
exp_effects <- exp(confint(fit_tweedie, parm = "beta_"))
exp_effects






# Extract lineage coefficients (on link scale), get CIs with confint
coefs <- summary(fit_tweedie)$coefficients$cond
confint(fit_tweedie, parm = "beta_", level = 0.95)  # may take time

# Marginal predictions per lineage (population-level)
newdata <- data.frame(copy_number_z = 0, node_type = levels(df$node_type)[1],
                      lineage = levels(df$lineage))
preds <- predict(fit_tweedie, newdata = newdata, type = "response", se.fit = TRUE)
preds

```



```{r}
# Full script: brms hurdle-lognormal with lineage fixed, sublineage random
# Assumptions: you have a data.frame `d` with columns:
#   rate       : numeric (>= 0)
#   TEcopies   : numeric
#   node_type  : factor or character ("terminal"/"internal")
#   lineage    : factor (fixed effect you want)
#   sublineage : factor (nested within lineage)  <-- RANDOM effect
#   node       : optional unique id per node (optional random effect)
#
# Required packages
library(brms)       # install.packages("brms")
library(cmdstanr)   # optional but recommended; install and set up cmdstan first
library(dplyr)
library(tibble)
library(coda)       # for HPDinterval if desired

# -----------------------
# 0. Preprocess & checks
# -----------------------
# Convert to factors and center/scale numeric predictors
treedata = read.csv('workflow/results/phylogeny/treedata.CN_ASR.tsv', sep='\t')
# Ignore 0-length branches, they only have few births
treedata <- subset(treedata, branch.length>0)
summary(treedata)

# Split L1 into 4 sublineages
table(subset(treedata, lineage=='L1')$sublineage)
treedata$lineage_mod <- treedata$lineage
treedata$lineage_mod[which(treedata$lineage=='L1')] <- treedata$sublineage[which(treedata$lineage=='L1')]

# Remove lineages with few observations
table(treedata$lineage_mod)
remove = which(table(treedata$lineage_mod) < 100)
treedata <- subset(treedata, !(lineage_mod %in% remove)) %>% droplevels()

# Assign node type
treedata$node_type <- rep('internal', nrow(treedata))
treedata$node_type[which(treedata$isTip)] <- 'terminal'


# Add birth rate
genome_size = 4435783
treedata$birthrate_genome = treedata$births.x/treedata$branch.length
treedata$deathrate_genome = treedata$deaths.x/treedata$branch.length

treedata$birthrate = treedata$birthrate_genome / genome_size
treedata$deathrate = treedata$deathrate_genome / genome_size
# Divide by genome size to get rate per position



treedata <- treedata %>%
  mutate(
    lineage    = as.factor(lineage_mod),
    node_type  = as.factor(node_type),
    node       = as.factor(node)
  )

# scale / center TEcopies for interpretability
treedata$copy_number_z <- scale(treedata$copy_number)[,1]


# -----------------------
# 1. Model specification
# -----------------------
# We'll fit a single bf with separate formulas for the conditional (mu) and the hurdle (hu).
# lineage is a fixed effect in both parts; sublineage is a random intercept.
#
# If you don't want node-level random effect, remove "(1 | node)" lines.
#
brms_formula <- bf(
  birthrate ~ copy_number_z + node_type + lineage_mod + (1 | lineage_mod),
  hu   ~ copy_number_z + node_type + lineage_mod + (1 | lineage_mod)
)

# Use hurdle_lognormal family (positive values on log scale; zeros handled by hurdle)
family_choice <- hurdle_lognormal()

# Priors (weakly informative; adjust if you have prior knowledge)
priors <- c(
  prior(normal(0, 2), class = "b"),          # fixed effects
  prior(student_t(3, 0, 2.5), class = "sd"), # group-level sds
  prior(student_t(3, 0, 2.5), class = "sigma") # residual for positive lognormal
)

# brms control: use cmdstanr backend if available
#options(brms.backend = "cmdstanr")  # set globally; remove if using rstan

# -----------------------
# 2. Fit the model
# -----------------------
fit <- brm(
  formula = brms_formula,
  data = treedata,
  family = family_choice,
  prior = priors,
  chains = 2, cores = 4, iter = 500, warmup = 100,
  control = list(adapt_delta = 0.95, max_treedepth = 15),
  seed = 2025
)

# Save fit object
# saveRDS(fit, "brms_hurdle_lineage_fixed_sublineage_random.rds")



# -----------------------
# 3. Basic diagnostics
# -----------------------
print(fit)          # summary including Rhat and ESS
plot(fit)           # traceplots + densities
pp_check(fit)       # posterior predictive check (default)
loo_fit <- try(loo(fit), silent = TRUE)  # approximate LOO; may be slow
if(!inherits(loo_fit, "try-error")) print(loo_fit)

# Check for divergences / warnings
sampler_stats <- sampler_params(fit)
# To find divergent transitions (cmdstanr):
if("divergent__" %in% names(sampler_stats[[1]])) {
  n_div <- sum(sapply(sampler_stats, function(x) sum(x[,"divergent__"])))
  message("Total divergent transitions: ", n_div)
}

# -----------------------
# 4. Extract lineage effect sizes (posterior)
# -----------------------
# Fixed effects: conditional (mu) will have names "b_mu_*", hurdle part has "b_hu_*"
# Use fixef() for summaries and posterior_samples() for full posterior draws.

fixefs <- fixef(fit)   # matrix: Estimate, Est.Error, Q2.5, Q97.5
# Rows include:
#   b_mu_Intercept, b_mu_TEcopies_z, b_mu_node_typeX, b_mu_lineageL3, ...
#   b_hu_Intercept, b_hu_TEcopies_z, b_hu_node_typeX, b_hu_lineageL3, ...
print(fixefs)

# Get posterior draws for lineage coefficients:
post <- posterior_samples(fit)  # data.frame-like of posterior draws

# Identify lineage coefficient names programmatically
mu_lineage_pars <- grep("^b_mu_lineage", colnames(post), value = TRUE)
hu_lineage_pars <- grep("^b_hu_lineage", colnames(post), value = TRUE)

# Summarize each lineage effect (median + 95% CI) for both parts
summarize_pars <- function(pars) {
  sapply(pars, function(p) {
    vec <- post[[p]]
    c(median = median(vec), q2.5 = quantile(vec, .025), q97.5 = quantile(vec, .975))
  })
}
mu_lineage_summary <- t(summarize_pars(mu_lineage_pars))
hu_lineage_summary <- t(summarize_pars(hu_lineage_pars))

mu_lineage_summary <- as.data.frame(mu_lineage_summary)
hu_lineage_summary <- as.data.frame(hu_lineage_summary)

mu_lineage_summary$parameter <- rownames(mu_lineage_summary)
hu_lineage_summary$parameter <- rownames(hu_lineage_summary)

# Clean names: extract lineage labels
mu_lineage_summary$lineage <- sub("^b_mu_lineage", "", mu_lineage_summary$parameter)
hu_lineage_summary$lineage <- sub("^b_hu_lineage", "", hu_lineage_summary$parameter)

# Print tables
cat("Conditional part (log-rate) lineage fixed effects (median and 95% CI):\n")
print(mu_lineage_summary[, c("lineage", "median", "q2.5", "q97.5")])

cat("\nHurdle part (prob>0, logit-scale) lineage fixed effects (median and 95% CI):\n")
print(hu_lineage_summary[, c("lineage", "median", "q2.5", "q97.5")])

# To get effect sizes on interpretable scales:
# - For conditional part: exponentiate median to get multiplicative factor on rate among positives
mu_lineage_summary$mult_factor <- exp(mu_lineage_summary$median)
mu_lineage_summary$mult_CI_low  <- exp(mu_lineage_summary$q2.5)
mu_lineage_summary$mult_CI_high <- exp(mu_lineage_summary$q97.5)

# - For hurdle part (logit/probit depending on link; default for hu is logit), transform to probability:
# Use inv_logit
inv_logit <- function(x) 1 / (1 + exp(-x))
hu_lineage_summary$prob_effect <- inv_logit(hu_lineage_summary$median)
hu_lineage_summary$prob_CI_low <- inv_logit(hu_lineage_summary$q2.5)
hu_lineage_summary$prob_CI_high <- inv_logit(hu_lineage_summary$q97.5)

# -----------------------
# 5. Pairwise contrasts between lineages (example)
# -----------------------
# Compute posterior distribution for difference lineage A - lineage B for conditional part
# Example: compare lineage "L3" minus "L1" (rename to your levels)
if(length(mu_lineage_pars) >= 2) {
  lvl <- gsub("^b_mu_lineage", "", mu_lineage_pars)
  # choose first two for demo:
  pA <- mu_lineage_pars[1]; pB <- mu_lineage_pars[2]
  post_diff_mu <- post[[pA]] - post[[pB]]
  cat(sprintf("\nPosterior median (mu) difference %s - %s: %f\n", lvl[1], lvl[2], median(post_diff_mu)))
  print(quantile(post_diff_mu, c(.025, .975)))
}

# -----------------------
# 6. Marginal expected rates per lineage (interpretable)
# -----------------------
# To produce per-lineage expected total rate (P(rate>0) * E[rate | rate>0]) averaged over random effects:
# Build newdata with mean TEcopies_z and a reference node_type (or iterate over multiple)
newdata <- tibble(
  TEcopies_z = mean(d$TEcopies_z, na.rm = TRUE),
  node_type  = levels(d$node_type)[1],
  lineage    = levels(d$lineage)
)

# For marginal population-level (fixed-effect only) predictions: re_formula = NA
# posterior_epred returns expected value on response scale (i.e., includes zeros)
epred_pop <- posterior_epred(fit, newdata = newdata, re_formula = NA) # matrix: draws x newdata rows

# Summarize: median and 95% CI per lineage
epred_med <- apply(epred_pop, 2, median)
epred_ci  <- apply(epred_pop, 2, quantile, c(.025, .975))

expected_rate_table <- tibble(
  lineage = newdata$lineage,
  expected_rate = epred_med,
  lower = epred_ci[1,],
  upper = epred_ci[2,]
)
print(expected_rate_table)

# If you prefer population+group-level conditional predictions (i.e., include sublineage random effects),
# supply representative sublineage values in newdata or set re_formula = NULL (but you must then include sublineage column).
# Example: conditional on each lineage but averaging over sublineages can be approximated by:
#  - draw many posterior_pred with re_formula = NULL, sampling sublineages at random from observed distribution,
#    then compute per-lineage averages. (Left as advanced exercise.)

# -----------------------
# 7. Variance partitioning / ICC for sublineage (continuous part)
# -----------------------
# VarCorr gives group-level sd estimates for both parts. For the continuous (mu) part:
vc <- VarCorr(fit)

# Access sd for sublineage on the conditional (mu) scale
# Structure: vc$sublineage$sd contains sd for each parameter; check names
print(vc$sublineage)

# For a simple approximate ICC on log-scale (conditional part):
sd_sublineage_mu <- as.numeric(vc$sublineage$sd$mu)  # sd of sublineage intercept for mu
sigma_mu <- as.numeric(vc$residual__sigma)          # residual sd for mu (log scale)
icc_sublineage <- sd_sublineage_mu^2 / (sd_sublineage_mu^2 + sigma_mu^2)
cat("\nApprox ICC (sublineage) for conditional log-rate:", icc_sublineage, "\n")

# For hurdle part, variance partitioning is on the latent (logit) scale:
sd_sublineage_hu <- as.numeric(vc$sublineage$sd$hu)
# Total latent variance approx = sd_sublineage_hu^2 + (pi^2 / 3)  # for logit link residual variance
icc_sublineage_hu <- sd_sublineage_hu^2 / (sd_sublineage_hu^2 + (pi^2 / 3))
cat("Approx ICC (sublineage) for hurdle (latent logit) :", icc_sublineage_hu, "\n")

# -----------------------
# 8. Posterior predictive checks & plots (suggested)
# -----------------------
# Use pp_check to compare observed vs predicted distribution (zeros and positives)
pp_check(fit, type = "hist")            # overall
pp_check(fit, type = "stat", stat = "prop_zero") # fraction of zeros
pp_check(fit, type = "dens_overlay")    # density overlay of positive values

# Plot lineage-level expected rates (table or ggplot)
library(ggplot2)
expected_rate_table %>%
  ggplot(aes(x = reorder(lineage, expected_rate), y = expected_rate)) +
  geom_point() +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2) +
  coord_flip() +
  labs(y = "Expected TE birth rate (median)", x = "Lineage (population-level)")

# -----------------------
# End of script
# -----------------------


```




## Birth rate vs. copy number
```{r}
# Does birth rate depend on copy number?
y_raw <- treedata$births.x / treedata$branch.length
y_pos <- y_raw[is.finite(y_raw) & !is.na(y_raw) & (y_raw > 0)]
log_min <- floor(log10(min(y_pos)))
log_max <- ceiling(log10(max(y_pos)))
# major breaks: 10^n
major_breaks <- 10^(log_min:log_max)
# minor breaks: 1..9 * 10^n (this gives the conventional minor gridlines)
minor_breaks <- as.vector(outer(1:9, 10^(log_min:log_max), `*`))
# keep only minor breaks within the data range
minor_breaks <- minor_breaks[minor_breaks >= min(y_pos) & minor_breaks <= max(y_pos)]
  
x_levels <- levels(as.factor(treedata$copy_number))

treedata_no7 <- subset(treedata, lineage!= 'L7')

ggplot(na.omit(treedata_no7), aes(as.factor(copy_number), births.x/branch.length)) + 
  geom_boxplot(outlier.size = 0.3) + facet_wrap(~lineage) +
  theme_minimal() +  xlab("Copy number") + ylab("IS6110 birth rate") + 
  scale_x_discrete(
    labels = ifelse(seq_along(x_levels) %% 5 == 1, x_levels, "")) + 
  scale_y_log10(
    breaks = major_breaks,
      minor_breaks = minor_breaks,
  )
```


## Birth rate vs. lineage


## fragmenti
```{r}
# Net births vs. branch length
a <- ggplot(treedata, aes(x=branch.length, y=births.x-deaths.x)) + 
  geom_point(alpha=0.1) + geom_smooth() +
  xlab('Branch length') + ylab('IS6110 births - deaths') +
  theme_bw(base_family = "CMU Sans Serif",base_size = 12)

# Distribution of birth and death rates
b <- melt(subset(treedata), id.vars=c('node'), measure.vars=c('birthrate', 'deathrate')) %>%
  ggplot(aes(x=variable, y=value)) + geom_boxplot() + scale_y_log10() +
  theme_bw(base_family = "CMU Sans Serif",base_size = 12) +
  xlab("") + ylab("Rate")


cowplot::plot_grid(a,b, ncol=2, labels = c('A', 'B'), label_fontfamily='CMU Sans Serif')


# Do births and deaths scale with branch length?
melt(subset(treedata), id.vars=c('node', 'branch.length'), measure.vars=c('births', 'deaths')) %>% ggplot(aes(branch.length, value)) + 
  geom_point(alpha=0.1) + 
  facet_wrap(~variable) +
  xlab('Branch length') + ylab('Births/deaths of IS6110') +
  theme_bw(base_family = "CMU Sans Serif",base_size = 12)


# Plot birth vs death rates
ggplot(treedata_noZero, aes(deathrate, birthrate)) + 
  geom_point() + geom_smooth() +
  theme_bw() 

# Distribution of branch lengths
ggplot(treedata, aes(branch.length)) + geom_density()



# Add distance to the root?? -> highest rates in terminal branches
distance_to_root <- function(tree, nodes) {
  root_node <- Ntip(tree) + 1
  sapply(nodes, function(node) {
    path <- nodepath(tree, root_node, node)
    sum(tree$edge.length[tree$edge[, 2] %in% path[-1]])
  })
}

treedata$dist2root <- distance_to_root(tree, treedata$node)
p + geom_tree(data=treedata, aes(col=dist2root)) + scale_color_viridis(option='turbo')  # sanity check

# Births vs distance to root
ggplot(subset(treedata), aes(x=dist2root, y=births.x/branch.length)) + geom_point(alpha=0.2) + geom_smooth()+ theme_bw() + facet_wrap(~lineage)


# Compare birth and death rates among lineages!
treedata$birthrate[which(treedata$birthrate == 0)] <- 1
treedata$deathrate[which(treedata$deathrate == 0)] <- 1

melt(treedata_noZero, measure.vars = c('birthrate', 'deathrate'),id.vars=c('parent','node', 'lineage')) %>% na.omit() %>%
  ggplot(aes(lineage, value, fill=variable)) + geom_boxplot() + 
  theme_bw() + scale_fill_brewer(palette='Dark2')  +
  xlab('Lineage') + ylab('Rate (log10)') +
  scale_y_log10()

ggsave('presentations/labmeeting.21-10-2025/figures/rates_per_lineage.pdf', width = 18, height = 10, units = "cm", device = cairo_pdf) 




# How many branches with 0 events?
table(treedata$births)
table(treedata$deaths)




# What if we add copy number?
melt(subset(treedata, births>0 | deaths>0), id.vars=c('node', 'branch.length', 'copy_number'), measure.vars=c('deaths', 'births')) %>% ggplot(aes(branch.length, value)) + 
  geom_jitter(aes(col=copy_number),alpha=0.1) + geom_smooth() + 
  facet_wrap(~variable) +
  xlab('Branch length') + ylab('Births/deaths of IS6110') +
  theme_bw() 



ggsave('presentations/labmeeting.21-10-2025/figures/birthdeath_vs_branchlength.pdf', width = 18, height = 8, units = "cm", device = cairo_pdf) 

m0 <- lm(births~branch.length, data=treedata)
summary(m0)





# Mean birth and death rate for simulation
birthrates  = treedata$birthrate[!is.na(treedata$birthrate) & !is.infinite(treedata$birthrate) & !is.nan(treedata$birthrate)]
summary(birthrates)
summary(birthrates[which(birthrates>0)])
quantile(birthrates[which(birthrates>0)],c(0.1,0.9))


deathrates  = treedata$deathrate[!is.na(treedata$deathrate) & !is.infinite(treedata$deathrate) & !is.nan(treedata$deathrate)]
summary(deathrates)
summary(deathrates[which(deathrates>0)])
quantile(deathrates[which(deathrates>0)],c(0.1,0.9))



# Distribution of birth and death rates, add color legend
ggplot(treedata, ) + geom_density(aes(birthrate)) + geom_density(aes(deathrate), col='red') + 
  scale_x_log10() + theme_minimal() + xlab('Rate (events per unit branch length)') + ylab('Density')

# Rate on tree (not really visible...)
p$data <- treedata
p + geom_tree(aes(col=birthrate),size=0.5, alpha=0.5) +  
  scale_color_viridis(option = 'turbo') +
  labs(col = "IS6110 birth rate")
```




What information can we get from the complete matrix? 

## 5' vs 3' similarity
```{r}
library(vegan)

mat1 = as.matrix(out$matrix5)
mat2 = as.matrix(out$matrix3)

dim(mat1)
dim(mat2)

row_jaccard <- function(a, b) {
  inter <- rowSums(a & b)
  union <- rowSums(a | b)
  mean(inter / union, na.rm = TRUE)
}

# Assume mat1 and mat2 are binary matrices with same rows but different columns
# We need to align them by collapsing columns into sets
mat1_sets <- lapply(1:nrow(mat1), function(i) which(mat1[i, ] == 1))
mat2_sets <- lapply(1:nrow(mat2), function(i) which(mat2[i, ] == 1))

# Compute average row-level Jaccard
row_jaccard_score <- mean(mapply(function(s1, s2) {
  inter <- length(intersect(s1, s2))
  union <- length(union(s1, s2))
  if (union == 0) return(NA) else inter / union
}, mat1_sets, mat2_sets), na.rm = TRUE)

row_jaccard_score

```





## Site-summaries
```{r}

library(ggplot2)

sitesmry <- list(
  five = read.csv("workflow/results/detettore/AS_5prime.site_summary.tsv", sep='\t'),
  three = read.csv("workflow/results/detettore/AS_3prime.site_summary.tsv", sep='\t')
)



ggplot(sitesmry$five, aes(x=births,y=deaths)) + 
  geom_point(alpha=0.2) + 
  theme(base_family = "CMU Sans Serif", base_line_size = 10)


```






## Conservation index: between lineages
```{r}
library(dplyr)

# --- Example data ---
set.seed(42)
n_strains <- 100
n_sites <- 500
mat <- matrix(rbinom(n_strains * n_sites, 1, 0.2),
              nrow = n_strains, ncol = n_sites)
rownames(mat) <- paste0("strain", 1:n_strains)

# Assign strains to two lineages
lineageA <- rownames(mat)[1:50]
lineageB <- rownames(mat)[51:100]

# --- Functions ---
# Compute conservation index for a subset of strains
compute_conservation_index <- function(mat, strain_subset) {
  submat <- mat[strain_subset, , drop = FALSE]
  freqs <- colSums(submat)
  p <- freqs / length(strain_subset)
  p <- pmin(pmax(p, 1e-10), 1 - 1e-10) # avoid log(0)
  entropies <- -p * log2(p) - (1 - p) * log2(1 - p)
  mean_entropy <- mean(entropies)
  CI <- 1 - mean_entropy
  return(CI)
}

# Bootstrap test for CI difference
bootstrap_ci_diff <- function(mat, subset1, subset2, n_boot = 1000) {
  # Observed difference
  ci1 <- compute_conservation_index(mat, subset1)
  ci2 <- compute_conservation_index(mat, subset2)
  obs_diff <- ci1 - ci2
  
  # Bootstrap differences
  boot_diffs <- replicate(n_boot, {
    cols <- sample(ncol(mat), replace = TRUE) # resample sites
    ci1b <- compute_conservation_index(mat[, cols, drop = FALSE], subset1)
    ci2b <- compute_conservation_index(mat[, cols, drop = FALSE], subset2)
    ci1b - ci2b
  })
  
  # Two-sided p-value
  p_val <- mean(abs(boot_diffs) >= abs(obs_diff))
  
  list(
    observed_diff = obs_diff,
    p_value = p_val,
    boot_distribution = boot_diffs
  )
}

# --- Run bootstrap test ---
res <- bootstrap_ci_diff(mat, lineageA, lineageB, n_boot = 5000)

cat("Observed CI difference:", res$observed_diff, "\n")
cat("Bootstrap p-value:", res$p_value, "\n")

# Optional: visualize bootstrap distribution
hist(res$boot_distribution, breaks = 50, main = "Bootstrap CI Differences",
     xlab = "CI difference (LineageA - LineageB)")
abline(v = res$observed_diff, col = "red", lwd = 2)

```

## Conservation index: along the genome!
```{r}
library(dplyr)

# --- Example simulated data ---
set.seed(42)
n_strains <- 100
n_sites <- 1000
# Simulate binary presence-absence
mat <- matrix(rbinom(n_strains * n_sites, 1, 0.2),
              nrow = n_strains, ncol = n_sites)
rownames(mat) <- paste0("strain", 1:n_strains)

# Assign each site a genomic position (sorted)
positions <- sort(sample(1:5e6, n_sites))  # positions in bp

# Assign two lineages
lineageA <- rownames(mat)[1:50]
lineageB <- rownames(mat)[51:100]

# --- Functions ---
compute_conservation_index <- function(mat, strain_subset) {
  submat <- mat[strain_subset, , drop = FALSE]
  freqs <- colSums(submat)
  p <- freqs / length(strain_subset)
  p <- pmin(pmax(p, 1e-10), 1 - 1e-10)
  entropies <- -p * log2(p) - (1 - p) * log2(1 - p)
  CI <- 1 - mean(entropies)
  return(CI)
}

# Sliding window bootstrap CI difference
sliding_ci_diff <- function(mat, positions, subset1, subset2,
                            window_size, step_size, n_boot = 500) {
  stopifnot(length(positions) == ncol(mat))
  
  windows <- seq(min(positions), max(positions) - window_size, by = step_size)
  results <- data.frame(
    window_start = windows,
    window_end = windows + window_size,
    observed_diff = NA_real_,
    p_value = NA_real_
  )
  
  for (i in seq_along(windows)) {
    start_pos <- windows[i]
    end_pos <- start_pos + window_size
    cols <- which(positions >= start_pos & positions < end_pos)
    if (length(cols) < 5) next  # skip if too few sites
    
    ci1 <- compute_conservation_index(mat[, cols, drop = FALSE], subset1)
    ci2 <- compute_conservation_index(mat[, cols, drop = FALSE], subset2)
    obs_diff <- ci1 - ci2
    
    # Bootstrap within window
    boot_diffs <- replicate(n_boot, {
      boot_cols <- sample(cols, replace = TRUE)
      ci1b <- compute_conservation_index(mat[, boot_cols, drop = FALSE], subset1)
      ci2b <- compute_conservation_index(mat[, boot_cols, drop = FALSE], subset2)
      ci1b - ci2b
    })
    p_val <- mean(abs(boot_diffs) >= abs(obs_diff))
    
    results$observed_diff[i] <- obs_diff
    results$p_value[i] <- p_val
  }
  
  return(results)
}

# --- Run sliding window ---
window_size <- 50000   # 50 kb windows
step_size <- 10000     # slide every 10 kb
res <- sliding_ci_diff(mat, positions, lineageA, lineageB,
                       window_size, step_size, n_boot = 500)

# --- Plot results ---
library(ggplot2)

ggplot(res, aes(x = window_start + window_size/2, y = observed_diff)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(aes(color = p_value < 0.05)) +
  scale_color_manual(values = c("black", "red")) +
  labs(x = "Genomic position (bp)",
       y = "CI difference (A - B)",
       color = "Significant (p < 0.05)",
       title = "Conservation index differences along genome") +
  theme_minimal()

```


## Ancestral states through time

Loop through tree and 


```{r}
library(ape)

branch_smry <- list(
  five = read.csv('workflow/results/detettore/AS_5prime.branch_summary.tsv', sep='\t'),
  three = read.csv('workflow/results/detettore/AS_3prime.branch_summary.tsv', sep='\t')
)

head(branch_smry$five)

root_node <- rootnode(out$tree)
sum(out$tree$edge.length[out$tree$edge[, 2] %in% c(, root_node)])



out$tree$
is.rooted(out$tree)


root_node <- which(!out$tree$edge[, 2] %in% out$tree$edge[, 1])


distance <- sum(tree$edge.length[tree$edge[, 2] %in% c("Tip1", mrca)])


# Function to traverse the tree and record state changes
traverse_tree <- function(node, tree, ancestral_states, state_changes) {
  # Get the state of the current node
  current_state <- ancestral_states[ancestral_states$node == node, "state"]
  
  # If the node has children, traverse them
  if (length(tree$edge[tree$edge[,1] == node, 2]) > 0) {
    for (child in tree$edge[tree$edge[,1] == node, 2]) {
      child_state <- ancestral_states[ancestral_states$node == child, "state"]
      # Record the state change if it exists
      if (!is.na(child_state) && current_state != child_state) {
        state_changes <- rbind(state_changes, data.frame(from = current_state, to = child_state, node = child))
      }
      # Recursively traverse the child
      state_changes <- traverse_tree(child, tree, ancestral_states, state_changes)
    }
  }
  return(state_changes)
}

# Initialize an empty data frame to record state changes
state_changes <- data.frame(from = character(), to = character(), node = integer(), stringsAsFactors = FALSE)

# Start traversing from the root (node 1)
state_changes <- traverse_tree(1, tree, ancestral_states, state_changes)
```



- plot state changes through time
- ancestral reconstruction: number of unique insertion events
- birth and death rates

```{r}
library(ape)
library(phangorn)
library(castor)
library(dplyr)
library(ggplot2)
library(reshape2)
library(magrittr)
library(ggtree)
library(treeio)

source('workflow/notebooks/utils.R')


# Clusters of unique anchors ---------------------------------------
ggplot(out$matrix_meta, aes(cluster)) + 
  geom_histogram(stat='count') +
  theme_bw()
  


# Analyze birth and death rates ------------------------------------
"
Reconstruct ancestral states for IS presence/absence. Produce two summaries,
per branch and per site, and add state transitions to the ggtree object.

"

branch_smry <- list(
  five = read.csv('workflow/results/10k/detettore/AS_5prime.branch_summary.tsv', sep='\t'),
  three = read.csv('workflow/results/10k/detettore/AS_3prime.branch_summary.tsv', sep='\t')
)

site_smry <- list(
  five = read.csv('workflow/results/10k/detettore/AS_5prime.site_summary.tsv', sep='\t'),
  three = read.csv('workflow/results/10k/detettore/AS_3prime.site_summary.tsv', sep='\t')
)

anc_states = list(
  five = read.csv('workflow/results/10k/detettore/AS_5prime.ancestral_likelihoods.tsv', sep='\t'),
  three = read.csv('workflow/results/10k/detettore/AS_3prime.ancestral_likelihoods.tsv', sep='\t')
)


# Distribution of births and deaths
melt(branch_smry, measure.vars=c('births', 'deaths')) %>%
  ggplot() + 
  geom_density(aes(value, col=variable), alpha=0.5)+ 
  theme_bw() + facet_wrap(~L1) + scale_x_log10()

melt(site_smry, measure.vars=c('births', 'deaths')) %>%
  ggplot() + 
  geom_density(aes(value, col=variable), alpha=0.5)+ 
  theme_bw() + facet_wrap(~L1) + scale_x_log10()



# Create tree containing all the ancestral states?
p <- ggtree(out$tree, size=0.0001) %<+% out$metadata
td <- as.treedata(p)

for (i in 2:ncol(anc_states$five)){
  site_id = names(anc_states$five[i])
  
  # skip singletons
  if (sum(anc_states$five[i])<1000) next
  
  td@data[[site_id]] <- anc_states$five[[site_id]][match(td@data$node, anc_states$five$node)]
}


td@data$node_id <- td@data$node
td@data$node_id[1:10000] <- td@phylo$tip.label[1:10000]

write.table(td@data, file = '~/Desktop/random_explorations/taxonium/md.tsv' ,quote=F, row.names=F, sep='\t')
write.beast.newick(td, '~/Desktop/random_explorations/taxonium/mtbc.is6110_anc_states.newick') 
write.beast(td, '~/Desktop/random_explorations/taxonium/mtbc.is6110_anc_states.nex') 





# Cut-n-paste ------------------------------
" Does absence at some sites correlate with presence at others? This
could indicate cut-and-paste transposition or methodological issues

"
library(Matrix)
matrix5 <- Matrix(as.matrix(out$matrix5), sparse = TRUE)

# 2. Compute phi correlations
phi_matrix <- binary_cor(matrix5)

# 3. Select pairs with strong negative correlation
neg_pairs <- which(phi_matrix < -0.2, arr.ind = TRUE)
pos_pairs <- which(phi_matrix > 0.2, arr.ind = TRUE)

# 4. Run Fisher's exact test on selected pairs
fisher_results <- apply(neg_pairs, 1, function(idx) {
  fisher.test(table(X[, idx[1]], X[, idx[2]]))
})








# Visualize birth/death on tree


# Plot distributions of birth and death rates
melt(
  as_branches, 
  id.vars = c('edge_id', 'lineage'), 
  measure.vars = c('birth_rate', 'death_rate')) %>%
  ggplot(aes(y=value, x=lineage, fill=variable)) + 
  geom_boxplot(position='dodge') + scale_y_log10()



```

## Sanity check: plot rates on tree
```{r}
# Test for sublineage
metadata_l = subset(out$metadata, LINEAGE_x == 'L6')
table(metadata_l$SUBLINEAGE_x)

lineage = 'L6.2.2'
lineage_sub <- subset_tree_matrix(out$tree, out$matrix5,out$metadata, lineage, 'sublineage')

tree_l <- lineage_sub[[1]]
matrix_l <- lineage_sub[[2]] %>% as.matrix()
  
# Ensure order matches tree tips
matrix_l <- matrix_l[tree_l$tip.label, ]

# Get birth and death rates
as_l <- is_birth_death(tree_l, matrix_l)



# Plot
p <- ggtree(tree_l) %<+% metadata_l
treedata <- p$data
treedata <- left_join(treedata, as_l$branch_summary, by=c("node" = "child"))
p <- p + 
  geom_nodelab(data=treedata,aes(label=insertions), col='red', nudge_x = 0.00001, nudge_y = 0.00001) + 
  geom_tiplab(data=treedata,aes(label=insertions), col='red', nudge_x = 0.00001, nudge_y = 0.00001)

gheatmap(p, matrix_l, colnames = FALSE) 
```


# E: chromosomal distribution

Fig X. Macroscale insertion bias. A) Genomic distribution of IS6110 insertion sites. To improve visibility, maximally 500 genomes were included per lineage. The order of the strains is the same as in the phylogeny in Fig. 2, such that vertical bands 

B) Density of insertion sites increases with distance to the origin of replication (oriC). C) Motif analysis. D) Genic versus intergenic insertion sites. 


## Circos plot: macroscale insertion bias
- add positions/density of DeJesus' essential genes

```{r}
library(ape)
library(magrittr)
library(ggplot2)
library(dplyr)
library(Cairo)
library(circlize)
library(ComplexHeatmap)

col_schemes <- load_color_schemes()

custom_order = c('L1', 'L2', 'L3', 'L4', 'L5', 'L6', 'L7', 'L8', 'L9', 'L10','La1', 'La2', 'La3')

# Sort according to the phylogenetic tree!
refins_ordered <- subset(refins, lineage %in% custom_order) %>% mutate(column_name = factor(strain, levels = rev(tree$tip.label))) %>% arrange(column_name)

head(refins_ordered)


# Reduce number of strains: Get n per lineage
n = 500
strains_reduced = c()
for (lineage in rev(custom_order)){
  strains_lineage = unique(refins_ordered$strain[refins_ordered$lineage==lineage])
  if (length(strains_lineage) == 0){next}
  if (length(strains_lineage) <= n){strains_reduced = c(strains_reduced, strains_lineage)}
  else {
    strains_sample = sample(strains_lineage, n)
    strains_reduced = c(strains_reduced, strains_sample)
  }
}
length(strains_reduced)

# Create list of bed files, per strain, for circos
bed_list = list()
colori = c()

refins_reduced <- subset(refins_ordered, strain %in% strains_reduced)
head(refins_reduced)

strains_ordered <- unique(refins_reduced$strain)

for (i in 1:length(strains_ordered)){
  
  df <- subset(refins_reduced, strain==strains_ordered[i])
  bed <- data.frame(
    'chr' = df$chromosome,
    'start' = df$position,
    'end' = df$position+1,
    'lineage' = df$lineage
  )
  
  bed_list[[i]] <- bed
  
  # Now the lineage color
  lineage =df$lineage[1]
  col = col_schemes$col_scale_black[lineage]
  colori <- c(colori, col)
}

length(bed_list)
length(colori)

genome = data.frame(chr  = "MTBC0", start = 1,end   = 4435783) 


# Define color legend
lineages = names(col_schemes$col_scale_black)
lineages = lineages[lineages %in% unique(refins$lineage)]
colorrs = col_schemes$col_scale[lineages]

lgd = Legend(
  at = lineages, type = "lines", 
  legend_gp = gpar(col = colorrs, lwd = 2), 
  title_gp = gpar(fontfamily = "CMU Sans Serif", fontsize = 10),
  labels_gp = gpar(fontfamily = "CMU Sans Serif", fontsize = 10),
  title_position = "topleft", 
  title = "Lineage", break_dist = 2)


# Plot insertions!
options(fig.width = 15, fig.height = 15, repr.plot.width = 15, repr.plot.height = 15)
par(bg = "white")

CairoPDF("figures/main/fig4a.circos.pdf",  width = 10, height = 10)

circlize_plot(genome, bed_list, colori)
draw(lgd, x = unit(10, "mm"), y = unit(20, "mm"), just = c("left", "bottom"))

dev.off()


```


## Genic vs. intergenic
Overrepresentation in intergenic as evidence for purifying selection. 
Also, a higher proportion of genic insertions are expected in singleton insertions. 

Still the weird thing that there are many identical positions with different strands...


Figure: 

```{r}
library(ggplot2)

# Create df with unique reference insertions 
refins_unique = data.frame()

for (ru in unique(refins$unique_id)){
  refins_sub = subset(refins, unique_id == ru)

  df = data.frame(
    unique_id = ru,
    position = refins_sub[1,]$position,
    gene = refins_sub[1,]$gene,
    rv_ortholog = refins_sub[1,]$rv_ortholog,
    context = refins_sub[1,]$context,
    nstrains = nrow(refins_sub),
    freqcat = ifelse(nrow(refins_sub) > 1, 'shared', 'singleton')
  )
  refins_unique = rbind(refins_unique, df)
}



# Are intergenic insertions overrepresented?
gff_path <- 'data/MTBC0/MTBC0v1.1_PGAP_annot.gff'

genic_observed <- unname(table(refins_unique$context)['genic'])
intergenic_observed <- unname(table(refins_unique$context)['intergenic'])
chromosome_length = 4435783

fisher_test(genic_observed, intergenic_observed,chromosome_length, gff_path)


# What about singleton vs. shared elements?
ggplot(refins_unique, aes(context)) + geom_histogram(stat='count') + 
  facet_wrap(~freqcat) + 
  xlab("") + ylab("Count") +
  theme_bw(base_family = "CMU Sans Serif", base_size=14)


# Fisher's exact test
refins_singletons <- subset(refins_unique, freqcat=='singleton')
refins_shared <- subset(refins_unique, freqcat=='shared')

fisher_test(
  unname(table(refins_singletons$context)['genic']),
  unname(table(refins_singletons$context)['intergenic']), 
  chromosome_length,
  gff_path
)

ratio_singletons = unname(table(refins_singletons$context)['intergenic']) /(unname(table(refins_singletons$context)['genic']) + unname(table(refins_singletons$context)['intergenic']))


fisher_test(
  unname(table(refins_shared$context)['genic']),
  unname(table(refins_shared$context)['intergenic']),
   chromosome_length,
  gff_path
)

ratio_shared = unname(table(refins_shared$context)['intergenic']) / (unname(table(refins_shared$context)['genic']) + unname(table(refins_shared$context)['intergenic']))

```


## Distance effect
```{r}
library(ggplot2)

bin_sizes = c(50000, 100000, 200000, 500000)
chrom_length = 4435783
positions <- unique(refins$position)

dist_from_ori = c()

for (pos in positions){
  if (pos < chrom_length/2){
    dist = pos
  }
  else {
    dist = chrom_length-pos
  }
  dist_from_ori = c(dist_from_ori, dist)
}


bin_size = 100000
n_breaks = as.integer(chrom_length/bin_size, 0)
bins <- hist(dist_from_ori, breaks=n_breaks, plot = F)

bins_df <- data.frame(
  'mids' = bins$mids,
  'counts' = bins$counts,
  'density' = bins$density
)

cor.test(bins_df$mids, bins_df$counts, method = 'pearson')


fig4b <- ggplot(bins_df, aes(mids/1000, counts)) + 
  geom_point() + geom_smooth() +
  theme_minimal(base_family = "CMU Sans Serif") +
  xlab("Distance form oriC (kb)") + ylab("Nr. insertion sites/100k bin")

fig4b  


ggplot(data.frame(dists=dist_from_ori), aes(x = dists/1000)) +
  geom_histogram(aes(y = ..density..), bins = 50, fill = "lightgray", color = NA) +
  geom_density(color = "red", size = 1) +
  #geom_rug(sides = "b", alpha = 0.05) +
  labs(
    x = "Distance from oriC (kb)",
    y = "Density"
  ) +
  theme_minimal(base_family = "CMU Sans Serif")

fig4b


```

Combine panels
```{r}
library(cowplot)
library(magick)

img_circors <- image_read("figures/main/fig4a.circos.pdf", density = 600)
p_circos  <- ggdraw() + draw_image(img_circors)

plot_grid(p_circos, fig4b, 
          labels = c('A', 'B'),
          label_size = 10,
          nrow=1, 
          rel_widths = c(2,1), 
          rel_heights=c(2,1),
          label_fontfamily='CMU Sans Serif')

```



## Symmetry analysis
Is the distribution of IS6110 more symmetrical than expected by chance?
If yes, this might reflect large-scale structural adaptations in the chromosome
to mitigate the effects of inversions/deletions (Rocha 2004)

```{r}
library(ggplot2)

head(refins_unique)
chrom_length = 4435783
oriC_start=1349

refins_shared = subset(refins_unique, nstrains > 1)
positions = refins_shared$position
weights = refins_shared$nstrains

angles <- angles_from_positions(positions, chrom_length)
res <- permutation_test_best_axis(angles, weights, m = 36, n_axes = 360, n_perm = 100)
res

res_noweights <-  permutation_test_best_axis(angles, NA, m = 36, n_axes = 360, n_perm = 100)
res_noweights


# Convert best axis (radians) to bp:
best_bp <- res$best_axis * chrom_length / (2*pi)
best_bp

# Symmetry statistic scan
plot_axis_scan(res, chrom_length)

# Plot observed vs mirrored bin counts
plot_paired_bins(angles, weights, best_axis = res$best_axis, m = 36)

plot_circular_histogram(angles, weights = NULL, m = 72, res$best_axis, genome_length = chrom_length)

# Compare observed and permutation stats
perm_df = data.frame(permutations = res$perm_stats)
ggplot(perm_df, aes(permutations)) + 
  geom_density() +
  geom_vline(xintercept=res$best_stat, col='red', linetype='dashed') 



# What is being permuted?
# How is rotation permutation? Shouldnt this result just in a different best axis?
rot <- runif(1, 0, 2*pi)
angles_rot <- (angles + rot) %% (2*pi)
angles_df <- data.frame(angles, angles_rot)
ggplot(angles_df, aes(angles, angles_rot)) + geom_point() 

res_rot <- permutation_test_best_axis(angles_rot, weights, m = 36, n_axes = 360, n_perm = 100)
res_rot

plot_axis_scan(res_rot, chrom_length)
plot_paired_bins(angles_rot, weights, best_axis = res_rot$best_axis, m = 36)

```







### distance to oriC
No clear increase of essential genes near the oriC.

```{r}
library(ggplot2)

ggplot(genemap, aes(x=essentiality, y=dist_oriC)) + geom_boxplot()

# Model: logistic regression 
genemap$essential_bin <- ifelse(genemap$essentiality == "ES", 1, 0)
fit <- glm(essential_bin ~ dist_oriC, data = genemap, family = binomial)
summary(fit)

kruskal.test(dist_oriC ~ essentiality, data = genemap)
pairwise.wilcox.test(genemap$dist_oriC, genemap$essentiality, p.adjust.method = "BH")



# Strand bias
table(genemap$strand, genemap$essentiality)

ggplot(na.omit(genemap), aes(x=strand, fill=essentiality)) + geom_bar(stat='count')


```








### Upset plot

Input format: df with genes in first column, other columns lineages and disrupted 0/1

```{r}
library(UpSetR)
library(ggplot2)
library(vegan)
library(magrittr)
library(reshape2)



upset_df = data.frame(
  genes = ko_genes
)

# Create input data frame
knockouts$lineage <- out$metadata$LINEAGE_x[match(knockouts$strain, out$metadata$GNUMBER)]

custom_order = c('L1', 'L2', 'L3', 'L4', 'L5', 'L6', 'L7', 'L8', 'L9', 'L10','La1', 'La2', 'La3')

for (l in custom_order){
  knockout_l = subset(knockouts, lineage==l)
  upset_df[[l]] = ifelse(ko_genes %in% knockout_l$gene, 1, 0)
}

head(upset_df)




upset(upset_df,nsets=9, text.scale = 1.5, sets.x.label='Nr. k.o. genes')

# Jaccard distances
upset_m <- as.matrix(upset_df[,-1])
rownames(upset_m) <- upset_df$genes
upset_m <- t(upset_m)
dist_mat <- vegdist(upset_m, method = "jaccard", binary = TRUE)

# Nr. knockouts per lineage
rowSums(upset_m) 
```




## Motif analysis
Create a sequence logo plot for the ~10 bases adjoining all insertions.
Also try different subsets: e.g. genic vs. intergenic.
```{r}

library(ggplot2)
library(Biostrings)
library(ggseqlogo)
library(cowplot)

fasta <- list(
  five = readDNAStringSet("workflow/results/detettore/ALL_presence-absence.5prime.fasta"),
  three = readDNAStringSet("workflow/results/detettore/ALL_presence-absence.3prime.fasta")
)

get_logo <- function(seqs, n_bases, where){
  subseqs = list()
  for (i in 1:length(seqs)) {
    seq = seqs[i]
    if (where == 'end'){
      end = width(seq)
      sseq = subseq(seq, start=end-(n_bases-1), end=end)
    }
    else if (where == 'start'){
      sseq = subseq(seq, start=1, end=n_bases)
    }
    subseqs = c(subseqs, unname(as.character(sseq)))
  }
  subseqs = unlist(subseqs)
  ggplot() + geom_logo(subseqs, method='prob') + theme_logo()

}

a <- get_logo(fasta$five, 10, 'end')
b <- get_logo(fasta$three, 10, 'start')

plot_grid(a,b)

```







# Natural mutagenesis experiment

## Gene knockouts
- add frequency aspect: shared vs. singletons
```{r}
source('workflow/notebooks/utils.R')

head(refins)

knockouts = refins[which(!grepl(';', refins$gene)),]
intergenic = refins[which(grepl(';', refins$gene)),]

print(paste("Nr. of genes with insertions:", length(unique(knockouts$gene))))
print(paste("Nr. of intergenic regions with insertions:", length(unique(intergenic$gene))))

ko_genes = unique(knockouts$gene)
ko_intergenic = unique(intergenic$gene)


# How many genes/intergenic regions affected?
context_df = data.frame()
context_l = list()

for (feature in unique(refins$gene)){
  
  feature_df = subset(refins, gene == feature)
  nr_pos = length(unique(feature_df$position))
  nr_lineages = length(unique(feature_df$lineage))
  nr_strains = length(unique(feature_df$strain))
  #resistance = feature_df$resistance[1]
  
  if (grepl(';', feature)){cat = 'intergenic'}
  else {cat = 'knockout'}

  row = data.frame(feature, nr_pos, nr_lineages, nr_strains, cat)
  context_df = rbind(context_df, row)
}

head(context_df)
dim(context_df)

dim(subset(context_df, cat=='knockout'))
dim(subset(context_df, cat=='intergenic'))

```


## Accumulation analysis
Given the limited sample size, how many new genes are knocked out when considering new strains? 
Do this per lineage! And for intergenic regions.

Also: exclude singleton insertions

```{r}
library(vegan)
library(ggplot2)

knockout_m = matrix(nrow=length(ko_genes), ncol=length(metadata$GNUMBER))
rownames(knockout_m) <- ko_genes
colnames(knockout_m) <- metadata$GNUMBER

for (i in 1:length(metadata$GNUMBER)){
  s = metadata$GNUMBER[i]
  knockout_s = subset(knockouts, strain==s)
  knockout_m[,s] <- ifelse(ko_genes %in% knockout_s$gene, 1, 0)
}

# Get accumulation data separately for lineages
acc_df = data.frame()

for (lineage in lineage_order){
  
  gnrs = metadata$GNUMBER[which(metadata$LINEAGE_x==lineage)]
  if (length(gnrs) < 100){next}
  
  m_l = knockout_m[,gnrs]
  m_l <- m_l[rowSums(m_l) != 0,]
  m_lt <- t(m_l)
  
  # Species accumulation curve (random permutations)
  acc <- specaccum(m_lt, method="random", permutations=100)

  df = data.frame(
    sites = acc$sites,
    richness = acc$richness,
    sd = acc$sd,
    perm = acc$perm,
    lineage = rep(lineage,length(acc$sites))
  )
  acc_df = rbind(acc_df, df)
}

summary(as.factor(acc_df$lineage))

col_schemes <- load_color_schemes()

ggplot(subset(acc_df, lineage %in% c('L1', 'L2', 'L3', 'L4', 'L5', 'L6')), aes(x=sites, y=richness, col=lineage)) + geom_line(alpha=1, size=1)+
  geom_line(aes(y=richness-sd,col=lineage), linetype='dashed',size=0.2) +
  geom_line(aes(y=richness+sd,col=lineage), linetype='dashed', size=0.2) +
  xlab("Sampled strains") + ylab("Knocked-out genes") +
  scale_color_manual(values=col_schemes$col_scale_black) +
  theme_bw()


ggsave('/home/cristobal/TB/projects/MTBC_IS6110-poly/figures/figX.accumulaiton_curves.pdf', width=14, height = 10, units = 'cm')

# Model nr. of k.o. genes at 500 strains ~ lineage diversity * mean IS CN
subset(acc_df, sites==500)


```



## Essentiality
How do genes never affected by IS6110 compare to in vitro essential genes, as defined by De Jesus et al. 2017?

Expectation: Many more genes should be needed in clinico!

How many of the knockouts occur only once, how many in multiple strains?

Are knockouts clade-specific, or are some genes affected throughout the MTBC?
```{r}
library(ggplot2)

# Table S3 in De Jesus et al.
orfs_dejesus <- read.csv('manuscript/literatura/DeJesus2017/orfs.csv', sep= '\t')


# Nr. of interrupted genes: 870
refins_dejesus <- merge(refins, orfs_dejesus, by.x='context', by.y="ORF.ID")
refins_dejesus_ne <- subset(refins_dejesus, Final.Call=='NE')

# NE in DeJesus that do not have IS6110 insertions
ne_genes = orfs_dejesus$ORF.ID[which(orfs_dejesus$Final.Call=='NE')]
length(ne_genes)



setdiff(unique(refins_ko$gene_rv), ne_genes)

table(refins_dejesus$Final.Call)

refins_dejesus_es <- subset(refins_dejesus, Final.Call=='ES')
length(unique(refins_dejesus_es$context))

# Plot essentiality categories
ggplot(refins_dejesus, aes(Final.Call)) + geom_histogram(stat='count')
```





# Hotspots of IS6110

Count how many times genic and intergenic regions were hit. 
For the niche model: can we obtain an estimate of how much of the genome is permissible?

Visualization:
  - tree: branches colored according to presence/absence of IS in region (on color per independent insertion)
  - genomic context: show gene model and insertion locations e.g. as triangles, size/col showing nr. of independent events



## Adaptive parallelism index
```{r}
library(ggplot2)
library(ggrepel)
library(viridis)

# Plot: nr of independent insertions vs. median size of clusters with a specific insertion!
head(hotspots)

regions = unique(hotspots$region)

hotspots_summary <- data.frame(
  region = regions,
  nsites = rep(0, length(regions)),
  nbirths = rep(0, length(regions)),
  ndeaths = rep(0, length(regions)),
  strains_min = rep(0, length(regions)),
  strains_max = rep(0, length(regions)),
  strains_mean = rep(0, length(regions)), 
  strains_total = rep(0, length(regions))
)



for (i in 1:length(regions)){
  r = regions[i]
  hotspots_sub = subset(hotspots, region == r)
  hotspots_summary$nsites[i] = nrow(hotspots_sub)
  hotspots_summary$nbirths[i] = sum(hotspots_sub$births)
  hotspots_summary$ndeaths[i] = sum(hotspots_sub$deaths)
  hotspots_summary$strains_min[i] = min(hotspots_sub$nstrains)
  hotspots_summary$strains_max[i] = max(hotspots_sub$nstrains)
  hotspots_summary$strains_mean[i] = mean(hotspots_sub$nstrains)
  hotspots_summary$strains_total[i] = sum(hotspots_sub$nstrains)
}


# Adaptive parallelism index API
hotspots_summary$API = hotspots_summary$nsites * log10(hotspots_summary$strains_mean + 1)
ggplot(hotspots_summary, aes(API)) + geom_density()

# How do measures of parallelism compare?
ggplot(subset(hotspots_summary, nbirths>1), aes(nsites, API)) + geom_point(alpha=0.6) +
  geom_label_repel(data=subset(hotspots_summary, API>20), aes(label=region), label.size=0, size = 3) +
  theme_bw() + 
  xlab("Number of independent insertion sites") + ylab("Parallelism index") +
  scale_color_viridis(option='A', end = 0.8)

ggsave('/home/cristobal/TB/projects/MTBC_IS6110-poly/presentations/labmeeting.21-10-2025/figures/adaptive_parallelism.pdf', width=22, height = 17, units = 'cm')


# Distribution of PI/nsites etc
table(hotspots_summary$nsites)

ggplot(hotspots_summary, aes(nsites)) + geom_histogram() +  theme_bw() + xlab('Number of independent insertion sites') + ylab("Count")

ggsave('/home/cristobal/TB/projects/MTBC_IS6110-poly/presentations/labmeeting.21-10-2025/figures/n_indepentend_sites_dist.pdf', width=15, height = 10, units = 'cm')

```

 
## Plot hotspot on tree
```{r}
library(ggplot2)
library(magrittr)
library(ggtree)
library(tidytree)
library(RColorBrewer)
library(ggrepel)
library(viridis)

hotspot_cnt = as.data.frame(table(hotspots$region))

# Get ancestral states. For region of interest, add states to tree
anc_states_five = read.csv('workflow/results/detettore/AS_5prime.ancestral_likelihoods.tsv', sep='\t')
names(anc_states_five) <- gsub('X', '', names(anc_states_five))

p <- ggtree(tree, layout="circular", col='grey90',size=0.1, alpha=0.8) %<+% metadata


# example: mtbc0_001868 (plcD)
hotspots_of_interest = c('dnaA;dnaN', 'mtbc0_000805;phoP', 'mtbc0_001868','mtbc0_001879;mtbc0_001880','mtbc0_002031')


focal = 'mtbc0_002031'
hotspots_sub = subset(hotspots, region == focal)  
anc_states_sub = anc_states_five[,c('node', hotspots_sub$site_id)]

# Are there strains with multiple insertions in the same region? Yes
table(rowSums(anc_states_sub[,-1]))


# Create two new columns for the tree: insertion_site, births
anc_state_compressed <- data.frame(
  'node' = anc_states_sub$node,
  'insertion_site' = rep(NA, nrow(anc_states_sub))
)

for (site_id in hotspots_sub$site_id){
  anc_state_compressed$insertion_site[which(anc_states_sub[[site_id]]>0.5)] <- site_id
}

p$data$insertion_site = anc_state_compressed$insertion_site[match(p$data$node, anc_state_compressed$node)]

p + geom_tree(data=subset(p$data, !(is.na(insertion_site))), aes(col=insertion_site), size=0.3) + 
  theme(legend.position = "none")


# Save tree to file and crop white space
dest = paste("presentations/labmeeting.21-10-2025/figures/", focal, '.pdf')
ggsave(dest, width = 15, height = 15, units = "cm", device = cairo_pdf) 


```

## Plot gene models and mutations
```{r}
library(ggplot2)

head(genemap)
head(hotspot)

focal = 'mtbc0_000805;phoP'

# Get mutations
hotspot_sub = subset(hotspots, region == focal)
hotspot_sub$position <- as.integer(hotspot_sub$position)

# Create data frame with gene coordinates
focal_splt = strsplit(focal, ';') %>% unlist()
geneA = focal_splt[1]
geneB = focal_splt[2]

genemap_sub = subset(genemap, mtbc0_gene_id == geneA | gene_id == geneA | mtbc0_gene_id == geneB| gene_id == geneB)
genemap_sub$length = genemap_sub$end - genemap_sub$start
genemap_sub$mid = genemap_sub$start + (genemap_sub$length/2)

# --- Compute x-axis limits based on mutation positions ---
x_min <- min(hotspot_sub$position) - 100
x_max <- max(hotspot_sub$position) + 100

# --- Define y positions for layers ---
gene_ymin <- 0
gene_ymax <- 0.02   # thin rectangles
label_y   <- 0.01   # vertically centered labels
mut_y     <- 0.025   # triangles above genes

# --- Plot ---
ggplot() +
  
  geom_rect(
    data = genemap_sub,
    aes(xmin = start, xmax = end, ymin = gene_ymin, ymax = gene_ymax),
    color = "black", fill = "steelblue", alpha = 0.6
  ) +
  geom_text(
    data = genemap_sub,
    aes(x = mid, y = label_y, label = gene_id),
    size = 5, vjust = 0.5, color = "black"
  ) +
  geom_point(
    data = hotspot_sub,
    aes(x = as.integer(position), fill = births, y = mut_y),
    shape = 25, size = 6, alpha = 0.4, color = "black"
  ) +
  scale_fill_viridis_c(option = "plasma", direction = -1) +
  xlab("Genomic position") +
  coord_cartesian(xlim = c(x_min, x_max), ylim = c(0, 0.05), expand = FALSE) +
  theme_minimal() +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    legend.title       = element_blank(),
    axis.title.y       = element_blank(),
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    axis.line.y        = element_blank()
  )
  
```




# Constrained niche model
A simple model to explain IS copy number variation, hotspots, symmetry etc.:
Few loci in the genome where IS can insert safely. This are hit by chance, and once hit,
additional copies increase the birth rate. Model implemented in contrained_niche_evolution.py.

Use downsampled tree, e.g. 500 strains.

Fit model to data using approximate Bayesian computation. Summary statistics: median CN, var(CN)_clades, ...
Chose initial distributions according to real data (birth rate, death rate, max_number ...). 
Fixed: initial_copies = 1, max_niches = 38
Fitted: p_lethal, r_birth, r_death

p_lethal = 0.95
r_transpose = 4000
r_delete = 0
max_niches = 40
initial_copies = 1



## Approximate Bayesian computation

Simulation test run
```{r}
args <- c(
  "workflow/scripts/niche_model.py",
  "--tree", "workflow/results/niche_model/subsampled_tree.rooted.nex",
  "--p_lethal",  "0.6621324",
  "--r_birth", "4507.4047070",
  "--r_death", "249.1706106",
  "--initial_copies", "1",
  "--max_copies", "40"
)


stdout = system2("/home/cristobal/anaconda3/envs/ete3/bin/python", args, stdout=TRUE)
cn_sim = read.delim(text=stdout, sep='\t')
```


Test run with the spatially explicit model
```{r}
args <- c(
  "workflow/scripts/niche_model.py",
  "--tree", "workflow/results/niche_model/subsampled_tree.rooted.nex",
  "--p_lethal",  "0.6621324",
  "--r_birth", "4507.4047070",
  "--r_death", "249.1706106",
  "--initial_copies", "1",
  "--max_copies", "40",
  "--spatial",
  "--niches", "10000"
)


stdout = system2("/home/cristobal/anaconda3/envs/ete3/bin/python", args, stdout=TRUE)
cn_sim = read.delim(text=stdout, sep='\t')
```


Summary statistics
```{r}

library(ape)
library(ggplot2)
library(ggtree)
library(viridis)
library(dplyr)

load('is6110.RData')

get_summary_stats <- function(copy_numbers, metadata){
  
  copy_numbers$lineage = metadata$lineage[match(copy_numbers[,1], metadata$GNUMBER)]
  
  cladevar = copy_numbers %>% 
    group_by(lineage) %>%
    summarize(Mean = mean(CN), SD = sd(.[[2]]))
  
  summdata = list(
    'Mean' = mean(copy_numbers[,2]),
    'SD' = sd(copy_numbers[,2]),
    'min' = min(copy_numbers[,2]),
    '25%' = quantile(copy_numbers[,2], 0.25),
    '50%' = quantile(copy_numbers[,2], 0.5),
    '75%' = quantile(copy_numbers[,2], 0.75),
    'max' = max(copy_numbers[,2]),
    'cladevar' = var(cladevar$Mean),
    'cladesd' = sd(cladevar$Mean)
  ) 
  return(summdata)
}


treesim = read.tree('workflow/results/niche_model/subsampled_tree.rooted.nex')
plot(treesim, show.tip.label = FALSE)
copy_numbers_sub <- subset(copy_numbers, strain %in% tree$tip.label)
head(copy_numbers_sub)
stats_obs = get_summary_stats(copy_numbers_sub,metadata)
stats_obs

```


## Get posterior distributions
```{r}
library(abc)
library(magrittr)


params_mat <- read.csv('workflow/results/niche_model/abc_params.tsv', sep='\t') %>% as.matrix()
colnames(params_mat) <- c('p_lethal', 'r_birth', 'r_death')
dim(params_mat)
summary(params_mat)



summaries_mat <- read.csv('workflow/results/niche_model/abc_summaries.tsv', sep='\t') %>% as.matrix()
colnames(summaries_mat) <- c('Mean', 'SD', 'min', '25%', '50%',  '75%', 'max', 'cladevar', 'cladesd')
dim(summaries_mat)

# Run abc (rejection followed by local linear regression)
abc_res <- abc(target = stats_obs,
               param = params_mat,
               sumstat = summaries_mat,
               tol = 0.001,   # keep top 1% closest sims
               method = "neuralnet")

summary(abc_res)
abc_res
plot(abc_res, params_mat)
hist(abc_res)

plot(density(abc_res$dist))

```

Plot parameter combinations vs distance to visualize peaks
```{r}
library(plotly)

str(abc_res)

abc_res_best = abc_res$region

abc_dists <- data.frame(
  p_lethal =  params_mat[,'p_lethal'][abc_res_best],
  r_birth = params_mat[,'r_birth'][abc_res_best],
  r_death = params_mat[,'r_death'][abc_res_best],
  p_lethal_adj = abc_res$adj.values[,'p_lethal'],
  r_birth_adj = abc_res$adj.values[,'r_birth'],
  r_death_adj = abc_res$adj.values[,'r_death'],
  dist = abc_res$dist[abc_res_best]
)

head(abc_dists)

# r_birth vs p_lethal
ggplot(abc_dists, aes(r_birth, p_lethal, col=dist)) + geom_point(size=2, alpha=0.5) +
  scale_color_viridis(option = 'turbo')

# with adjusted values
ggplot(abc_dists, aes(r_birth_adj, p_lethal_adj, col=dist)) + geom_point(size=3, alpha=0.5) +
  scale_color_viridis(option = 'turbo') + theme_bw() + xlab('Birth rate') + ylab('Proportion of inaccessible sites')

# How do adjusted and unadjusted values compare?
ggplot(abc_dists, aes(r_birth_adj, r_birth, col=dist)) + geom_point(size=2, alpha=0.5) +
  scale_color_viridis(option = 'turbo')

ggplot(abc_dists, aes(p_lethal_adj, p_lethal, col=dist)) + geom_point(size=2, alpha=0.5) +
  scale_color_viridis(option = 'turbo')


# 3d plot
plot_ly(
  abc_dists,
  x = ~r_birth, y = ~p_lethal, z = ~-dist,
  intensity = ~-dist,
  type = "mesh3d",
  colors= colorRamp(rainbow(5))
)


```

Cross validation
```{r}
cv = cv4abc(params_mat,summaries_mat, abc_res, 100, c(0.01, 0.001), 'mean')

```




Get best model and re-run to get copy numbers
```{r}
best_i = which(abc_res$dist==min(abc_res$dist))
params_mat[best_i,]
summaries_mat[best_i,]
stats_obs


args <- c(
  "workflow/scripts/niche_model.py",
  "--tree", "workflow/results/niche_model/subsampled_tree.rooted.nex",
  "--p_lethal",  "0.6621324",
  "--r_birth", "4507.4047070",
  "--r_death", "249.1706106",
  "--initial_copies", "1",
  "--max_copies", "40"
)


stdout = system2("/home/cristobal/anaconda3/envs/ete3/bin/python", args, stdout=TRUE)
cn_sim = read.delim(text=stdout, sep='\t')
head(cn_sim)
```


## Figure

A) Observed and simulated copy numbers with parameter values sampled from the posterior. B) Posterior distributions for birth rate, death rate, and the proportion of suitable niches in the genome.

```{r}
library(ggtree)
library(ggplot2)
library(ape)
library(viridis)


tree <- read.tree('workflow/results/niche_model/subsampled_tree.rooted.nex')

p <- ggtree(tree, alpha=0.1, size=0) 
# Add observed values to tree
p$data$copy_number <- copy_numbers$CN[match(p$data$label, copy_numbers$strain)]
# Add simulated values to tree
p$data$copy_number_simulated <- cn_sim$copies[match(p$data$label, cn_sim$tip)]

# Plot trees
## Observed
a <- p + geom_tree(aes(col=copy_number)) + scale_color_viridis(option = 'turbo', guide="none") + 
  theme(panel.border = element_rect(color = "black", fill = NA, linewidth = 0.5, linetype='dashed'))

## Simulated
b <- p + geom_tree(aes(col=copy_number_simulated)) + scale_color_viridis(option = 'turbo', guide="none")

## Combine
cowplot::plot_grid(a,b)

# Plot posterior distributions



```


## Posterior predictive checks

```{r}
post_params <- post$adj.values    # or post$unadj.values

# Draw posterior samples
n_sim <- 1000       # how many posterior predictive simulations  
idx <- sample(1:nrow(post_params), n_sim, replace = TRUE)
pp_params <- post_params[idx, ]

# Simulate data
pp_sims = 

# Compute summary stats
pp_sumstats <- t(sapply(pp_sims, compute_sumstats)) 


# Compare predictive distribution vs observed summary stats
pp_check <- abc::gfit(target = obs_stats,
                      sumstat = sumstats_matrix,
                      nb.repl = 1000)  # posterior predictive sims
plot(pp_check)

# Posterior predictive P values
pp_pvals <- colMeans(pp_sumstats >= matrix(obs_stats,
                                           nrow=n_sim,
                                           ncol=length(obs_stats),
                                           byrow=TRUE))
pp_pvals



```

## Figure

