---
title: "explorations"
output: html_document
date: "2025-07-10"
editor_options: 
  chunk_output_type: console
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# What is lineage C?
```{r}
genomes = read.csv("/home/cristobal/TB/projects/MTBC_IS6110-poly/metadata/GENOMES_2025-02-18_14-49-13.tsv", sep='\t')

C_gnrs = metadata$GNUMBER[which(metadata$LINEAGE_x=='C')]
genomes_c = subset(genomes, G_NUMBER %in% C_gnrs)
View(genomes_c)



```




# Birth and death of IS6110
- Parsimony mapping of insertions: phangorn::ancestral.pars() or castor::asr_max_parsimony()
- Stochastic Mapping of Insertions (e.g., via phytools::make.simmap)
- BadiRate, COUNT, or GLOOME: tools for gene family evolution
- TE load as quantitative trait: geiger::fitContinuous or phytools::fastAnc



## Shannon entropy to test wheter IS is less or more active
```{r}
# Compare lineages: how conserved are patterns? 
library(dplyr)
library(ggplot2)

# Function to compute SFS and entropy for a given subset of strains
compute_sfs_entropy <- function(mat) {

  # Frequency of '1's per site
  freqs <- colSums(mat)
  
  # Site Frequency Spectrum (as table)
  sfs <- table(freqs) # counts per frequency
  sfs <- data.frame(freq = as.numeric(names(sfs)), count = as.numeric(sfs))
  
  # Shannon entropy per site
  p <- freqs / nrow(mat)
  p[p == 0 | p == 1] <- ifelse(p[p == 0 | p == 1] == 0, 1e-10, 1 - 1e-10) # avoid log(0)
  entropies <- -p * log2(p) - (1 - p) * log2(1 - p)
  
  list(sfs = sfs, mean_entropy = mean(entropies), entropy_values = entropies)
}


compute_sfs_entropy <- function(mat) {

  # Frequency of '1's per site
  freqs <- colSums(mat)
  
  # Site Frequency Spectrum (as table)
  sfs <- table(freqs) # counts per frequency
  sfs <- data.frame(freq = as.numeric(names(sfs)), count = as.numeric(sfs))
  
  # Shannon entropy per site
  p <- freqs / nrow(mat)
  p[p == 0 | p == 1] <- ifelse(p[p == 0 | p == 1] == 0, 1e-10, 1 - 1e-10) # avoid log(0)
  entropies <- -p * log2(p) - (1 - p) * log2(1 - p)
  
  list(sfs = sfs, mean_entropy = mean(entropies), entropy_values = entropies)
}

pairwise_diff <- function(mat){
  # Convert to numeric matrix if logical
  mat <- as.matrix(mat) * 1
  L <- ncol(mat)

  # Precompute number of 1's per strain
  ones <- mat %*% rep(1, L)         # vector of row sums

  # Agreement matrix: 
  # shared 1's = mat %*% t(mat)
  # shared 0's = (1-mat) %*% t(1-mat)
  shared_ones  <- mat %*% t(mat)
  shared_zeros <- (1 - mat) %*% t(1 - mat)

  # Total matches = shared ones + shared zeros
  matches <- shared_ones + shared_zeros

  # Hamming distance matrix (normalized)
  dist_mat <- 1 - matches / L
  return(dist_mat)
}






is6110_entropy <- list()
is6110_sfs <- data.frame()

for (l in lineage_order){
  if (l %in% c('L7')){next}
  tm_sub <- subset_tree_matrix(tree, matrices$`5`, metadata, l, 'lineage') 
  entrpy <- compute_sfs_entropy(tm_sub[[2]])
  is6110_entropy[[l]] <- entrpy
  print(paste(l, entrpy$mean_entropy))
}








# Combine SFS data for plotting
sfs_df <- bind_rows(lapply(results, function(x) {
  mutate(x$sfs, subset = x$subset)
}))

# Plot Site Frequency Spectra for each subset
ggplot(sfs_df, aes(x = freq, y = count, color = factor(subset))) +
  geom_line() +
  geom_point() +
  theme_minimal() +
  labs(x = "Number of strains with insertion",
       y = "Number of sites",
       color = "Subset",
       title = "Site Frequency Spectra (SFS) per subset")

# Mean entropy per subset
entropy_df <- data.frame(
  subset = 1:n_reps,
  mean_entropy = sapply(results, function(x) x$mean_entropy)
)

print(entropy_df)



```

## Ancestral state reconstruction: output likelihoods?
```{r}
library(magrittr)
library(ape)
library(castor)

table(out$metadata$LINEAGE_x)

# Test with L5
lineage_sub <- subset_tree_matrix(
      out$tree, 
      out$matrix5, 
      out$metadata, 
      'L5', 
      'lineage')
    
tree <- out$tree
matrix <- out$matrix5

site_states <- data.frame(
  'node' = 1:(length(tree$tip.label) + tree$Nnode)
)

# Loop over each insertion site (column)
for (i in 1:ncol(matrix)) {
  
  if (i%%500 == 0){
    print(i)
  }
  
  site_vec <- matrix[, i]
  site_id = colnames(matrix)[i]
  
  # Skip constant or NA-only sites
  if (any(is.na(site_vec)) || length(unique(site_vec)) == 1) next
  
  # Ensure it's numeric binary. Add 1 because a trait can' be zero with asr_max_parsimony
  site_vec_mapped <- map_to_state_space(site_vec)
  
  # Reconstruct ancestral states
  asr_result <- asr_max_parsimony(
    tree, 
    tip_states = site_vec_mapped$mapped_states
    )
  
  # For each node, store likelihood of presence
  lh_presence <- asr_result$ancestral_likelihoods[,2]
  
  # Combine tip and internal node states
  full_states <- c(site_vec, lh_presence)
  full_states <- as.data.frame(full_states)
  names(full_states) <- site_id
  site_states <- cbind(site_states, full_states)
  
}
```



## Lineage-wise ancestral state reconstruction
```{r}
# Lineage-wise
as_branches = data.frame()
as_sites = data.frame()

trees = list()
matrices = list()

for (lineage in schemata$lineage_order){
  
  print(lineage)
  
  nr_strains = nrow(subset(out$metadata, LINEAGE_x == lineage))
  
  if (nr_strains > 5){
    lineage_sub <- subset_tree_matrix(
      out$tree, 
      out$matrix5, 
      out$metadata, 
      lineage, 
      'lineage')
    
    tree_l <- lineage_sub[[1]]
    matrix_l <- lineage_sub[[2]] %>% as.matrix()
    matrix_l <- matrix_l[tree_l$tip.label, ]  # Ensure order matches tree tips
    
    # Reconstruct ancestral sites
    as = is_birth_death(tree_l, matrix_l)
    as$branch_summary$lineage <- rep(lineage, nrow(as$branch_summary))
    as$site_summary$lineage <- rep(lineage, nrow(as$site_summary))
    
    as_branches <- rbind(as_branches, as$branch_summary)
    as_sites <- rbind(as_sites, as$site_summary)
    
    trees[[lineage]] <- tree_l
    matrices[[lineage]] <- matrix_l
  }
}
```


## "Instable" sites: check sites with many deaths/births
```{r}
library(plotly)

head(as_branches)
head(as_sites)

# Per lineage
ggplot(subset(as_sites), aes(ins_per_site, del_per_site)) + 
  geom_jitter(alpha=0.5, size=1.5) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) +
  xlab("Births per site") + ylab("Deaths per site") +
  facet_wrap(~lineage)


# With the full tree
ggplot(as_full$site_summary, aes(ins_per_site, del_per_site)) + 
  geom_jitter(alpha=0.5, size=1.5) +
  theme_bw(base_family = "CMU Sans Serif", base_size = 12) +
  xlab("Births per site") + ylab("Deaths per site")

table(as_full$site_summary$ins_per_site)
table(as_full$site_summary$del_per_site)

# Color deaths and births for single site on branches -------------------------
tree = out$tree
matrix = out$matrix5
site_id = 'X5.ua619'
site_vec <- matrix[, site_id]

# Ensure it's numeric binary. Add 1 because a trait can' be zero with asr_max_parsimony
site_vec <- as.integer(site_vec)
site_vec <- as.integer(site_vec) + 1

# Reconstruct ancestral states
asr_result <- asr_max_parsimony(tree, tip_states = site_vec, Nstates = 2)
states <- asr_result$ancestral_states  # matrix: Nnodes x Nstates
states <- states - 1  # Reconvert to 0/1

# Combine tip and internal node states
full_states <- c(site_vec-1, states)
names(full_states) <- 1:(length(tree$tip.label) + tree$Nnode)

site_states = data.frame()

for (j in 1:nrow(tree$edge)) {
  parent <- tree$edge[j, 1]
  child  <- tree$edge[j, 2]
  state <- paste(full_states[parent], full_states[child], sep = '_')
  row = data.frame(
    'parent' = parent, 
    'child' = child,
    'state' = state
  )
  site_states = rbind(site_states, row)
}

site_states$id <- paste(site_states$parent, site_states$child, sep='-')

p <- ggtree(tree, size=0.0001, alpha=0.5) %<+% out$metadata
p$data$state <- NULL
p$data$id <- paste(p$data$parent, p$data$node, sep='-')
p$data$state <- site_states$state[match(p$data$id, site_states$id)]
table(p$data$state)

p <- p + geom_tree(aes(color=state)) + 
  scale_color_manual(values = c("0_0" = "grey90", "0_1" = "green","1_1" = "grey70","1_0" = "red"))

ggplotly(p)




# Plot CN change on tree ------------------------------------------------------
head(as_sites)
lineage='L3'

lineage_sub <- subset_tree_matrix(
    out$tree, 
    out$matrix5, 
    out$metadata, 
    lineage, 
    'lineage')

metadata_l = subset(out$metadata, LINEAGE_x == lineage)

p <- ggtree(lineage_sub[[1]], layout='circular', size=0.001) %<+% metadata_l

p$data$insertions <- rep(0, nrow(p$data))
p$data$deletions <- rep(0, nrow(p$data))
p$data$net_change <- rep(0, nrow(p$data))

for (i in 1:nrow(as_branches)){
  row = as_branches[i,]
  if (row$lineage == lineage){
    is_change = row$insertions - row$deletions
    
    p$data$insertions[which((p$data$node == row$child) & (p$data$parent==row$parent))] = row$insertions
    p$data$deletions[which((p$data$node == row$child) & (p$data$parent==row$parent))] = row$deletions
    p$data$net_change[which((p$data$node == row$child) & (p$data$parent==row$parent))] = is_change
  }
}

p + geom_tree(aes(color=net_change)) + 
  scale_color_viridis(option = 'turbo')
  
```


```{r}
library(ape)
library(phangorn)
library(castor)
library(dplyr)
library(ggplot2)
library(reshape2)
library(magrittr)
library(ggtree)

source('workflow/notebooks/utils.R')

metadata <- read.csv('data/10k/metadata_reduced.tsv', sep='\t')
# Remove the two outlier orygis strains
metadata <- subset(metadata, !(GNUMBER %in% c('G32738', 'G32741'))) %>% droplevels()

tree <- read.tree('explorations/njtree/10k.njtree.noCanettii.cleaned.nex') 
matrix <- read.csv('workflow/results/10k/detettore/ALL_presence-absence.tsv', sep='\t')
rownames(matrix) <- matrix[,1]
matrix <- matrix[,-1]

rates = data.frame()
trees = list()
matrices = list()

for (lineage in lineage_order){
  
  print(lineage)
  
  lineage_sub <- subset_tree_matrix(tree, matrix, metadata, lineage, 'lineage')
  tree_l <- lineage_sub[[1]]
  matrix_l <- lineage_sub[[2]] %>% as.matrix()
  
  # Ensure order matches tree tips
  matrix_l <- matrix_l[tree_l$tip.label, ]

  # Get birth and death rates
  rates_l <- is_birth_death(tree_l, matrix_l)
  rates_l$lineage <- rep(lineage, nrow(rates_l))
  rates <- rbind(rates, rates_l)
  
  trees[[lineage]] <- tree_l
  matrices[[lineage]] <- matrix_l
}


# Plot distributions of birth and death rates
melt(
  rates, 
  id.vars = c('edge_id', 'lineage'), 
  measure.vars = c('birth_rate', 'death_rate')) %>%
  ggplot(aes(y=value, x=lineage, fill=variable)) + 
  geom_boxplot(position='dodge') + scale_y_log10()
  

rates$lineage <- factor(rates$lineage, levels=lineage_order)

# Do births correspond to deaths, suggesting cut-and-paste (or methodological bias)
ggplot(subset(rates, birth_rate>0 & death_rate>0), aes(x=birth_rate, y=death_rate)) +
  geom_point() + geom_smooth() +
  scale_x_log10() + scale_y_log10() +
  facet_wrap(~lineage, scales='free') 


# Branch length vs number of insertions/deletions
melt(
  rates, 
  id.vars=c('edge_id', 'lineage', 'branch_length'), 
  measure.vars = c('insertions', 'deletions')) %>%
  ggplot(aes(x=branch_length, y=value, col=variable)) +
  geom_point() + geom_smooth()



# Plot rates on tree
lineage = 'La1'
rates_l  <- subset(rates,lineage==lineage) %>% droplevels()
names(rates_l)[1] <- 'parent_x'
metadata_l <- subset(metadata, LINEAGE_x == lineage)

p <- ggtree(trees[[lineage]]) %<+% metadata_l
treedata <- p$data
treedata <- left_join(treedata, rates_l, by=c("node" = "child"))
p <- p + geom_nodelab(data=treedata,aes(label=insertions), col='red', nudge_x = 0.00001, nudge_y = 0.00001)

gheatmap(p, matrices[[lineage]], colnames = FALSE) + scale_fill_manual(values=c("white", "black"))




# Test for sublineage
metadata_l = subset(metadata, LINEAGE_x == 'L4')
table(metadata_l$SUBLINEAGE_x)

lineage = 'L4.4.1.1'
lineage_sub <- subset_tree_matrix(tree, matrix, metadata, lineage, 'sublineage')

tree_l <- lineage_sub[[1]]
matrix_l <- lineage_sub[[2]] %>% as.matrix()
  
# Ensure order matches tree tips
matrix_l <- matrix_l[tree_l$tip.label, ]

# Get birth and death rates
rates_l <- is_birth_death(tree_l, matrix_l)

# Plot
p <- ggtree(tree_l) %<+% metadata_l
treedata <- p$data
treedata <- left_join(treedata, rates_l, by=c("node" = "child"))
p <- p + 
  geom_nodelab(data=treedata,aes(label=insertions), col='red', nudge_x = 0.00001, nudge_y = 0.00001) + 
  geom_tiplab(data=treedata,aes(label=insertions), col='red', nudge_x = 0.00001, nudge_y = 0.00001)

gheatmap(p, matrix_l, colnames = FALSE) + scale_fill_manual(values=c("white", "black"))

```

# GWAS

## Burden analysis with all genes
```{r}
library(ape)
library(ggplot2)
library(ggtree)
library(magrittr)
library(viridis)
library(RColorBrewer)
library(cowplot)
library(ggtreeExtra)
library(ggrepel)


gwas <- read.csv('workflow/results/10k/gwas/burden_all_genes.tsv', sep='\t')

# Remove sample names to speed up things
gwas <- gwas[,-c(8,9 )]


permutation_results <- list.files('workflow/results/10k/gwas/permutations', pattern="burden_all*", full.names = T)

permut <- data.frame()
for (f in permutation_results){
  
  df <- read.csv(f, sep='\t')
  df <- df[,-c(8,9,10,11)]
  
  bn <- basename(f)
  match <- regexpr("[0-9]+", bn)
  n <- regmatches(bn, match)
  df$n <- rep(n, nrow(df))
  permut <- rbind(permut, df)
}

# Plot distribution of P values from permuted data sets
head(permut)
permut$logp <- -log10(permut$lrt.pvalue)

ggplot(permut, aes(lrt.pvalue)) + 
  geom_density(aes(group=n), alpha=0.2) + 
  theme_bw()

# Get threshold: 1 and 5 percentiles
gwas_thresholds = quantile(permut$lrt.pvalue, probs=c(0.01, 0.05))

gwas$logp <- -log10(gwas$lrt.pvalue)
sign <- -log10(gwas_thresholds[[1]])

gwas$i <- 1:nrow(gwas)

ggplot(gwas, aes(x=i, y=logp)) + 
  geom_point(alpha=0.2) + theme_minimal() +
  geom_hline(aes(yintercept=sign), col='red', linetype='dashed') +
  ylab('-Log10(P)') + xlab('') + theme(axis.text.x=element_blank())


# Volcano plot
ggplot(gwas, aes(y=logp, x=beta, col=100*variant_h2)) + geom_point(alpha=1) +
  geom_hline(aes(yintercept=sign), col='red', linetype='dashed') +
  theme_bw(base_family = "CMU Sans Serif") + ylab('-Log10(P)') + xlab('Effect size (beta)') +
  geom_label_repel(data=subset(gwas, logp>sign),aes(label=variant), label.size=0, size = 3) +
  scale_color_viridis(option='turbo')

# Better: same tree as in fig2, with band showing presence/absence of variants in genes

```



## GWAS with all variants
```{r}
library(ggplot2)

gwas_snps <- read.csv('workflow/results/10k/gwas/all_variants.tsv', sep='\t')

names(gwas_snps)
gwas_snps <- gwas_snps[,-c(8,9)]
head(gwas_snps)

gwas_snps$logp <- -log10(gwas_snps$lrt.pvalue)
gwas_snps$pos <- as.numeric(gsub(".*_([0-9]+).*", "\\1", gwas_snps$variant))

sign = 0.05
sign_corr = 0.05/nrow(gwas_snps)
sign_corr_logp = -log10(sign_corr)

# Remove duplicate frequencies
#gwas_snps <- gwas_snps[!duplicated(gwas_snps$lrt.pvalue), ]

ggplot(gwas_snps, aes(x=pos, y=logp)) + geom_point(alpha=0.2) +
theme_bw() + ylab('-Log10(P)') + xlab('Position') + geom_hline(aes(yintercept=sign_corr_logp), col='red', linetype='dashed')

# Check top 10 SNPs
gwas_snps <- gwas_snps[with(gwas_snps, order(-logp)),]
head(gwas_snps[,-c(8,9)], 10)

# volcano plot
ggplot(gwas_snps, aes(y=logp, x=beta, col=100*variant_h2)) + geom_point(alpha=0.5) +
  geom_hline(aes(yintercept=sign_corr_logp), col='red', linetype='dashed') +
  theme_bw() + ylab('-Log10(P)') + xlab('Effect size (beta)') +
  scale_color_viridis(option='turbo')

  #geom_label_repel(data=subset(gwas_snps, logp>sign_corr_logp),aes(label=variant), label.size=0)
  




# Show best hit on tree
check <- hit_on_tree(
  gwas_snps, out$tree, 
  out$copy_numbers, 
  out$metadata, 
  'MTB_anc_1107917_G_T'
  )

check$table
check$tree







```

## Burden analysis with all genes
```{r}
library(ggplot2)
library(viridis)

burden <- read.csv('workflow/results/10k/gwas/burden_all_genes.tsv', sep='\t')
head(burden)

rec_genes <- read.csv('explorations/pyseer/r3_genes.metadata.tsv', sep='\t')
burden$r3 <- rep('No', nrow(burden))
burden$r3[which(burden$variant %in% rec_genes$Rv_number)] <- 'Yes'
burden$logp <- -log10(burden$lrt.pvalue)

ggplot(burden, aes(x=variant, y=logp, col=beta)) + geom_point() +
  scale_color_viridis_c(option = 'B', end = 0.8)

burden <- burden[order(burden$lrt.pvalue), ]
burden$indx <- row(burden[1])


# Remove duplicate frequencies
burden_corr <- burden[!duplicated(burden$lrt.pvalue), ]
ggplot(burden_corr, aes(x=variant, y=logp, col=beta)) + geom_point() +
  scale_color_viridis_c(option = 'B', end = 0.8)

# Volcano plot
ggplot(burden, aes(y=logp, x=beta)) + geom_point(alpha=0.3) +
  geom_hline(aes(yintercept=sign_corr_logp), col='red', linetype='dashed') +
  theme_bw() + ylab('-Log10(P)') + xlab('Effect size (beta)') +
  xlim(c(-25,25))
  


```



# Reference insertions


## Number of mutated position per feature
```{r}
library(ggplot2)
library(ggrepel)

refins <- read.delim('workflow/results/10k/detettore/ALL_reference_insertions.extended.tsv', sep='\t', header=TRUE)

context_df = data.frame()
context_l = list()

for (feature in unique(refins$gene_alt)){
  
  feature_df = subset(refins, gene_alt == feature)
  nr_pos = length(unique(feature_df$position))
  nr_lineages = length(unique(feature_df$lineage))
  nr_strains = length(unique(feature_df$strain))
  
  resistance = feature_df$resistance[1]
  
  if (grepl(';', feature)){
    cat = 'intergenic'
  }
  else {
    cat = 'knockout'
  }

  row = data.frame(feature, nr_pos, nr_lineages, nr_strains, cat, resistance)
  context_df = rbind(context_df, row)
}

View(context_df)


# How many regions with more than 10, 20 mutated positions?
length(which(context_df$nr_pos>10))
length(which(context_df$nr_pos>20))

# Nr strains versus nr of mutated positions
known_hotspots = c("plcC","plcB", "dnaA;dnaN", 'ctpD', 'mtbc0_000805;phoP', 'mmpS1')


# Known hotspots
ggplot(context_df, aes(nr_pos, nr_strains)) + geom_point(alpha=0.5) +
  xlab("Number of mutated positions") + ylab("Number of strains") + 
  geom_point(data=subset(context_df, feature %in% known_hotspots),col='red') +
  geom_text_repel(data=subset(context_df, feature %in% known_hotspots), aes(label=feature), col='red') + 
  theme_bw()

# More hotspots...
ggplot(context_df, aes(nr_pos, nr_strains)) + geom_point(alpha=0.5) +
  xlab("Number of mutated positions") + ylab("Number of strains") + 
  geom_point(data=subset(context_df, nr_pos>40 | nr_strains > 500),col='blue') +
  geom_text_repel(data=subset(context_df, nr_pos>40 | nr_strains > 500), aes(label=feature), col='blue') + 
  theme_bw() + facet_wrap(~cat)


# What about resistance genes?
a <- ggplot(subset(context_df, resistance == 'Yes'), aes(nr_pos, nr_strains)) + geom_point(alpha=0.5) + 
  xlab("Number of mutated positions") + ylab("Number of strains") + 
  geom_text_repel(aes(label=feature)) +
  theme_bw()

b <- ggplot(subset(context_df, resistance == 'Yes' & nr_pos < 10), aes(nr_pos, nr_strains)) + geom_jitter(alpha=0.5) + 
  xlab("Number of mutated positions") + ylab("Number of strains") + 
  geom_text_repel(aes(label=feature)) +
  theme_bw()


cowplot::plot_grid(a,b, labels('A', 'B'), nrow = 1, ncol=2)

ggplot(context_df, aes(nr_pos)) + geom_density()
summary(context_df$nr_pos)

```

## Word clouds
```{r}
library(wordcloud)

# Get context per lineage
refins <- read.delim('workflow/results/10k/detettore/ALL_reference_insertions.extended.tsv', sep='\t', header=TRUE)


context_lineage = list()

for (l in unique(refins$lineage)){
  
  refins_l = subset(refins, lineage==l) %>% droplevels()
  context_df = data.frame()
  
  for (feature in unique(refins_l$gene_alt)){
    
    feature_df = subset(refins_l, gene_alt == feature)
    
    nr_pos = length(unique(feature_df$position))
    nr_lineages = length(unique(feature_df$lineage))
    nr_strains = length(unique(feature_df$strain))
    
    resistance = feature_df$resistance[1]
    
    if (grepl(';', feature)){
      cat = 'intergenic'
    }
    else {
      cat = 'knockout'
    }
    row = data.frame(feature, nr_pos, nr_lineages, nr_strains, cat, resistance)
    context_df = rbind(context_df, row)
  }
  context_lineage[[l]] = context_df
}

str(context_lineage)

wordcloud(context_lineage$L1$feature, context_lineage$L1$nr_strains)


# Nr of mutated positions, by lineage!
l="L2"
df = context_lineage[[l]]

ggplot(df, aes(nr_pos, nr_strains)) + geom_point(alpha=0.5) +
  xlab("Number of mutated positions") + ylab("Number of strains") + 
  geom_point(data=subset(df, nr_pos>10 | nr_strains > 50),col='blue') +
  geom_text_repel(data=subset(df, nr_pos>10 | nr_strains > 50), aes(label=feature), col='blue') + 
  theme_bw() + facet_wrap(~cat) + ggtitle(l)


```


## Gviz: visualize region of interest
```{r}
library(Gviz)
library(rtracklayer)
library(GenomicFeatures)
library(viridis)
library(RColorBrewer)
library(stringr)


options(ucscChromosomeNames=FALSE)
gtrack <- GenomeAxisTrack(labelPos = "below")
sTrack <- SequenceTrack(
  "/home/cristobal/TB/projects/detettore6110/resources/reference/MTBC0_v1.1.fasta")

# Read genes and create genes track
gff = read.delim("/home/cristobal/TB/projects/detettore6110/resources/reference/MTBC0v1.1_PGAP_annot.gff", sep='\t', comment.char = '#', header=F)

gff = subset(gff, V3=='CDS')

genes_df = data.frame()

for (i in 1:nrow(gff)){
  row = gff[i,]
  chromosome = row$V1
  start = row$V4
  end = row$V5
  strand = row$V7
  
  parent = str_extract(row$V9, "(?<=Parent=)[^;]+")
  parent = str_sub(parent, 6) 
  gene = str_extract(row$V9, "(?<=gene=)[^;]+")
  product = str_extract(row$V9, "(?<=product=)[^;]+")
  
  rv_name = get_alternative_gene_name(parent, genemap)
  
  row_df = data.frame(chromosome, start, end, strand, parent, gene, product, rv_name)
  genes_df = rbind(genes_df, row_df)
}

genesTrack <- AnnotationTrack(genes_df, id=genes_df$rv_name, name="Gene", featureAnnotation="id", fontcolor.feature = "black", fill="#ffe680", cex=0.8)


# Create track with IS insertions
refins <- read.delim('workflow/results/10k/detettore/ALL_reference_insertions.extended.tsv', sep='\t', header=TRUE)

refins$ID <- paste(refins$position, refins$strand, sep = '_')
length(unique(refins$ID))
length(unique(refins$position))

lineage_order = c('L1', 'L2', 'L3', 'L4', 'L5', 'L6', 'L7', 'L8', 'L9', 'L10','La1', 'La2', 'La3')


# Better: one annotation track per lineage, is fill color
# for frequencies
lineage_order_no7 <- lineage_order[-7]

lineage_annotations = list()

for (i in 1:length(lineage_order_no7)){  # Ignore L7
  
  l = lineage_order_no7[i]
  print(l)
  
  lineage_df = data.frame()
  
  refins_l = subset(refins, lineage==l) %>% droplevels()
  dim(refins_l)
  #event_count = table(refins_l$ID)
  #events = unique(refins_l$ID)
  event_count = table(refins_l$position)
  events = unique(refins_l$position)
  
  if (length(events) == 0){
    next
  }
  
  for (j in 1:length(events)){
    
    #ins = unique(refins_l$ID)[j]
    #ins_splt = strsplit(ins, '_')
    #pos = ins_splt[[1]][1]
    #strand = ins_splt[[1]][2]
    pos = events[j]
    strand = '+'
    
    #count = event_count[[ins]]
    count = event_count[[as.character(pos)]]
    n_strains = length(unique(refins_l$strain))
    freq =  count / n_strains
      
    df = data.frame(
      chromosome = 'MTBC0',
      start = as.numeric(pos),
      end = as.numeric(pos),
      strand = strand,
      name = pos,
      count = count,
      freq = freq
    )
    lineage_df = rbind(lineage_df, df)
  }
  
  # Set panel color
  if (i%%2 == 0){
    bg_color = 'white'
  }
  else {
    bg_color = 'grey95'
  }
  
  lineage_annotations[[l]] = AnnotationTrack(
    lineage_df, 
    name=l, 
    background.title=mtb_col_scale[[l]],
    #fill=mtb_col_scale[[l]],
    background.panel=bg_color,
    fill='black'
    )
}

tracklist = list(gtrack, genesTrack)
tracklist = c(tracklist, lineage_annotations)

# Plot region
subset(genes_df, rv_name=='phoP')

start=856033
end=856776
offset = 500


plotTracks(tracklist, from = start-offset, to = end+offset, sizes = c(2,2,1,1,1,1,1,1,1,1,1,1,1,1))

```


## Explore genes of interest
```{r}
library(ggplot2)
library(wordcloud)
library(magrittr)

refins <- read.delim('workflow/results/10k/detettore/ALL_reference_insertions.extended.tsv', sep='\t', header=TRUE)

focal = 'phoP'
refins_gene = refins[grepl(focal, refins$gene),]

refins_gene_cnt = table(refins_gene$sublineage) %>% as.data.frame()

wordcloud(refins_gene_cnt$Var1, refins_gene_cnt$Freq)


ggplot(refins_gene, aes(x=as.factor(position))) + geom_histogram(stat = 'count',position = 'dodge') + facet_wrap(~lineage)



```



# Anchors: find filter thresholds
```{r}
library(ggplot2)

anchors <- read.csv('local/ALL_anchors.tsv', sep=',', header=T)
anchors$num_reads <- as.numeric(anchors$num_reads)
anchors$z_score <- as.numeric(anchors$z_score)
anchors$mod_z_score <- as.numeric(anchors$mod_z_score)
anchors$side <- as.factor(anchors$side)
summary(anchors)

# Add coverage
md <- read.csv('metadata/10k/metadata_reduced.tsv', sep='\t')
head(md)
anchors$coverage <- md$COVERAGE.MEAN[match(anchors$strain ==  md$GNUMBER)]




# z-score vs. modified z-score
ggplot(anchors, aes(z_score, mod_z_score)) + geom_point() +
  xlim(-6,6) + ylim(-6,6) +
  geom_abline(slope=1, intercept=0)

# 
ggplot(anchors, aes(num_reads, mod_z_score)) + 
  geom_point(alpha=0.5) +
  geom_point(data=subset(anchors, (num_reads < 10) | mod_z_score < -2), col='red')




```




# Trees with presence-absence matrices
Use small test tree/matrix to make sure things make sense!!


```{r}
library(ape)
library(ggtree)
library(plotly)
library(magrittr)
library(ggplot2)

tree <- read.tree('workflow/results/10k/phylogeny/snp_alignment.rooted.treefile') 

# Plot
p <- ggtree(tree)

# Subsample tree: 100 random per lineage

# Add metadata: keep only non-singleton sites, show them along the tree!
md <- read.delim('metadata/10k/metadata_reduced.tsv', sep='\t')
md$LINEAGE_x <- as.factor(md$LINEAGE_x)
md$SUBLINEAGE_x <- as.factor(md$SUBLINEAGE_x)


# Load presence-absence matrix
m <- read.csv('workflow/results/10k/detettore/ALL_presence-absence.5prime.tsv', sep='\t')
rownames(m) <- m[,1]
m <- m[,-1]
ncol(m)


# Sort insertions sites by frequency
m_counts <- colSums(m) %>% as.data.frame()
m_counts$ins <- rownames(m_counts)
names(m_counts)[1] <- 'count'
m_counts <- m_counts[order(m_counts$count, decreasing = T),]

ggplot(m_counts, aes(count)) + geom_histogram() + scale_x_log10() +
  theme_bw() + xlab("Number of strains with insertion site") + ylab("Count")


# Presence-absence metadata
m_meta <- read.csv('workflow/results/10k/detettore/ALL_presence-absence.metadata.tsv', sep='\t',
                   header=T)

m_meta <- subset(m_meta, side == '5')

mtb_col_scale <- scale_color_manual(values = c(
  "A1"="#fbf390", 
  "A2"="#c1f9dd", 
  "A3"="#dfe1f3", 
  "A4"="#f7cfe7", 
  "L1" = "#ff00ff", 
  "L2" = "#0000ff", 
  "L3" = "#a000cc", 
  "L4" = "#ff0000", 
  "L5" = "#663200", 
  "L6" = "#00cc33", 
  "L7" = "#ede72e", 
  "L8" ="#FF7F00" ,
  "L9" = "#006400",
  "La1"="black",
  "La2" ="black",
  "La3"="black",
  "La4"="black"
))

p <- ggtree(tree, layout="fan") %<+% md
p + geom_tippoint(aes(col=LINEAGE_x), size=0.0001, alpha=0.5) + mtb_col_scale

# Lineage or sublineage trees
l = 'La3'



# Use hclust to order columns
dist_cols <- dist(t(m_l), method = "binary") 
hc <- hclust(dist_cols, method = "complete")
ordered_cols <- hc$order
new_colnames <- colnames(m_l)[ordered_cols]
m_l_sorted <- m_l[, new_colnames]

# Order columns based on their colsums
m_l_sorted <- m_l[, order(colSums(m_l))]
m_l_sorted[] <- lapply(m_l_sorted, factor)

gheatmap(q, m_lf, colnames = FALSE) + scale_fill_manual(values=c("white", "black")) +
  ggtitle(l)


m_l_counts <- colSums(m_l) %>% as.data.frame()
m_l_counts$ins <- rownames(m_l_counts)
names(m_l_counts)[1] <- 'count'
m_l_counts <- m_l_counts[order(m_l_counts$count, decreasing = T),]
m_l_counts

for (i in 1:10){
  print(m_meta[m_meta$site_id == m_l_counts$ins[i],5])
}


```




# Reference vs. reference-free
```{r}
# Load presence-absence matrix
m <- read.delim('workflow/results/10k/detettore/ALL_presence-absence.tsv', sep='\t') %>% 
  as.matrix()

rown <- m[,1]
m <- m[,-1]
rownames(m) <- rown
m <- apply(m, 2, as.numeric)
dim(m)

# Load reference insertions
refins <- read.delim('/home/cristobal/TB/projects/MTBC_IS6110-poly/workflow/results/10k/detettore/ALL_reference_insertions.tsv', sep='\t', header=TRUE)
# Keep insertions where the reference support is at most 10% of the 5' support
dim(refins)
refins = subset(refins, support_ref < 0.1*support_5)


```

# Unused
```{r}
"
Create DataTrack for Gviz with the frequency of each unique insertion per 
lineage. 

Too small-scale to visualize...
"

# Unique positions found by detettore, to create GRanges object
positions = c()
strands = c()

for (ins in unique(refins$ID)){
  splt = strsplit(ins, '_')
  pos = as.numeric(splt[[1]][1])
  strand = splt[[1]][2]
  positions = c(positions, pos)
  strands = c(strands, strand)
}

gr <- GRanges(
  seqnames = "MTBC0", 
  ranges = IRanges(start = positions, end = positions), strand = '+')


# matrix with one column per range
matrix <- matrix(
  nrow = length(lineage_order), ncol = length(positions),
  dimnames = list(lineage_order, paste(positions, strands, sep='_')))

# Fill matrix with the frequency of each insertion per lineage
for (i in 1:length(unique(refins$ID))){
  ins = unique(refins$ID)[i]
  print(i)
  
  for (j in 1:length(lineage_order)){
   l = lineage_order[j]
   
   ins_count = nrow(subset(refins, ID == ins & lineage == l))
   lineage_strains = nrow(subset(md, LINEAGE_x == l))
   freq = ins_count/lineage_strains
   
   matrix[j,i] <- freq
 }
}

dTrack <- DataTrack(
  gr,
  data=matrix,
  type = c("heatmap"), 
  showSampleNames = TRUE, 
  cex.sampleNames = 0.6,
  gradient = c("white", "red"), # specify the color range
  )
```


# Check alignment: Identify samples with many missing alleles
```{r}
library(ggplot2)
source('workflow/notebooks/utils.R')

metadata <- read.csv('metadata/10k/metadata.tsv', sep='\t')
msamples <- read.csv('workflow/results/10k/alignment/snp_alignment.samples.tsv', sep='\t', header=FALSE)

msamples$zscore <- get_zscore(msamples$V3)


str(msamples)

# Are there some bad samples?
ggplot(msamples, aes(V3)) + geom_boxplot()
summary(msamples$V3)
samples_remove <- msamples$V1[which(msamples$zscore>2)]
length(samples_remove)
samples_remove_md <- metadata[which(metadata$GNUMBER %in% samples_remove),]

# Plot distributions of percentage < 5% and percentage not mapped
ggplot(metadata, aes(PERCENTAGE.LESS.5)) + geom_density()
ggplot(metadata, aes(PERCENTAGE.NOT.COVERED)) + geom_density()
summary(metadata$PERCENTAGE.LESS.5)
summary(metadata$PERCENTAGE.NOT.COVERED)



# Coverage drop??
library(readr)

depth <- read_csv("local/G734964.depth.gz", col_names = F)
head(depth)
depth <- as.data.frame(depth)
depth$pos <- c(1:nrow(depth))
tail(depth)

depth$bin <- cut(depth$pos, breaks = seq(0, nrow(depth), by = 5000), right = FALSE)

bin_data <- aggregate(X1 ~ bin, depth, mean)
head(bin_data)
tail(bin_data)



ggplot(bin_data, aes(x = bin, y = X1)) +
  geom_point() +
  labs(x = "Position (bins of 5000 base pairs)", y = "Average Coverage", title = "Sequencing Depth per Position in Bins of 5000 Base Pairs") +
  theme_minimal()


```

# Determine thresholds for sites

Less obvious, a site can still be informative for a small clade when it has been 
deleted in all other strains. 

```{r}
library(ggplot2)
source('workflow/notebooks/utils.R')

msites <- read.csv('workflow/results/10k/alignment/snp_alignment.sites.tsv', sep='\t', header=FALSE)
msites$zscore <- get_zscore(msites$V3)
summary(msites)
nrow(msites)


# Determine threshold for sites
ggplot(msites, aes(V3)) + geom_boxplot()
summary(msites$V3)


sites_remove <- subset(msites, V3>0.01)
nrow(sites_remove)
summary(sites_remove)

```



# Create presence-absence matrix for visulization in inkscape
```{r}
# Load necessary libraries
library(ggplot2)
library(tidyr)
library(dplyr)

# for visualization in R
table(metadata$lineage)



# Create a random 10x10 matrix
set.seed(123)  # For reproducibility
presence_absence_matrix <- matrix(sample(0:1, 100, replace = TRUE), nrow = 10, ncol = 10)

# Convert the matrix into a data frame for ggplot
matrix_df <- as.data.frame(presence_absence_matrix)
matrix_df$row <- rep(1:10, each = 1)
matrix_df$col <- rep(1:10, times = 1)

# Convert to long format for ggplot
matrix_long <- matrix_df %>%
  gather(key = "variable", value = "presence", -row, -col)

# Plot using ggplot2
ggplot(matrix_long, aes(x = col, y = row, fill = factor(presence))) +
  geom_tile(color = "white", size = 0.5) +  # Create the tiles
  scale_fill_manual(values = c("white", "blue")) +  # Color for presence (blue) and absence (white)
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1),  # Rotate x-axis labels
        axis.title = element_blank(),  # Remove axis titles
        axis.text = element_blank(),   # Remove axis text
        panel.grid = element_blank()) +  # Remove gridlines
  coord_fixed(ratio = 1)  # Maintain square shape of the matrix

```

